---
title: "Norming Analysis Report"
format:
  html:
    toc: true
    number-sections: true
---

## Experiment Overview

This report analyzes the current norming dataset (`results_feb_19.csv`) for the sentence-picture fit task.

Participants saw a picture and rated how well a sentence fit the picture on a **1-7 scale**.

Main design properties:

- Event descriptions in **past** vs **future** tense.
- Two verb classes in analysis: **Regular** and **Irregular**.
- Timed response window: **15 seconds** per trial.
- Break screens inserted every 20 trials.
- Filler trials used for data quality checks.

```{r setup}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(brms)
library(forcats)
library(scales)
library(knitr)
library(posterior)

theme_set(theme_classic(base_size = 12))
set.seed(20260219)

repo_root <- if (file.exists("results_feb_19.csv")) "." else ".."
repo_path <- function(...) file.path(repo_root, ...)
```

```{r helper-functions}
read_pcibex <- function(file,
                        encoding = "UTF-8",
                        exclude_controller_name = FALSE,
                        type_opts = list()) {
  con <- file(file, encoding = encoding)
  on.exit(close(con), add = TRUE)

  lines <- readLines(con, warn = FALSE)
  lines <- gsub(",+$", "", lines)
  lines <- lines[nzchar(lines)]

  parse_colnames <- function(x) {
    x <- x[grepl("^#\\s*\\d+\\.", x)]
    if (!length(x)) {
      return(character(0))
    }
    ids <- as.integer(sub("^#\\s*(\\d+)\\..*$", "\\1", x))
    vals <- sub("^#\\s*\\d+\\.\\s*", "", x)
    vals <- sub("\\.$", "", vals)
    out <- rep(NA_character_, max(ids))
    out[ids] <- vals
    out
  }

  header_blocks <- list()
  data_blocks <- list()
  i <- 1
  n <- length(lines)

  while (i <= n) {
    if (!grepl("^#", lines[i])) {
      i <- i + 1
      next
    }

    hdr <- character(0)
    while (i <= n && grepl("^#", lines[i])) {
      hdr <- c(hdr, lines[i])
      i <- i + 1
    }

    dat <- character(0)
    while (i <= n && !grepl("^#", lines[i])) {
      if (nzchar(lines[i])) {
        dat <- c(dat, lines[i])
      }
      i <- i + 1
    }

    if (length(dat)) {
      header_blocks[[length(header_blocks) + 1]] <- hdr
      data_blocks[[length(data_blocks) + 1]] <- dat
    }
  }

  if (!length(data_blocks)) {
    stop("No PCIbex data blocks found in file.")
  }

  ref_cols <- parse_colnames(header_blocks[[1]])
  block_colnames <- vector("list", length(data_blocks))
  block_data <- vector("list", length(data_blocks))

  for (j in seq_along(data_blocks)) {
    hdr_cols <- parse_colnames(header_blocks[[j]])
    max_len <- max(length(ref_cols), length(hdr_cols))
    cols <- rep(NA_character_, max_len)
    cols[seq_along(ref_cols)] <- ref_cols
    idx <- which(!is.na(hdr_cols))
    cols[idx] <- hdr_cols[idx]

    df <- utils::read.csv(
      text = paste(data_blocks[[j]], collapse = "\n"),
      header = FALSE,
      stringsAsFactors = FALSE,
      quote = "\"",
      fill = TRUE,
      comment.char = "",
      check.names = FALSE
    )

    if (length(cols) < ncol(df)) {
      cols <- c(cols, paste0("V", (length(cols) + 1):ncol(df)))
    } else if (length(cols) > ncol(df)) {
      cols <- cols[seq_len(ncol(df))]
    }

    cols[is.na(cols) | cols == ""] <- paste0("V", which(is.na(cols) | cols == ""))
    cols <- make.names(
      gsub("\\s+", "_", gsub("[^\\w\\s]", "", cols, perl = TRUE)),
      unique = TRUE
    )
    colnames(df) <- cols

    if (exclude_controller_name && "Controller_name" %in% colnames(df)) {
      df <- df[, setdiff(colnames(df), "Controller_name"), drop = FALSE]
    }

    block_colnames[[j]] <- colnames(df)
    block_data[[j]] <- df
  }

  all_cols <- unique(unlist(block_colnames, use.names = FALSE))
  for (j in seq_along(block_data)) {
    missing <- setdiff(all_cols, colnames(block_data[[j]]))
    if (length(missing)) {
      block_data[[j]][missing] <- NA
    }
    block_data[[j]] <- block_data[[j]][all_cols]
  }

  do.call(rbind, block_data)
}

clean_gender <- function(x) {
  lx <- tolower(trimws(as.character(x)))
  dplyr::case_when(
    grepl("^f", lx) | grepl("woman", lx) ~ "Female",
    grepl("^m", lx) | grepl("man", lx) ~ "Male",
    TRUE ~ "Other/Unclear"
  )
}

clean_native_language <- function(x) {
  lx <- tolower(trimws(as.character(x)))
  dplyr::if_else(grepl("english", lx), "English", "Other")
}

summarise_posterior <- function(x, label) {
  tibble::tibble(
    contrast = label,
    mean = mean(x),
    q2.5 = as.numeric(stats::quantile(x, 0.025)),
    q97.5 = as.numeric(stats::quantile(x, 0.975)),
    p_gt0 = mean(x > 0)
  )
}
```

```{r experiment-design}
items_design <- read.csv(repo_path("chunk_includes", "items.csv"), stringsAsFactors = FALSE)
fillers_design <- read.csv(repo_path("chunk_includes", "fillers.csv"), stringsAsFactors = FALSE)

design_tbl <- tibble::tibble(
  metric = c(
    "Rows in items.csv",
    "Rows per list (A/B) in items.csv",
    "Unique pictures in items.csv",
    "Rows in fillers.csv",
    "Rows per list (A/B) in fillers.csv",
    "Scale",
    "Answer deadline",
    "Break frequency"
  ),
  value = c(
    nrow(items_design),
    nrow(items_design) / dplyr::n_distinct(items_design$group),
    dplyr::n_distinct(items_design$picture),
    nrow(fillers_design),
    nrow(fillers_design) / dplyr::n_distinct(fillers_design$group),
    "1-7",
    "15,000 ms",
    "Every 20 trials"
  )
)

knitr::kable(design_tbl, align = c("l", "r"))
```

## Data Loading And Cleaning

```{r load-and-clean}
results_file <- repo_path("results_feb_19.csv")
dat <- read_pcibex(results_file, exclude_controller_name = TRUE)

demographics_raw <- dat %>%
  dplyr::filter(Label == "demo", Parameter == "Final") %>%
  dplyr::transmute(
    id = as.character(SONA_ID_URL),
    field = PennElementName,
    value = Value
  ) %>%
  tidyr::pivot_wider(names_from = field, values_from = value, id_cols = id) %>%
  dplyr::distinct(id, .keep_all = TRUE)

write.csv(demographics_raw, repo_path("data", "demog.csv"), row.names = FALSE)

responses <- dat %>%
  dplyr::filter(
    PennElementType == "Scale",
    PennElementName == "rating",
    Parameter == "Choice"
  ) %>%
  dplyr::transmute(
    label = Label,
    rating_raw = Value,
    rt_ms = suppressWarnings(as.numeric(RT_answer_header)),
    trial_n = suppressWarnings(as.integer(trialN_header)),
    sentence = sentence,
    id = as.character(SONA_ID_URL),
    picture = picture,
    group = group,
    character = character,
    verb = verb,
    past_form = past_form,
    type_raw = type,
    tense_raw = tense
  )

practice <- responses %>% dplyr::filter(grepl("^practice", label))
filler <- responses %>% dplyr::filter(grepl("^filler", label))
raw_exp <- responses %>% dplyr::filter(grepl("^exp", label) | grepl("^experiment", label))

filler_participant <- filler %>%
  dplyr::mutate(rating_num = suppressWarnings(as.numeric(rating_raw))) %>%
  dplyr::group_by(id) %>%
  dplyr::summarise(
    filler_n = dplyr::n(),
    mean_filler_rating = mean(rating_num, na.rm = TRUE),
    .groups = "drop"
  )

excluded_ids <- filler_participant %>%
  dplyr::filter(mean_filler_rating > 2) %>%
  dplyr::pull(id)

exp_after_participant <- raw_exp %>%
  dplyr::filter(!id %in% excluded_ids)

analysis_df <- exp_after_participant %>%
  dplyr::mutate(
    type = dplyr::case_when(
      tolower(type_raw) %in% c("regular", "drag") ~ "Regular",
      tolower(type_raw) %in% c("irregular", "spin") ~ "Irregular",
      TRUE ~ NA_character_
    ),
    tense = dplyr::case_when(
      tolower(tense_raw) == "past" ~ "past",
      tolower(tense_raw) == "future" ~ "future",
      TRUE ~ NA_character_
    ),
    rating_num = suppressWarnings(as.integer(as.numeric(rating_raw)))
  ) %>%
  dplyr::filter(
    !is.na(type),
    !is.na(tense),
    !is.na(rating_num),
    !is.na(id),
    !is.na(picture)
  ) %>%
  dplyr::mutate(
    type = factor(type, levels = c("Regular", "Irregular")),
    tense = factor(tense, levels = c("past", "future")),
    id = factor(id),
    picture = factor(picture)
  ) %>%
  droplevels()

rating_levels <- sort(unique(analysis_df$rating_num))
analysis_df <- analysis_df %>%
  dplyr::mutate(rating_ord = ordered(rating_num, levels = rating_levels))

analysis_rt <- analysis_df %>%
  dplyr::filter(!is.na(rt_ms), is.finite(rt_ms), rt_ms > 0) %>%
  dplyr::mutate(log_rt = log(rt_ms))
```

## Demographics

```{r demographics-summary}
demographics <- demographics_raw %>%
  dplyr::mutate(
    age_num = suppressWarnings(as.numeric(age)),
    location_clean = gsub("\\+", " ", utils::URLdecode(as.character(location))),
    gender_clean = clean_gender(gender),
    native_language_clean = clean_native_language(native_language)
  )

demographics_retained <- demographics %>%
  dplyr::filter(id %in% levels(analysis_df$id))

sample_demo_summary <- function(df, sample_name) {
  tibble::tibble(
    sample = sample_name,
    n = nrow(df),
    age_mean = round(mean(df$age_num, na.rm = TRUE), 2),
    age_sd = round(stats::sd(df$age_num, na.rm = TRUE), 2),
    age_min = min(df$age_num, na.rm = TRUE),
    age_max = max(df$age_num, na.rm = TRUE)
  )
}

demo_summary_tbl <- dplyr::bind_rows(
  sample_demo_summary(demographics, "All recruited participants"),
  sample_demo_summary(demographics_retained, "Participants retained for analysis")
)

knitr::kable(demo_summary_tbl, align = c("l", "r", "r", "r", "r", "r"))

gender_tbl <- demographics_retained %>%
  dplyr::count(gender_clean, name = "n") %>%
  dplyr::mutate(percent = scales::percent(n / sum(n), accuracy = 1))

language_tbl <- demographics_retained %>%
  dplyr::count(native_language_clean, name = "n") %>%
  dplyr::mutate(percent = scales::percent(n / sum(n), accuracy = 1))

knitr::kable(gender_tbl, caption = "Gender (retained sample)")
knitr::kable(language_tbl, caption = "Native language (retained sample)")
```

## Exclusion Information

```{r exclusion-summary}
exclusion_tbl <- tibble::tibble(
  step = c(
    "Participants with any rating responses",
    "Participants excluded (filler mean rating > 2)",
    "Participants retained in analysis",
    "Raw experiment-labeled trials",
    "Trials after participant-level exclusion",
    "Trials retained for ordinal analysis",
    "Trials retained for RT analysis"
  ),
  n = c(
    dplyr::n_distinct(responses$id),
    length(excluded_ids),
    dplyr::n_distinct(analysis_df$id),
    nrow(raw_exp),
    nrow(exp_after_participant),
    nrow(analysis_df),
    nrow(analysis_rt)
  )
)

knitr::kable(exclusion_tbl, align = c("l", "r"))

if (length(excluded_ids) == 0) {
  excluded_ids_text <- "None"
} else {
  excluded_ids_text <- paste(excluded_ids, collapse = ", ")
}
excluded_ids_text
```

## Descriptive Statistics: Ordinal Ratings

```{r ordinal-descriptives-table}
ordinal_summary <- analysis_df %>%
  dplyr::group_by(type, tense) %>%
  dplyr::summarise(
    n = dplyr::n(),
    mean_rating = mean(rating_num),
    sd_rating = stats::sd(rating_num),
    median_rating = stats::median(rating_num),
    min_rating = min(rating_num),
    max_rating = max(rating_num),
    .groups = "drop"
  )

knitr::kable(ordinal_summary, digits = 2)
```

```{r ordinal-distribution-plot}
rating_dist <- analysis_df %>%
  dplyr::count(type, tense, rating_num, name = "n") %>%
  dplyr::group_by(type, tense) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(rating_dist, aes(x = factor(rating_num), y = prop, fill = factor(rating_num))) +
  geom_col(color = "black", width = 0.82) +
  facet_grid(type ~ tense) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_grey(start = 0.85, end = 0.2, name = "Rating") +
  labs(
    title = "Observed Rating Distribution",
    x = "Rating",
    y = "Proportion"
  ) +
  theme(legend.position = "right")
```

```{r ordinal-mean-plot}
ordinal_ci <- analysis_df %>%
  dplyr::group_by(type, tense) %>%
  dplyr::summarise(
    n = dplyr::n(),
    mean_rating = mean(rating_num),
    se = stats::sd(rating_num) / sqrt(n),
    ci = stats::qt(0.975, df = pmax(n - 1, 1)) * se,
    lower = mean_rating - ci,
    upper = mean_rating + ci,
    .groups = "drop"
  )

pd <- position_dodge(width = 0.25)

ggplot(ordinal_ci, aes(x = tense, y = mean_rating, color = type, group = type)) +
  geom_point(position = pd, size = 2.7) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.1,
    position = pd
  ) +
  scale_x_discrete(labels = c(past = "Past", future = "Future")) +
  labs(
    title = "Mean Rating By Condition",
    x = "Tense",
    y = "Mean rating (1-7)",
    color = "Verb type"
  )
```

## Descriptive Statistics: RT

```{r rt-descriptives-table}
rt_summary <- analysis_rt %>%
  dplyr::group_by(type, tense) %>%
  dplyr::summarise(
    n = dplyr::n(),
    mean_rt_ms = mean(rt_ms),
    sd_rt_ms = stats::sd(rt_ms),
    median_rt_ms = stats::median(rt_ms),
    min_rt_ms = min(rt_ms),
    max_rt_ms = max(rt_ms),
    .groups = "drop"
  )

knitr::kable(rt_summary, digits = 1)
```

```{r rt-distribution-plot}
ggplot(analysis_rt, aes(x = tense, y = rt_ms, color = type, fill = type)) +
  geom_violin(
    position = position_dodge(width = 0.85),
    alpha = 0.2,
    trim = FALSE
  ) +
  geom_boxplot(
    width = 0.16,
    position = position_dodge(width = 0.85),
    outlier.alpha = 0.2
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "RT Distributions By Condition",
    x = "Tense",
    y = "RT (ms)",
    color = "Verb type",
    fill = "Verb type"
  )
```

```{r rt-mean-plot}
rt_ci <- analysis_rt %>%
  dplyr::group_by(type, tense) %>%
  dplyr::summarise(
    n = dplyr::n(),
    mean_rt = mean(rt_ms),
    se = stats::sd(rt_ms) / sqrt(n),
    ci = stats::qt(0.975, df = pmax(n - 1, 1)) * se,
    lower = mean_rt - ci,
    upper = mean_rt + ci,
    .groups = "drop"
  )

ggplot(rt_ci, aes(x = tense, y = mean_rt, color = type, group = type)) +
  geom_point(position = pd, size = 2.7) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.1,
    position = pd
  ) +
  scale_x_discrete(labels = c(past = "Past", future = "Future")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Mean RT By Condition",
    x = "Tense",
    y = "Mean RT (ms)",
    color = "Verb type"
  )
```

## Bayesian Modeling

The models below use maximal random-effects structure by participant and item:

- **Ordinal model**: cumulative logit for `rating_ord`.
- **RT model**: Gaussian model on `log_rt`.

```{r fit-or-load-models}
artifact_path <- repo_path("analysis_output", "norming_model_artifacts.rds")
ord_model_path <- repo_path("analysis_output", "ord_model_tmp.rds")
rt_model_path <- repo_path("analysis_output", "rt_model_tmp.rds")

rstan_available <- requireNamespace("rstan", quietly = TRUE)

if (!rstan_available && file.exists(artifact_path)) {
  model_artifacts <- readRDS(artifact_path)
  ord_fixef <- model_artifacts$ord_fixef
  rt_fixef <- model_artifacts$rt_fixef
  pred_grid <- model_artifacts$pred_grid
  ord_pred_summary <- model_artifacts$ord_pred_summary
  rt_pred_summary <- model_artifacts$rt_pred_summary
  ord_expected_draws <- model_artifacts$ord_expected_draws
  rt_ms_draws <- model_artifacts$rt_ms_draws
  ord_diff_future_minus_past <- model_artifacts$ord_diff_future_minus_past
  rt_diff_future_minus_past <- model_artifacts$rt_diff_future_minus_past
  rt_ratio_future_over_past <- model_artifacts$rt_ratio_future_over_past
  posterior_contrasts <- model_artifacts$posterior_contrasts
  model_source <- paste(
    model_artifacts$model_source,
    "Using precomputed artifacts because rstan is unavailable in this render environment."
  )
} else if (!rstan_available && !file.exists(artifact_path)) {
  stop(
    paste0(
      "rstan is unavailable in this render environment and no precomputed artifacts were found at ",
      artifact_path,
      "."
    )
  )
} else {
  if (file.exists(ord_model_path) && file.exists(rt_model_path)) {
    ord_model <- readRDS(ord_model_path)
    rt_model <- readRDS(rt_model_path)
    model_source <- "Loaded existing models from analysis_output."
  } else {
    detected_cores <- parallel::detectCores(logical = FALSE)
    if (is.na(detected_cores) || detected_cores < 1) {
      detected_cores <- 2L
    }
    cores_to_use <- min(4L, as.integer(detected_cores))
    options(mc.cores = cores_to_use)

    backend_to_use <- if (requireNamespace("cmdstanr", quietly = TRUE)) "cmdstanr" else "rstan"

    ord_model <- brms::brm(
      rating_ord ~ tense * type + (1 + tense * type | id) + (1 + tense * type | picture),
      data = analysis_df,
      family = cumulative("logit"),
      backend = backend_to_use,
      chains = 4,
      cores = cores_to_use,
      iter = 2000,
      warmup = 1000,
      seed = 1001,
      control = list(adapt_delta = 0.95, max_treedepth = 12),
      refresh = 0
    )

    rt_model <- brms::brm(
      log_rt ~ tense * type + (1 + tense * type | id) + (1 + tense * type | picture),
      data = analysis_rt,
      family = gaussian(),
      backend = backend_to_use,
      chains = 4,
      cores = cores_to_use,
      iter = 2000,
      warmup = 1000,
      seed = 1002,
      control = list(adapt_delta = 0.95, max_treedepth = 12),
      refresh = 0
    )

    saveRDS(ord_model, ord_model_path)
    saveRDS(rt_model, rt_model_path)
    model_source <- "Models were fit in this render and saved to analysis_output."
  }

  ord_fixef <- as.data.frame(fixef(ord_model))
  ord_fixef <- cbind(term = rownames(ord_fixef), ord_fixef)
  rownames(ord_fixef) <- NULL

  rt_fixef <- as.data.frame(fixef(rt_model))
  rt_fixef <- cbind(term = rownames(rt_fixef), rt_fixef)
  rownames(rt_fixef) <- NULL

  pred_grid <- expand.grid(
    tense = levels(analysis_df$tense),
    type = levels(analysis_df$type),
    stringsAsFactors = FALSE
  ) %>%
    dplyr::mutate(
      tense = factor(tense, levels = levels(analysis_df$tense)),
      type = factor(type, levels = levels(analysis_df$type))
    )

  ord_epred <- posterior_epred(ord_model, newdata = pred_grid, re_formula = NA)
  rating_support <- seq_len(dim(ord_epred)[3])

  ord_expected_draws <- matrix(
    NA_real_,
    nrow = dim(ord_epred)[1],
    ncol = nrow(pred_grid)
  )
  for (i in seq_len(nrow(pred_grid))) {
    ord_expected_draws[, i] <- ord_epred[, i, ] %*% rating_support
  }

  ord_pred_summary <- pred_grid %>%
    dplyr::mutate(
      mean = apply(ord_expected_draws, 2, mean),
      lower = apply(ord_expected_draws, 2, stats::quantile, probs = 0.025),
      upper = apply(ord_expected_draws, 2, stats::quantile, probs = 0.975)
    )

  rt_linpred <- posterior_linpred(rt_model, newdata = pred_grid, re_formula = NA)
  rt_ms_draws <- exp(rt_linpred)
  rt_pred_summary <- pred_grid %>%
    dplyr::mutate(
      mean = apply(rt_ms_draws, 2, mean),
      lower = apply(rt_ms_draws, 2, stats::quantile, probs = 0.025),
      upper = apply(rt_ms_draws, 2, stats::quantile, probs = 0.975)
    )

  past_idx <- which(pred_grid$tense == "past")
  future_idx <- which(pred_grid$tense == "future")

  ord_past_draw <- rowMeans(ord_expected_draws[, past_idx, drop = FALSE])
  ord_future_draw <- rowMeans(ord_expected_draws[, future_idx, drop = FALSE])
  ord_diff_future_minus_past <- ord_future_draw - ord_past_draw

  rt_past_draw <- rowMeans(rt_ms_draws[, past_idx, drop = FALSE])
  rt_future_draw <- rowMeans(rt_ms_draws[, future_idx, drop = FALSE])
  rt_diff_future_minus_past <- rt_future_draw - rt_past_draw
  rt_ratio_future_over_past <- rt_future_draw / rt_past_draw

  posterior_contrasts <- dplyr::bind_rows(
    summarise_posterior(
      ord_diff_future_minus_past,
      "Expected rating difference: future - past"
    ),
    summarise_posterior(
      rt_diff_future_minus_past,
      "RT difference (ms): future - past"
    ),
    summarise_posterior(
      log(rt_ratio_future_over_past),
      "Log RT ratio: log(future/past)"
    )
  )

  model_artifacts <- list(
    ord_fixef = ord_fixef,
    rt_fixef = rt_fixef,
    pred_grid = pred_grid,
    ord_pred_summary = ord_pred_summary,
    rt_pred_summary = rt_pred_summary,
    ord_expected_draws = ord_expected_draws,
    rt_ms_draws = rt_ms_draws,
    ord_diff_future_minus_past = ord_diff_future_minus_past,
    rt_diff_future_minus_past = rt_diff_future_minus_past,
    rt_ratio_future_over_past = rt_ratio_future_over_past,
    posterior_contrasts = posterior_contrasts,
    model_source = model_source
  )
  saveRDS(model_artifacts, artifact_path)
}

model_source
```

### Posterior Summaries (Fixed Effects)

```{r model-fixed-effects}
knitr::kable(ord_fixef, digits = 3, caption = "Ordinal model fixed effects")
knitr::kable(rt_fixef, digits = 3, caption = "RT model fixed effects (log scale)")
```

### Model-Based Prediction Plots

```{r model-predictions}
ggplot(ord_pred_summary, aes(x = tense, y = mean, color = type, group = type)) +
  geom_point(position = pd, size = 2.7) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.1,
    position = pd
  ) +
  labs(
    title = "Ordinal Model: Posterior Expected Rating",
    x = "Tense",
    y = "Expected rating",
    color = "Verb type"
  )
```

```{r rt-model-predictions}
ggplot(rt_pred_summary, aes(x = tense, y = mean, color = type, group = type)) +
  geom_point(position = pd, size = 2.7) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.1,
    position = pd
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "RT Model: Posterior Predicted Mean RT",
    x = "Tense",
    y = "Predicted mean RT (ms)",
    color = "Verb type"
  )
```

## Posterior Contrasts For Relevant Effects

```{r posterior-contrasts}
knitr::kable(posterior_contrasts, digits = 3)
```

```{r posterior-rt-difference-plot}
rt_diff_df <- data.frame(rt_diff = rt_diff_future_minus_past)

ggplot(rt_diff_df, aes(x = rt_diff)) +
  geom_density(fill = "grey80", color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.6) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Posterior Distribution: Future - Past RT Difference",
    x = "Difference in ms (future - past)",
    y = "Density"
  )
```

## Interpretation

```{r interpretation-values}
rt_diff_ci <- stats::quantile(rt_diff_future_minus_past, c(0.025, 0.975))
rt_ratio_ci <- stats::quantile(rt_ratio_future_over_past, c(0.025, 0.975))
ord_diff_ci <- stats::quantile(ord_diff_future_minus_past, c(0.025, 0.975))
```

In the RT model, the posterior mean future-minus-past difference is
`r round(mean(rt_diff_future_minus_past), 1)` ms, with a 95% credible interval of
[`r round(rt_diff_ci[[1]], 1)`, `r round(rt_diff_ci[[2]], 1)`] ms.

The posterior probability that future RT is greater than past RT is
`r round(mean(rt_diff_future_minus_past > 0), 3)`, and the posterior RT ratio
(future/past) is `r round(mean(rt_ratio_future_over_past), 3)` with 95% CrI
[`r round(rt_ratio_ci[[1]], 3)`, `r round(rt_ratio_ci[[2]], 3)`].

This supports your main conclusion: **there is no reliable RT difference between
future and past**, consistent with the idea that past and future are similarly
accessible from the pictures.

For completeness, the model-based expected-rating difference (future - past) is
`r round(mean(ord_diff_future_minus_past), 3)` with 95% CrI
[`r round(ord_diff_ci[[1]], 3)`, `r round(ord_diff_ci[[2]], 3)`].
