---
title: "Tense Timing Experiments: Analysis & Report"
format:
  typst:
    toc: true
    toc-depth: 2
    papersize: us-letter
    margin:
      x: 0.5in
      y: 0.5in
    fontsize: 10pt
    number-sections: true
fig-cap-location: bottom
tbl-cap-location: bottom
execute:
  echo: false
  warning: false
  message: false
  fig-align: center
  fig-width: 7
  fig-height: 2.5
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(include = FALSE)
# Typst embeds raster and SVG images but not PDF, so figures are rendered as
# PNG at print resolution rather than as vector PDF.
knitr::opts_chunk$set(dev = "png", dpi = 300)
library(ggplot2)
theme_set(theme_classic(base_size = 13, base_family = "Helvetica Neue"))
```

```{r}
# ─── FILTERING SWITCHES ──────────────────────────────────────────────────────
# Every filter below is optional. Set a switch to FALSE to skip that step; the
# reported participant counts and the write-up in "Participant filtering"
# follow whatever is set here.

filter_participants <- TRUE # master switch for all participant-level exclusion
filter_low_accuracy <- TRUE # drop participants below min_subject_accuracy
filter_adverbial <- FALSE # drop participants over max_adverbial_rate
filter_incomplete <- FALSE # drop participants with a failed upload chunk

filter_trials <- TRUE # drop individual trials failing the utterance screen

fit_models <- TRUE # TRUE fits the brms models (slow; cached to
# analysis/fits/*.rds, so it is a one-off cost)

min_subject_accuracy <- 0.50 # min proportion of trials passing all trial-level flags
max_adverbial_rate <- 0.20 # max proportion of "... in the past/future" responses
n_trials_full <- 108 # trials per participant in the CSV (all participants)

# Sessions whose recordings contain no speech. The microphone never delivered
# audio, so these are a recording failure rather than a performance one and are
# dropped before anything is counted, including the participant totals. The
# proportion of their recordings that transcribed as silence:
#   52505  48/48    52336  96/108    52385  84/108    52399  36/84
# Every other participant in the September run is at or below 2/108.
exclude_speakers <- c("52505", "52336", "52385", "52399")

# Experiment 3 is a speeded keypress task; RTs outside this window are either
# anticipations or disengagement. The raw range runs from 0 ms to 173 s.
rt_min <- 200 # ms
rt_max <- 5000 # ms

# Which priming interactions to estimate. The two priming variables are
# correlated, so their interaction asks whether tense priming depends on the
# morphophonological form also repeating. The three-way term additionally asks
# whether that dependency differs by tense. This was out of reach when Exp 3 had
# 771 trials from 15 participants; after the September collection it has roughly
# 2660 trials from 46, so an eight-cell split now leaves a few hundred trials per
# cell and the three-way term is worth estimating. Both stay off until that is a
# deliberate choice rather than a side effect of more data arriving.
fit_prime_interaction <- FALSE # tense prime x regularity prime
fit_three_way <- FALSE # ... additionally crossed with tense

# An effect is highlighted when this much of its posterior falls on one side of
# zero. It governs the bold rows in the coefficient tables and the "leaning"
# verdict in the prose, so both report the same standard.
pd_close <- 0.85

# Leave-one-out model comparison. It is by far the slowest part of a render and
# says nothing about the effects themselves, only about which spec in the ladder
# predicts best, so it is off while the report is being written. Set it to TRUE
# to restore the model-comparison tables in the appendix.
run_loo <- FALSE
# ─────────────────────────────────────────────────────────────────────────────

library(tidyverse)
library(janitor)
library(scales)
library(patchwork)

# Todo: make sure that it saves the subject id in the recording name.
# Todo: check if you can burn just the first 1 and 12.
#  in addition to primed and unprimed, can you make primed to be F->F Prime/ T->T Prime, and also T->F Switch, F->T Switch. And also same thing with the irregulars. But do the prime x tense first, and then do prime type x tense later in the plots.
```


```{=typst}
#import "@preview/eggs:0.6.0": *
#show: eggs

// The contents list stands on its own page.
#pagebreak()

// Captions a step below the body text, so a long one does not compete with it.
#show figure.caption: set text(size: 9pt)
// Tables may break across pages; a half figure is not worth showing.
#show figure.where(kind: "quarto-float-tbl"): set block(breakable: true)
#show table: set text(size: 8pt)

// Shared pieces for the procedure diagrams of all three experiments.
#let stage(body, fill: white) = box(
  inset: (x: 6pt, y: 5pt), radius: 2pt,
  stroke: 0.6pt + luma(40%), fill: fill,
  align(center + horizon, body)
)
#let arw = align(center + horizon, text(fill: luma(45%))[#sym.arrow.r])
#let note(body) = align(center, text(size: 7.5pt, fill: luma(35%), body))
#let sub(body) = text(size: 7.5pt, body)
#let cell(body, fill: white) = box(
  width: 100%, inset: (x: 2pt, y: 4pt),
  fill: fill, stroke: 0.5pt + luma(60%),
  align(center, text(size: 7.5pt, body))
)

#let unit(body, tone: 100%) = box(
  inset: (x: 5pt, y: 3pt), radius: 2pt, fill: luma(tone),
  stroke: 0.5pt + luma(55%), text(size: 7.5pt, body)
)
#let nests = block(width: 100%)[
  #grid(
    columns: (auto, auto), column-gutter: 6pt, row-gutter: 3pt, align: horizon,
    text(size: 7.5pt)[Session =],
    grid(columns: 3, column-gutter: 3pt,
      unit(tone: 90%)[meta-block], unit(tone: 90%)[meta-block], unit(tone: 90%)[meta-block]),
    text(size: 7.5pt)[Meta-block =],
    grid(columns: 3, column-gutter: 3pt,
      unit[block], unit[block], unit[block]),
    text(size: 7.5pt)[Block =],
    unit(tone: 78%)[*shown below*],
  )
]

// Columns of the timeline. "The" and the verb are each split so a bracket can
// start or stop inside a word rather than only at its edges.
#let tl = (1.0fr, 0.30fr, 0.30fr, 0.95fr, 0.55fr, 0.38fr, 0.90fr, 0.70fr)

#let region(body, tone: 96%) = box(
  width: 100%, fill: luma(tone), radius: 2pt,
  stroke: 0.5pt + luma(62%), inset: (x: 3pt, y: 7pt),
  align(center, text(size: 9pt, body))
)
#let span(body, tone: 84%) = box(
  width: 100%, fill: luma(tone), radius: 2pt,
  stroke: 0.8pt + luma(30%),
  inset: (x: 3pt, y: 7pt),
  align(center, text(size: 8pt, body))
)
#let blank = box(width: 100%, inset: (y: 7pt))[]
#let hyp(body) = align(right + horizon, text(size: 9pt, weight: "bold", body))

// The axis is one arrow, drawn as a rule with a head on its end.
#let timearrow = box(width: 100%, height: 9pt)[
  #place(left + horizon, line(length: 100% - 7pt, stroke: 1pt + luma(25%)))
  #place(right + horizon, polygon(
    fill: luma(25%), (0pt, 0pt), (7pt, 3.5pt), (0pt, 7pt),
  ))
]

// Linguistic examples come from the eggs package, which numbers them, turns a
// numbered list into lettered subexamples, and gives them referenceable labels.
```

```{r}
# User-defined function to read in PCIbex Farm results files
read.pcibex <- function(
  filepath,
  auto.colnames = TRUE,
  fun.col = function(col, cols) {
    cols[cols == col] <- paste(col, "Ibex", sep = ".")
    return(cols)
  }
) {
  n.cols <- max(count.fields(filepath, sep = ",", quote = NULL), na.rm = TRUE)
  if (auto.colnames) {
    cols <- c()
    con <- file(filepath, "r")
    while (TRUE) {
      line <- readLines(con, n = 1, warn = FALSE)
      if (length(line) == 0) {
        break
      }
      m <- regmatches(line, regexec("^# (\\d+)\\. (.+)\\.$", line))[[1]]
      if (length(m) == 3) {
        index <- as.numeric(m[2])
        value <- m[3]
        if (is.function(fun.col)) {
          cols <- fun.col(value, cols)
        }
        cols[index] <- value
        if (index == n.cols) {
          break
        }
      }
    }
    close(con)
    return(read.csv(
      filepath,
      comment.char = "#",
      header = FALSE,
      col.names = cols
    ))
  } else {
    return(read.csv(
      filepath,
      comment.char = "#",
      header = FALSE,
      col.names = seq(1:n.cols)
    ))
  }
}


subj_id <- function(
  df,
  cols = c("results_reception_time", "md5_hash_of_participant_s_ip_address"),
  label = "subj"
) {
  lookup <- df %>%
    distinct(across(all_of(cols))) %>%
    mutate(!!label := paste0("S[", row_number(), "]"))

  df %>% left_join(lookup, by = cols)
}

# ─── PLOTTING STYLE ──────────────────────────────────────────────────────────
# Grayscale print figures, matching the insensitivity/SPR paper: conditions are
# carried by shape + fill + linetype rather than colour, all ink is black, error
# bars are capless, and the minor grid is dropped.

theme_paper <- function(base_size = 13, legend_position = "right") {
  theme_classic(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.line = element_line(colour = "black", linewidth = 0.6),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", size = base_size - 2),
      legend.position = legend_position,
      legend.key.width = unit(1.2, "cm")
    )
}

pd <- position_dodge(width = 0.3)

# Filled marker = the marked/primed level, open marker = its baseline.
cond_shapes <- function(name, levels) {
  scale_shape_manual(name = name, values = setNames(c(16, 21), levels))
}
cond_fills <- function(name, levels) {
  scale_fill_manual(name = name, values = setNames(c("black", "white"), levels))
}
cond_linetypes <- function(name, levels) {
  scale_linetype_manual(
    name = name,
    values = setNames(c("solid", "22"), levels)
  )
}

# Morey (2008) within-subject SE: remove each subject's own mean, add the group
# mean back, then apply the sqrt(M/(M-1)) correction for M within-subject cells.
#
# `strata` names a variable whose levels are on different scales and must not be
# pooled (here `measure`: an onset in seconds and a duration in seconds are not
# comparable quantities). Normalisation, the cell count M, and the correction are
# all computed separately within each stratum. `...` gives the condition
# variables that define the cells.
within_se <- function(df, value, subject, ..., strata = NULL) {
  df %>%
    group_by({{ strata }}) %>%
    mutate(
      .grand = mean({{ value }}, na.rm = TRUE),
      .cells = n_distinct(paste(!!!enquos(...)))
    ) %>%
    group_by({{ strata }}, {{ subject }}) %>%
    mutate(.norm = {{ value }} - mean({{ value }}, na.rm = TRUE) + .grand) %>%
    group_by({{ strata }}, ...) %>%
    summarize(
      m = mean({{ value }}, na.rm = TRUE),
      n = sum(!is.na(.norm)),
      .cells = first(.cells),
      se = sd(.norm, na.rm = TRUE) /
        sqrt(n) *
        sqrt(.cells / max(.cells - 1, 1)),
      .groups = "drop"
    ) %>%
    mutate(lwr = m - 1.96 * se, upr = m + 1.96 * se) %>%
    select(-.cells)
}

save_figure <- function(
  plot,
  name,
  width = 6.2,
  height = 2.6,
  dir = "analysis/fig_output"
) {
  dir.create(dir, recursive = TRUE, showWarnings = FALSE)
  ggsave(
    file.path(dir, paste0(name, ".pdf")),
    plot,
    width = width,
    height = height,
    device = cairo_pdf
  )
  invisible(plot)
}
# ─────────────────────────────────────────────────────────────────────────────

add_trial_info <- function(
  df,
  subject_col = "subj",
  block_col = "block",
  tense_col = "tense",
  n_cycle = 12,
  burn_pos = c(1, 6, 7, 12)
) {
  df %>%
    group_by(.data[[subject_col]]) %>%
    mutate(trial_n = row_number()) %>%
    group_by(.data[[subject_col]], .data[[block_col]]) %>%
    mutate(
      block_trial_n = (row_number() - 1) %% n_cycle + 1,
      prime_status = case_when(
        block_trial_n %in% burn_pos ~ "burn",
        .data[[tense_col]] == lag(.data[[tense_col]]) ~ "primed",
        TRUE ~ "unprimed"
      )
    ) %>%
    ungroup()
}
```


```{r}
concept <- read.pcibex("data/concept_june6.csv") %>% clean_names() %>% subj_id()
exp <- concept %>%
  filter(penn_element_name == "tenseChoice") %>%
  select(
    subj,
    label,
    choice = value,
    assigned_list,
    block,
    verb,
    regularity,
    tense,
    entity,
    on_left = right_opt,
    on_right = practice_correct,
    decision_rt
  ) %>%
  # Tense repetition is the primary priming variable; regularity repetition is
  # tracked alongside it because the two are correlated in this sequence.
  add_trial_info() %>%
  rename(prime_status_main = prime_status) %>%
  add_trial_info(tense_col = "regularity") %>%
  rename(prime_reg = prime_status, prime_status = prime_status_main) %>%
  # Case differs across the three CSVs: concept logs "PAST"/"FUTURE" and
  # "Irregular"/"Regular", the alignments use lowercase tense from the file
  # stem, and phon logs "IRREGULAR"/"REGULAR". Normalise here so that scales,
  # contrast coding and tables match across experiments.
  mutate(
    tense = str_to_lower(tense),
    regularity = str_to_upper(regularity),
    decision_rt = as.double(decision_rt)
  ) %>%
  mutate(
    rt_ok = !is.na(decision_rt) & decision_rt >= rt_min & decision_rt <= rt_max
  ) %>%
  # Carry-over control, the Experiment 3 analogue of the previous utterance's
  # length: how long the preceding decision took.
  arrange(subj, trial_n) %>%
  group_by(subj) %>%
  mutate(prev_len = lag(if_else(rt_ok, decision_rt, NA_real_))) %>%
  ungroup()
```



```{r}
# 4. Read aligned TextGrids + join PCIbex metadata per experiment
library(readtextgrid)
library(fs)

read_tgs <- function(aligned_dir, exp_name) {
  tg_files <- dir_ls(aligned_dir, recurse = TRUE, glob = "*.TextGrid")
  if (length(tg_files) == 0) {
    return(tibble())
  }

  map_dfr(tg_files, \(fp) {
    tg <- read_textgrid(fp)
    stem <- path_file(fp) %>% str_remove("\\.TextGrid$")
    sona_id <- path_file(path_dir(fp))

    # Two naming generations. Recordings written before the naming change are
    # resp_m{mb}_block{bl}_p{pass}_{verb}_{tense} and carry no identifier, so the
    # participant is only known from the folder prepare_recordings.py chose.
    # Later ones are resp_{participant}_{session}_m{mb}_... and identify
    # themselves. The middle group is optional so both parse here.
    parts <- str_match(
      stem,
      paste0(
        "^resp_(?:([a-z0-9]+)_([a-z0-9]+)_)?",
        "m(\\d+)_block(\\d+)_p(\\d+)_([a-z]+)_(past|future|pres)$"
      )
    )
    id_in_name <- parts[, 2]
    speaker <- if (is.na(id_in_name)) sona_id else id_in_name

    tg %>%
      filter(text != "") %>%
      transmute(
        exp = exp_name,
        sona_id = speaker,
        recording = stem,
        metablock = parts[, 4],
        block = parts[, 5],
        pass = parts[, 6],
        verb = parts[, 7],
        tense = parts[, 8],
        tier = tier_name,
        word_phone = text,
        t_start = xmin,
        t_end = xmax,
        duration = xmax - xmin
      )
  })
}

# Takes one or more results files. Experiment 1 was collected in two runs that
# logged different URL parameters (the September run adds subject_id, a
# recruitment flag and the Prolific ids), so the shared columns sit at
# different positions and the files can only be combined once read.pcibex has
# named them. Participants are numbered after the bind so that subj is unique
# across runs, and each file's rows stay in presentation order, which is what
# add_trial_info counts positions from.
read_pcibex_meta <- function(
  filepaths,
  prime_col = "tense",
  also_prime = NULL
) {
  # Participants are numbered within each file, exactly as a single-file read
  # would number them, and later files continue the count so that subj stays
  # unique across runs. Numbering before the row filter keeps the labels a
  # one-file experiment produces identical to what it produced before.
  subj_offset <- 0L
  parts <- lapply(filepaths, function(filepath) {
    raw <- read.pcibex(filepath) %>%
      clean_names() %>%
      subj_id()
    n_here <- n_distinct(raw$subj)
    out <- raw %>%
      filter(parameter == "Filename", str_detect(value, "^resp")) %>%
      transmute(
        subj = sprintf(
          "S[%d]",
          as.integer(str_extract(subj, "\\d+")) + subj_offset
        ),
        sona_id = as.character(sona_id_url),
        recording = str_remove(value, "\\.webm$"),
        block = as.character(block),
        list = as.character(assigned_list),
        tense = as.character(tense),
        regularity = as.character(regularity),
        entity = as.character(entity),
        target_sentence = as.character(target_canonical_sentence)
      )
    subj_offset <<- subj_offset + n_here
    out
  })

  bind_rows(parts) %>%
    mutate(regularity = str_to_upper(regularity)) %>%
    add_trial_info(tense_col = prime_col) %>%
    # A second, non-primary priming variable: whether the OTHER factor repeated
    # from the preceding trial. Burn positions are defined on block position, so
    # both priming columns mark the same trials as "burn".
    {
      if (is.null(also_prime)) {
        .
      } else {
        rename(., prime_status_main = prime_status) %>%
          add_trial_info(tense_col = also_prime) %>%
          rename(prime_reg = prime_status, prime_status = prime_status_main)
      }
    } %>%
    select(-block, -tense) # drop: conflicts with stem-parsed columns in alignments
}

# phon: priming is by regularity (regular vs irregular morphophonology). Tense is
#       constant (all past), so tense priming is undefined here.
# syn:  priming is by tense (past vs future); regularity repetition is tracked
#       alongside it as prime_reg.
phon_meta <- read_pcibex_meta("data/phon_june6.csv", prime_col = "regularity")
syn_meta <- read_pcibex_meta(
  c("data/syn_june6.csv", "data/syn_sep6.csv"),
  prime_col = "tense",
  also_prime = "regularity"
)

alignments <- bind_rows(
  read_tgs("analysis/mfa/aligned/phon", "phon") %>%
    left_join(phon_meta, by = c("sona_id", "recording")),
  read_tgs("analysis/mfa/aligned/syntax", "syntax") %>%
    left_join(syn_meta, by = c("sona_id", "recording"))
) %>%
  filter(str_detect(recording, "^resp"), !sona_id %in% exclude_speakers)

glimpse(alignments)
```


```{r}
# 4b. Response accuracy screen
#
# Trial-level flags compare the spoken word sequence (MFA words tier, i.e. the
# ASR transcript) against the TargetCanonicalSentence logged by PCIbex for that
# trial. Subjects failing the screen are dropped from `alignments` here, so every
# downstream chunk works on retained subjects only.

past_forms <- c(
  blow = "blew",
  build = "built",
  carry = "carried",
  climb = "climbed",
  dig = "dug",
  drag = "dragged",
  drink = "drank",
  eat = "ate",
  paint = "painted",
  peel = "peeled",
  play = "played",
  push = "pushed",
  read = "read",
  ride = "rode",
  shake = "shook",
  smell = "smelled",
  spin = "spun",
  stir = "stirred",
  sweep = "swept",
  wash = "washed"
)

resp_spoken <- alignments %>%
  filter(tier == "words") %>%
  group_by(exp, sona_id, recording, verb, tense, entity, target_sentence) %>%
  arrange(t_start, .by_group = TRUE) %>%
  summarize(spoken_raw = paste(word_phone, collapse = " "), .groups = "drop")

response_flags <- resp_spoken %>%
  mutate(
    said = str_squish(str_to_lower(str_remove_all(spoken_raw, "[^a-zA-Z ]"))),
    target_words = str_squish(str_to_lower(str_remove_all(
      target_sentence,
      "[^a-zA-Z ]"
    ))),
    past_target = coalesce(unname(past_forms[verb]), paste0(verb, "ed")),

    # correct number of words
    n_words_ok = str_count(said, "\\S+") == str_count(target_words, "\\S+"),
    # correct head: the subject entity for this trial is named
    head_ok = !is.na(entity) &
      str_detect(said, paste0("\\b", str_to_lower(entity), "\\b")),
    # correct verb, in the inflection the trial called for
    verb_ok = case_when(
      # "blew" and "blue" are perfect homophones: the ASR cannot distinguish them
      # and writes the colour on most blow trials. Accepting it recovers correct
      # productions. "blowed" stays an error, being a real over-regularisation
      # rather than a transcription artefact.
      tense == "past" ~ str_detect(said, paste0("\\b", past_target, "\\b")) |
        (verb == "smell" & str_detect(said, "\\bsmelt\\b")) |
        (verb == "blow" & str_detect(said, "\\bblue\\b")),
      TRUE ~ str_detect(said, paste0("\\b", verb, "\\b"))
    ),
    # correct tense marking on the verb itself
    tense_ok = case_when(
      tense == "past" ~ verb_ok & !str_detect(said, "\\bwill\\b"),
      tense == "future" ~ verb_ok & str_detect(said, "\\bwill\\b"),
      TRUE ~ NA
    ),
    # adverbial register: tense carried by "in the past/future" instead of the
    # verb, e.g. "the wizard climbs a ladder in the future". No tense planning on
    # the verb, so these trials cannot speak to the question and are excluded.
    adverbial = str_detect(said, "\\bin the (past|future)\\b"),
    empty_resp = said %in% c("", "sil"),
    exact = said == target_words,

    # Scoring stops at the verb. Every measure (utterance onset, NP onset,
    # subject duration, gap, verb duration) lies at or before the verb, so what
    # follows it is irrelevant. Word count stays as a diagnostic but no longer
    # excludes trials: it rejected correct productions differing only in the
    # object's determiner, and the four items whose target object is not "a/an"
    # (blow/bubbles, drink/coffee, sweep/the floor, play/the guitar) absorbed
    # nearly all of that loss.
    utt_ok = head_ok &
      verb_ok &
      tense_ok &
      !adverbial &
      !empty_resp
  ) %>%
  select(
    exp,
    sona_id,
    recording,
    exact,
    n_words_ok,
    head_ok,
    verb_ok,
    tense_ok,
    adverbial,
    empty_resp,
    utt_ok
  )

subject_stats <- response_flags %>%
  group_by(exp, sona_id) %>%
  summarize(
    n_trials = n(),
    accuracy = mean(utt_ok, na.rm = TRUE),
    adverbial_rate = mean(adverbial, na.rm = TRUE),
    empty_rate = mean(empty_resp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Every participant has 108 trials in the CSV. Those below that lost one
  # upload chunk (one metablock-3 block, both passes = 12 trials); their data
  # are incomplete, not duplicated.
  mutate(complete_session = n_trials == n_trials_full)

# Sequential exclusion cascade. Each step sees only the participants that
# survived the earlier ones, so the counts add up to the final N.
filter_steps <- list(
  # Adverbial register first: it is a task-compliance criterion, and a response
  # that marks tense adverbially carries no tense planning on the verb whatever
  # its accuracy. Running it before the accuracy screen keeps the two counts
  # interpretable rather than letting accuracy absorb these participants.
  list(
    on = filter_adverbial,
    name = "Adverbial register",
    desc = sprintf(
      paste(
        "marked tense with \"in the past/future\" rather than on",
        "the verb on more than %.0f%% of trials"
      ),
      100 * max_adverbial_rate
    ),
    keep = function(d) d$adverbial_rate <= max_adverbial_rate
  ),
  list(
    on = filter_low_accuracy,
    name = "Low accuracy",
    desc = sprintf(
      "passed fewer than %.0f%% of trials on the utterance criteria",
      100 * min_subject_accuracy
    ),
    keep = function(d) d$accuracy >= min_subject_accuracy
  ),
  list(
    on = filter_incomplete,
    name = "Incomplete session",
    desc = sprintf(
      "had fewer than %d recordings, from a failed upload chunk",
      n_trials_full
    ),
    keep = function(d) d$complete_session
  )
)

alive <- subject_stats
filter_log <- tibble()
excluded <- tibble()

for (st in filter_steps) {
  if (!filter_participants || !st$on) {
    next
  }
  keep <- st$keep(alive)
  filter_log <- bind_rows(
    filter_log,
    alive %>%
      mutate(.keep_it = keep) %>%
      group_by(exp) %>%
      summarize(
        step = st$name,
        desc = st$desc,
        n_before = n(),
        n_removed = sum(!.keep_it),
        n_after = sum(.keep_it),
        .groups = "drop"
      )
  )
  excluded <- bind_rows(
    excluded,
    alive %>% filter(!keep) %>% mutate(step = st$name)
  )
  alive <- alive %>% filter(keep)
}

kept_subjects <- alive
subject_screen <- subject_stats %>%
  mutate(
    subject_ok = paste(exp, sona_id) %in%
      paste(kept_subjects$exp, kept_subjects$sona_id)
  )

if (filter_participants) {
  alignments <- alignments %>%
    semi_join(kept_subjects, by = c("exp", "sona_id"))
}

# Criterion-by-criterion pass rates, before any exclusion
response_flags %>%
  group_by(exp) %>%
  summarize(
    n = n(),
    across(
      c(
        exact,
        n_words_ok,
        head_ok,
        verb_ok,
        tense_ok,
        utt_ok,
        adverbial,
        empty_resp
      ),
      ~ mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) %>%
  print()

if (nrow(excluded)) {
  excluded %>% arrange(exp, accuracy) %>% print(n = Inf)
}
```




```{r}
# 5. Syntax + Phon: correct utterances + timing measures
# Toggle correctness criterion:
#   strict = TRUE  → spoken words must exactly match target sentence
#   strict = FALSE → utt_ok from the 4b screen (word count + head + verb + tense,
#                    excluding adverbial-register and empty responses)
strict_matching <- FALSE

syn_words <- alignments %>%
  filter(exp == "syntax", tier == "words") %>%
  group_by(sona_id, recording) %>%
  arrange(t_start, .by_group = TRUE) %>%
  mutate(word_n = row_number()) %>%
  ungroup()

# Reconstruct spoken form + tense-based correctness
# past   → word 3 is NOT "will" (speaker used a past form)
# future → word 3 IS "will"
syn_spoken <- syn_words %>%
  group_by(sona_id, recording) %>%
  summarize(spoken = paste(word_phone, collapse = " "), .groups = "drop")

syn_w3 <- syn_words %>%
  filter(word_n == 3) %>%
  select(sona_id, recording, w3 = word_phone)

syn_check <- syn_words %>%
  distinct(
    sona_id,
    recording,
    tense,
    verb,
    entity,
    regularity,
    subj,
    list,
    trial_n,
    block_trial_n,
    prime_status,
    prime_reg,
    target_sentence
  ) %>%
  left_join(syn_spoken, by = c("sona_id", "recording")) %>%
  left_join(syn_w3, by = c("sona_id", "recording")) %>%
  left_join(
    response_flags %>%
      select(
        sona_id,
        recording,
        utt_ok,
        adverbial,
        n_words_ok,
        head_ok,
        verb_ok,
        tense_ok
      ),
    by = c("sona_id", "recording")
  ) %>%
  mutate(
    target_words = str_to_lower(str_remove_all(
      target_sentence,
      "[^a-zA-Z ]"
    )) %>%
      str_squish(),
    correct = if (!filter_trials) {
      TRUE
    } else if (strict_matching) {
      spoken == target_words
    } else {
      coalesce(utt_ok, FALSE)
    }
  )

# Timing: pivot words 1–3 wide (the | entity | will/verb)
syn_timing <- syn_words %>%
  filter(word_n <= 3) %>%
  select(sona_id, recording, word_n, t_start, t_end) %>%
  pivot_wider(
    names_from = word_n,
    values_from = c(t_start, t_end),
    names_sep = "_w"
  )

# Join and compute measures
syn_analysis <- syn_check %>%
  left_join(syn_timing, by = c("sona_id", "recording")) %>%
  mutate(
    utt_onset = t_start_w1, # onset of "the"
    det_dur = t_end_w1 - t_start_w1, # duration of the determiner
    np_onset = t_start_w2,
    np_head_dur = t_end_w2 - t_start_w2, # duration of entity noun (NP head)
    np_offset = t_end_w2, # offset of full subject NP
    crit_onset = t_start_w3, # onset of "will" (future) or past verb
    gap = t_start_w3 - t_end_w2 # silent interval between NP and verb/will
  ) %>%
  select(
    sona_id,
    recording,
    subj,
    list,
    tense,
    verb,
    entity,
    regularity,
    trial_n,
    block_trial_n,
    prime_status,
    prime_reg,
    spoken,
    correct,
    adverbial,
    n_words_ok,
    head_ok,
    verb_ok,
    tense_ok,
    utt_onset,
    det_dur,
    np_onset,
    np_head_dur,
    np_offset,
    crit_onset,
    gap
  )

syn_analysis %>% filter(correct) %>% glimpse()
```


```{r}
# 8. Phon: same timing measures
phon_words <- alignments %>%
  filter(exp == "phon", tier == "words") %>%
  group_by(sona_id, recording) %>%
  arrange(t_start, .by_group = TRUE) %>%
  mutate(word_n = row_number()) %>%
  ungroup()

phon_spoken <- phon_words %>%
  group_by(sona_id, recording) %>%
  summarize(spoken = paste(word_phone, collapse = " "), .groups = "drop")

phon_w3 <- phon_words %>%
  filter(word_n == 3) %>%
  select(sona_id, recording, w3 = word_phone)

phon_timing <- phon_words %>%
  filter(word_n <= 3) %>%
  select(sona_id, recording, word_n, t_start, t_end) %>%
  pivot_wider(
    names_from = word_n,
    values_from = c(t_start, t_end),
    names_sep = "_w"
  )

phon_analysis <- phon_words %>%
  distinct(
    sona_id,
    recording,
    tense,
    verb,
    entity,
    regularity,
    subj,
    list,
    trial_n,
    block_trial_n,
    prime_status,
    target_sentence
  ) %>%
  left_join(phon_spoken, by = c("sona_id", "recording")) %>%
  left_join(phon_w3, by = c("sona_id", "recording")) %>%
  left_join(phon_timing, by = c("sona_id", "recording")) %>%
  left_join(
    response_flags %>%
      select(
        sona_id,
        recording,
        utt_ok,
        adverbial,
        n_words_ok,
        head_ok,
        verb_ok,
        tense_ok
      ),
    by = c("sona_id", "recording")
  ) %>%
  mutate(
    target_words = str_to_lower(str_remove_all(
      target_sentence,
      "[^a-zA-Z ]"
    )) %>%
      str_squish(),
    correct = if (!filter_trials) {
      TRUE
    } else if (strict_matching) {
      spoken == target_words
    } else {
      coalesce(utt_ok, FALSE)
    },
    utt_onset = t_start_w1,
    det_dur = t_end_w1 - t_start_w1,
    np_onset = t_start_w2,
    np_head_dur = t_end_w2 - t_start_w2,
    np_offset = t_end_w2,
    crit_onset = t_start_w3,
    gap = t_start_w3 - t_end_w2
  ) %>%
  select(
    sona_id,
    recording,
    subj,
    list,
    tense,
    verb,
    entity,
    regularity,
    trial_n,
    block_trial_n,
    prime_status,
    spoken,
    correct,
    adverbial,
    n_words_ok,
    head_ok,
    verb_ok,
    tense_ok,
    utt_onset,
    det_dur,
    np_onset,
    np_head_dur,
    np_offset,
    crit_onset,
    gap
  )

phon_analysis %>% filter(correct) %>% glimpse()
```

---


```{r, include=FALSE}
knitr::opts_chunk$set(include = TRUE, echo = FALSE)
library(knitr)

# ─── SHARED REPORT PREP ──────────────────────────────────────────────────────
# 1. All timing measures go to milliseconds here, once. Everything downstream
#    (tables, plots, models) is in ms.
# 2. Surface-form frequency: the Zipf frequency of the word form the trial
#    actually requires — the past form for past trials, the bare form for
#    future trials. Generated by analysis/scripts/make_verb_frequencies.py.
verb_freq <- read_csv("data/verb_frequencies.csv", show_col_types = FALSE)

# Word-level durations, with the phone count MFA aligned inside each word.
# Duration measures are then expressed per phoneme, so that a segmentally longer
# word is not counted as slower speech: "swept" and "carried" differ in length
# before any planning effect is considered.
phones_in_words <- {
  w <- alignments %>%
    filter(tier == "words") %>%
    select(
      exp,
      sona_id,
      recording,
      verb,
      tense,
      word = word_phone,
      w_start = t_start,
      w_end = t_end
    ) %>%
    group_by(exp, sona_id, recording) %>%
    arrange(w_start, .by_group = TRUE) %>%
    mutate(word_n = row_number()) %>%
    ungroup()
  ph <- alignments %>%
    filter(tier == "phones", !word_phone %in% c("sil", "sp", "spn", "")) %>%
    select(sona_id, recording, p_start = t_start, p_end = t_end)
  w %>%
    left_join(
      ph,
      by = c("sona_id", "recording"),
      relationship = "many-to-many"
    ) %>%
    filter(p_start >= w_start - 1e-6, p_end <= w_end + 1e-6) %>%
    count(
      exp,
      sona_id,
      recording,
      word_n,
      word,
      w_start,
      w_end,
      verb,
      tense,
      name = "n_phones"
    )
}

# The verb is found by matching the word against the form the trial required
# (past form for past trials, bare form for future), not by position: "will"
# takes slot 3 in the future condition and pushes the verb to slot 4.
word_metrics <- phones_in_words %>%
  left_join(verb_freq %>% select(verb, past_form), by = "verb") %>%
  mutate(target_form = if_else(tense == "past", past_form, verb)) %>%
  group_by(exp, sona_id, recording) %>%
  summarize(
    subj_nphones = n_phones[match(2, word_n)],
    verb_dur = {
      i <- match(TRUE, word == first(target_form))
      if (is.na(i)) NA_real_ else (w_end - w_start)[i]
    },
    verb_nphones = {
      i <- match(TRUE, word == first(target_form))
      if (is.na(i)) NA_integer_ else n_phones[i]
    },
    .groups = "drop"
  )

# Articulation rate for the trial, computed only from words that are NOT
# measured (i.e. excluding the subject noun and the verb). Using the whole
# utterance would be endogenous: a slowly produced subject would lower the very
# rate used to adjust it.
speech_rate <- phones_in_words %>%
  left_join(verb_freq %>% select(verb, past_form), by = "verb") %>%
  mutate(target_form = if_else(tense == "past", past_form, verb)) %>%
  group_by(exp, sona_id, recording) %>%
  filter(word_n != 2, word != first(target_form)) %>%
  summarize(
    rate = sum(n_phones) / sum(w_end - w_start),
    .groups = "drop"
  )

# Spoken duration of each utterance. AutoRecordMS is a constant 4500 ms for
# every trial, so the raw recording length carries no information; the length
# that varies is how long the participant actually spoke.
utt_duration <- alignments %>%
  filter(tier == "words") %>%
  group_by(exp, sona_id, recording) %>%
  summarize(utt_dur = max(t_end) - min(t_start), .groups = "drop")

attach_word_metrics <- function(df, exp_name) {
  df %>%
    left_join(
      word_metrics %>% filter(exp == exp_name) %>% select(-exp),
      by = c("sona_id", "recording")
    ) %>%
    left_join(
      speech_rate %>% filter(exp == exp_name) %>% select(-exp),
      by = c("sona_id", "recording")
    ) %>%
    left_join(
      utt_duration %>% filter(exp == exp_name) %>% select(-exp),
      by = c("sona_id", "recording")
    ) %>%
    # Carry-over control: how long the participant spoke on the PRECEDING trial.
    # trial_n is contiguous within participant, so lag() gives the true
    # predecessor; the first trial of each 12-trial cycle is a burn trial and is
    # excluded anyway, so no lag crosses a block boundary into the analysis set.
    arrange(sona_id, trial_n) %>%
    group_by(sona_id) %>%
    mutate(prev_len = lag(utt_dur)) %>%
    ungroup() %>%
    mutate(
      subj_dur_pp = np_head_dur / subj_nphones,
      verb_dur_pp = verb_dur / verb_nphones,
      z_rate = as.numeric(scale(rate)),
      z_prev = as.numeric(scale(prev_len)),
      entity = factor(entity)
    )
}
syn_analysis <- attach_word_metrics(syn_analysis, "syntax")
phon_analysis <- attach_word_metrics(phon_analysis, "phon")

time_cols <- c(
  "utt_onset",
  "det_dur",
  "np_onset",
  "np_head_dur",
  "np_offset",
  "crit_onset",
  "gap",
  "verb_dur",
  "subj_dur_pp",
  "verb_dur_pp"
)
syn_analysis <- syn_analysis %>% mutate(across(all_of(time_cols), ~ . * 1000))
phon_analysis <- phon_analysis %>% mutate(across(all_of(time_cols), ~ . * 1000))

# An onset of exactly 0 means the aligner placed speech at the very start of the
# recording: the participant was already talking, so the latency is not
# measurable rather than being zero.
drop_zero_onsets <- function(df) {
  df %>%
    mutate(across(
      any_of(c("utt_onset", "np_onset", "crit_onset")),
      ~ if_else(.x <= 0, NA_real_, .x)
    ))
}
syn_analysis <- drop_zero_onsets(syn_analysis)
phon_analysis <- drop_zero_onsets(phon_analysis)


attach_freq <- function(df) {
  df %>%
    left_join(
      verb_freq %>% select(verb, zipf_lemma, zipf_past),
      by = "verb"
    ) %>%
    mutate(
      zipf_form = if_else(tense == "past", zipf_past, zipf_lemma),
      # Frequency enters as a RANK over the distinct word forms: least frequent
      # = 1, most frequent = K. Ranking discards the spacing between Zipf
      # values, so a handful of very frequent verbs no longer dominates an
      # otherwise flat predictor.
      rank_freq = dense_rank(zipf_form),
      # Centred, then divided by its SD. Centring makes the intercept the
      # mid-frequency item; the SD division is a units choice that keeps the
      # existing normal(0, 0.3) prior on b meaningful. On raw centred ranks
      # (range +/- 9.5) that prior would imply implausible effects across the
      # range and would stop regularising.
      z_freq = as.numeric(scale(rank_freq))
    )
}
syn_analysis <- attach_freq(syn_analysis)
phon_analysis <- attach_freq(phon_analysis)

measure_levels <- c(
  "Utterance onset",
  "Determiner duration\n(ms)",
  "NP onset",
  "Subject duration\n(ms/phone)",
  "Verb duration\n(ms/phone)",
  "Gap\n(NP-verb/will)"
)

long_measures <- function(df) {
  df %>%
    filter(correct, prime_status != "burn") %>%
    select(
      sona_id,
      tense,
      regularity,
      prime_status,
      `Utterance onset` = utt_onset,
      `Determiner duration\n(ms)` = det_dur,
      `NP onset` = np_onset,
      `Subject duration\n(ms/phone)` = subj_dur_pp,
      `Verb duration\n(ms/phone)` = verb_dur_pp,
      `Gap\n(NP-verb/will)` = gap
    ) %>%
    pivot_longer(
      all_of(measure_levels),
      names_to = "measure",
      values_to = "value"
    ) %>%
    mutate(measure = factor(measure, levels = measure_levels))
}

syn_plot_data <- long_measures(syn_analysis)
phon_plot_data <- long_measures(phon_analysis)

cumulative_words <- function(exp_name, analysis_df) {
  alignments %>%
    filter(exp == exp_name, tier == "words") %>%
    group_by(sona_id, recording) %>%
    arrange(t_start, .by_group = TRUE) %>%
    mutate(word_n = row_number()) %>%
    ungroup() %>%
    semi_join(
      analysis_df %>% filter(correct),
      by = c("sona_id", "recording")
    ) %>%
    filter(prime_status != "burn", word_n <= 6) %>%
    mutate(across(c(t_start, t_end), ~ . * 1000))
}
syn_cumulative <- cumulative_words("syntax", syn_analysis)
phon_cumulative <- cumulative_words("phon", phon_analysis)

# 3. Counts for inline use.
concept_n_subj <- n_distinct(concept$sona_id_url)
concept_total <- nrow(exp)
concept_burn <- exp %>% filter(prime_status == "burn") %>% nrow()
concept_included <- exp %>% filter(prime_status != "burn", rt_ok) %>% nrow()
concept_rt_trimmed <- exp %>% filter(prime_status != "burn", !rt_ok) %>% nrow()

# How strongly tense repetition and regularity repetition co-vary in the
# conceptual sequence (they are anti-correlated by list construction).
concept_rep <- exp %>%
  filter(prime_status != "burn", prime_reg != "burn")
concept_rrep_given_tprime <- mean(
  concept_rep$prime_reg[concept_rep$prime_status == "primed"] == "primed"
)
concept_rrep_given_tswitch <- mean(
  concept_rep$prime_reg[concept_rep$prime_status == "unprimed"] == "primed"
)

exp_counts <- function(df) {
  list(
    n_subj = n_distinct(df$sona_id),
    total = nrow(df),
    wrong = sum(!df$correct, na.rm = TRUE),
    burn = sum(df$prime_status == "burn", na.rm = TRUE),
    included = sum(df$correct & df$prime_status != "burn", na.rm = TRUE)
  )
}
syn_c <- exp_counts(syn_analysis)
phon_c <- exp_counts(phon_analysis)
syn_c$pct_wrong <- round(100 * syn_c$wrong / syn_c$total)
phon_c$pct_wrong <- round(100 * phon_c$wrong / phon_c$total)

# ─── REPORT PLOT HELPERS ─────────────────────────────────────────────────────
# Facets always in a single row; y always milliseconds.
condition_means_plot <- function(plot_data, cond, cond_label, cond_levels) {
  plot_data %>%
    within_se(value, sona_id, {{ cond }}, prime_status, strata = measure) %>%
    ggplot(aes(
      x = prime_status,
      y = m,
      shape = {{ cond }},
      fill = {{ cond }},
      linetype = {{ cond }},
      group = {{ cond }}
    )) +
    facet_wrap(~measure, scales = "free_y", nrow = 1) +
    geom_line(position = pd, linewidth = 0.7, colour = "black") +
    geom_errorbar(
      aes(ymin = lwr, ymax = upr),
      position = pd,
      width = 0,
      linewidth = 0.7,
      colour = "black",
      linetype = "solid"
    ) +
    geom_point(position = pd, size = 2.5, colour = "black") +
    cond_shapes(cond_label, cond_levels) +
    cond_fills(cond_label, cond_levels) +
    cond_linetypes(cond_label, cond_levels) +
    labs(x = "Prime status", y = "Time (ms)") +
    theme_paper(base_size = 9, legend_position = "bottom")
}

timeline_plot <- function(cum_data, cond, cond_label) {
  cum_data %>%
    group_by({{ cond }}, prime_status, word_n) %>%
    summarize(
      m_onset = mean(t_start, na.rm = TRUE),
      m_offset = mean(t_end, na.rm = TRUE),
      se_onset = sd(t_start, na.rm = TRUE) / sqrt(n()),
      se_offset = sd(t_end, na.rm = TRUE) / sqrt(n()),
      .groups = "drop"
    ) %>%
    pivot_longer(
      c(m_onset, m_offset),
      names_to = "bound",
      values_to = "m",
      names_prefix = "m_"
    ) %>%
    mutate(
      se = if_else(bound == "onset", se_onset, se_offset),
      cond = interaction({{ cond }}, prime_status, sep = " / ")
    ) %>%
    ggplot(aes(
      x = word_n + if_else(bound == "offset", 0.3, 0),
      y = m,
      shape = cond,
      fill = cond,
      linetype = cond,
      group = interaction(cond, bound)
    )) +
    geom_line(linewidth = 0.8, alpha = 0.5, colour = "black") +
    geom_errorbar(
      aes(ymin = m - 1.96 * se, ymax = m + 1.96 * se),
      width = 0,
      linewidth = 0.6,
      colour = "black",
      linetype = "solid"
    ) +
    geom_point(size = 2.5, colour = "black") +
    scale_shape_manual(name = cond_label, values = c(16, 21, 17, 24)) +
    scale_fill_manual(
      name = cond_label,
      values = c("black", "white", "black", "white")
    ) +
    scale_linetype_manual(
      name = cond_label,
      values = c("solid", "22", "solid", "22")
    ) +
    scale_x_continuous(
      breaks = 1:4,
      labels = c("the", "NP", "will / Vpast", "V / obj")
    ) +
    labs(x = "Sentence region", y = "Time from trial onset (ms)") +
    theme_paper(base_size = 9) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

means_table <- function(plot_data, cond, caption) {
  # The plotting labels carry their units on a second line, which makes a very
  # wide first column and squeezes the rest. The table uses the short names and
  # states the units in its caption instead.
  short <- c(
    "Utterance onset" = "Utt. onset",
    "Determiner duration\n(ms)" = "Det. dur.",
    "NP onset" = "NP onset",
    "Subject duration\n(ms/phone)" = "Subj. dur.",
    "Verb duration\n(ms/phone)" = "Verb dur.",
    "Gap\n(NP-verb/will)" = "Gap"
  )
  d <- plot_data %>%
    within_se(value, sona_id, {{ cond }}, prime_status, strata = measure) %>%
    mutate(
      measure = coalesce(
        unname(short[as.character(measure)]),
        str_replace_all(as.character(measure), "\n", " ")
      ),
      Condition = as.character({{ cond }}),
      `Prime status` = prime_status,
      N = n,
      Mean = sprintf("%.0f", m),
      SE = sprintf("%.1f", se)
    )
  side_by_side(d, c("Condition", "Prime status", "N", "Mean", "SE"), caption)
}

# ─── MODEL WRAPPER ───────────────────────────────────────────────────────────
# One call per experiment fits every measure. `fixed` is the full population
# formula; `subj_slopes` / `item_slopes` name the terms that get random slopes
# (frequency is between-verb, so it never takes a by-verb slope).
fit_timing <- function(
  data,
  dv,
  fixed,
  subj_slopes,
  item_slopes,
  tag,
  family = shifted_lognormal(),
  priors = priors_lognormal,
  subject = "sona_id",
  extra_re = NULL,
  adapt_delta = 0.95
) {
  file <- file.path("analysis/fits", paste0(tag, "_", dv))
  if (!fit_models && !file.exists(paste0(file, ".rds"))) {
    return(NULL)
  }
  # Experiment 3 is keyed by `subj`, the production experiments by `sona_id`.
  stopifnot(all(c(subject, "verb", dv) %in% names(data)))
  main <- as.formula(sprintf(
    "%s ~ %s + (1 + %s | %s) + (1 + %s | verb)%s",
    dv,
    fixed,
    subj_slopes,
    subject,
    item_slopes,
    if (is.null(extra_re)) "" else paste(" +", extra_re)
  ))
  # A hurdle family splits the response into "did it happen at all" and "how
  # large was it". The gap is 67-79% exact zeros, so the same predictors are
  # allowed to act on both parts: priming may change how OFTEN speakers pause
  # rather than how long they pause for. In brms `hu` is the probability of a
  # ZERO, so a positive hu coefficient means MORE no-pause trials.
  is_hurdle <- grepl("hurdle", family$family %||% "", fixed = TRUE)
  form <- if (is_hurdle) {
    bf(main, as.formula(sprintf("hu ~ %s + (1 | %s)", fixed, subject)))
  } else {
    main
  }
  brm(
    form,
    data = data,
    family = family,
    prior = priors,
    chains = 4,
    cores = 4,
    iter = 4000,
    warmup = 1000,
    control = list(adapt_delta = adapt_delta, max_treedepth = 12),
    seed = 1234,
    # Stan's progress lines otherwise land in the results='asis' chunks that
    # emit the model tables, and pandoc then escapes the LaTeX around them.
    refresh = 0,
    silent = 2,
    file = file,
    file_refit = "on_change"
  )
}

# Facet titles for the coefficient plots, matching the condition-mean figures.
measure_pretty <- c(
  utt_onset = "Utt. onset",
  det_dur = "Det. dur.",
  np_onset = "NP onset",
  subj_dur_pp = "Subject dur.",
  verb_dur_pp = "Verb dur.",
  gap = "Gap length"
)

# ─── MODEL LADDER ────────────────────────────────────────────────────────────
# Each spec is fitted to every measure of the experiment. `slopes` is the
# by-participant random slope structure, always a subset of `fixed`: frequency,
# entity and rate are between-item or trial-level covariates and take no
# participant slope.
# Spec 6 (entity as a maximal random effect) is not fitted. With only three
# entity levels the funnel geometry produced 198-434 divergent transitions at
# adapt_delta = 0.95 and still 47-198 at 0.99, at roughly 8x the runtime of any
# other spec. Divergences mean the posterior was not properly explored, so those
# estimates would be biased rather than merely noisy. Spec 5 estimates the same
# entity contrasts as fixed effects and samples cleanly. The machinery in
# fit_timing() (extra_re, adapt_delta) is retained so the attempt is
# reproducible.
specs_tense <- list(
  list(
    id = "s1",
    label = "1: Tense x Prime",
    fixed = "c_tense * c_prime",
    slopes = "c_tense * c_prime"
  ),
  list(
    id = "s2",
    label = "2: + Frequency",
    fixed = "c_tense * c_prime + z_freq",
    slopes = "c_tense * c_prime"
  ),
  list(
    id = "s3",
    label = "3: + Reg. prime",
    fixed = "c_tense * c_prime + z_freq + c_rprime",
    slopes = "c_tense * c_prime + c_rprime"
  ),
  list(
    id = "s5",
    label = "5: + Entity, rate",
    fixed = "c_tense * c_prime + z_freq + c_rprime + entity + z_rate",
    slopes = "c_tense * c_prime + c_rprime"
  )
)

# Experiment 3 is a keypress task with no audio, so there is no articulation
# rate to control for; its spec 5 adds entity alone.
specs_tense_norate <- map(specs_tense, function(sp) {
  sp$fixed <- sub(" \\+ z_rate$", "", sp$fixed)
  sp
})

specs_reg <- list(
  list(
    id = "s1",
    label = "1: Regularity x Prime + Freq",
    fixed = "c_reg * c_prime + z_freq",
    slopes = "c_reg * c_prime"
  ),
  list(
    id = "s5",
    label = "2: + Entity, rate",
    fixed = "c_reg * c_prime + z_freq + entity + z_rate",
    slopes = "c_reg * c_prime"
  )
)

# Spec 5 controls for the subject noun and the trial's articulation rate. The
# three entity nouns differ sharply in per-phoneme duration (Chef 162, Pirate
# 118, Wizard 107 ms/phone), so leaving entity out loads that variance onto the
# residual for every duration measure.
# Every spec in the ladder is fitted and held in memory, and a brms fit is tens
# of megabytes. Only the headline spec is reported; the rest exist for the LOO
# comparison, so when that is switched off they are skipped rather than loaded
# and discarded. On a machine that has to swap, this is the difference between a
# render taking minutes and taking most of an hour.
fit_ladder <- function(data, measures, specs, tag, subject = "sona_id") {
  specs <- set_names(specs, map_chr(specs, "label"))
  if (!run_loo) {
    keep <- intersect(c(headline_spec, headline_spec_reg), names(specs))
    if (length(keep)) {
      specs <- specs[keep[1]]
    }
  }
  specs %>%
    map(function(sp) {
      set_names(measures, coalesce(measure_pretty[measures], measures)) %>%
        map(
          ~ fit_timing(
            data,
            .x,
            sp$fixed,
            sp$slopes,
            "c_prime",
            paste0(tag, "_", sp$id),
            extra_re = sp$extra_re,
            adapt_delta = sp$adapt_delta %||% 0.95,
            family = if (.x == "gap") {
              hurdle_lognormal()
            } else {
              shifted_lognormal()
            },
            priors = if (.x == "gap") priors_hurdle else priors_lognormal,
            subject = subject
          )
        ) %>%
        compact()
    })
}

# One table + one figure per spec, so every model in the ladder is inspectable.
# The reported model. The ladder is kept for the comparison table, but every
# figure and coefficient table in the body comes from this spec.
headline_spec <- "5: + Entity, rate"
headline_spec_reg <- "2: + Entity, rate"

pick_headline <- function(ladder) {
  nm <- intersect(c(headline_spec, headline_spec_reg), names(ladder))
  if (!length(nm)) {
    return(list())
  }
  ladder[[nm[1]]]
}

# ── Item-level forest plot ───────────────────────────────────────────────────
# By-verb deviations from the population intercept: how much each item pulls the
# measure up or down once the fixed effects are accounted for. Points are
# posterior medians, bars 95% CrIs; a verb whose interval excludes zero is
# behaving differently from the rest of the set.
item_forest <- function(models, exp_label) {
  if (!length(models)) {
    return(invisible(NULL))
  }
  d <- imap_dfr(models, function(fit, nm) {
    r <- ranef(fit, groups = "verb", probs = c(0.025, 0.975))$verb
    tibble(
      measure = nm,
      verb = dimnames(r)[[1]],
      est = r[, "Estimate", "Intercept"],
      lo = r[, "Q2.5", "Intercept"],
      hi = r[, "Q97.5", "Intercept"]
    )
  }) %>%
    mutate(measure = factor(measure, levels = names(models)))

  ord <- d %>%
    filter(measure == levels(d$measure)[1]) %>%
    arrange(est) %>%
    pull(verb)

  ggplot(d, aes(x = est, y = factor(verb, levels = ord))) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "gray50") +
    geom_errorbar(
      aes(xmin = lo, xmax = hi),
      orientation = "y",
      width = 0,
      linewidth = 0.5,
      colour = "black"
    ) +
    geom_point(size = 1.6, colour = "black") +
    facet_wrap(~measure, nrow = 1, scales = "free_x") +
    labs(
      x = "By-verb deviation from the intercept",
      y = NULL,
      title = exp_label
    ) +
    theme_paper(base_size = 9, legend_position = "none") +
    theme(axis.text.y = element_text(size = 6.5))
}

# ── Bayesian sanity checks ───────────────────────────────────────────────────
# Divergences mean the sampler could not explore part of the posterior, so the
# estimates are biased rather than merely noisy. Rhat > 1.01 or bulk ESS < 400
# means the chains have not mixed well enough to trust the summaries.
convergence_table <- function(models, caption) {
  if (!length(models)) {
    return(invisible(NULL))
  }
  tbl <- imap_dfr(models, function(fit, nm) {
    np <- nuts_params(fit)
    dr <- summarise_draws(as_draws_df(fit), "rhat", "ess_bulk", "ess_tail")
    dr <- dr[!is.na(dr$rhat), ]
    tibble(
      measure = nm,
      divergences = sum(subset(np, Parameter == "divergent__")$Value),
      max_rhat = round(max(dr$rhat), 3),
      min_ess_bulk = round(min(dr$ess_bulk)),
      min_ess_tail = round(min(dr$ess_tail))
    )
  }) %>%
    mutate(ok = divergences == 0 & max_rhat <= 1.01 & min_ess_bulk >= 400)
  print(kable(tbl, caption = caption))
  invisible(tbl)
}

# Posterior predictive check: simulated datasets from the fitted model against
# the observed distribution. Systematic mismatch means the likelihood is wrong,
# regardless of how well the chains converged.
pp_grid <- function(models, exp_label) {
  if (!length(models)) {
    return(invisible(NULL))
  }
  ps <- imap(models, function(fit, nm) {
    pp_check(fit, ndraws = 50) +
      labs(title = nm, x = NULL, y = NULL) +
      theme_paper(base_size = 7, legend_position = "none") +
      theme(
        plot.title = element_text(size = 7, face = "bold"),
        axis.text = element_text(size = 5.5)
      )
  })
  wrap_plots(ps, nrow = 1) +
    plot_annotation(title = paste(exp_label, "- posterior predictive checks"))
}

report_ladder <- function(ladder, term_labels, exp_label, fig_prefix) {
  for (nm in names(ladder)) {
    models <- ladder[[nm]]
    if (!length(models)) {
      next
    }
    cat(sprintf("\n\n**%s -- model %s**\n\n", exp_label, nm))
    report_models(
      models,
      term_labels,
      sprintf(
        "%s, model %s. Positive = slower/longer. Bold marks terms with P(dir) at or above %.2f.",
        exp_label,
        nm,
        pd_close
      ),
      figure_name = sprintf(
        "%s-%s",
        fig_prefix,
        tolower(gsub("[^a-z0-9]+", "-", tolower(nm)))
      )
    )
  }
}

# Predictive comparison across the ladder, per measure.
compare_ladder <- function(ladder, caption) {
  ms <- unique(unlist(lapply(ladder, names)))
  rows <- map_dfr(ms, function(m) {
    fits <- compact(lapply(ladder, function(l) l[[m]]))
    if (length(fits) < 2) {
      return(NULL)
    }
    # loo_compare() dispatches on its first argument: a bare list of brmsfit
    # objects lands in the default method and errors. Pass the loo objects.
    out <- try(
      {
        fits <- lapply(fits, add_criterion, "loo")
        loos <- lapply(fits, function(f) f$criteria$loo)
        cmp <- loo_compare(loos)
        tibble(
          measure = m,
          model = rownames(cmp),
          elpd_diff = round(cmp[, "elpd_diff"], 1),
          se_diff = round(cmp[, "se_diff"], 1)
        )
      },
      silent = TRUE
    )
    # A failed comparison must not discard the fits that took hours to produce.
    if (inherits(out, "try-error")) {
      warning(
        "LOO comparison failed for ",
        m,
        ": ",
        conditionMessage(attr(out, "condition"))
      )
      NULL
    } else {
      out
    }
  })
  if (nrow(rows)) print(kable(rows, caption = caption))
}

fit_all_measures <- function(
  data,
  measures,
  fixed,
  subj_slopes,
  item_slopes,
  tag
) {
  set_names(measures, coalesce(measure_pretty[measures], measures)) %>%
    map(
      ~ fit_timing(
        data,
        .x,
        fixed,
        subj_slopes,
        item_slopes,
        tag,
        family = if (.x == "gap") hurdle_lognormal() else shifted_lognormal(),
        priors = if (.x == "gap") priors_hurdle else priors_lognormal
      )
    ) %>%
    compact()
}


# Readable table: one block per measure, terms spelled out, and the rows with
# most of the posterior on one side of zero in bold.
# Probabilities are never printed as 0.00 or 1.00, which would claim more
# certainty than a finite posterior sample can support.
fmt_p <- function(p) {
  ifelse(p >= 0.995, "> 0.99", ifelse(p <= 0.005, "< 0.01", sprintf("%.2f", p)))
}

# Long tables are split into two side-by-side panels so a five-measure model
# fits on one page. The split falls on a measure boundary, never inside one.
# Long tables are split into two side-by-side panels so that a table with many
# measures fits on one page. The split falls on a measure boundary, never inside
# one, and a blank row separates the blocks. Emphasis is markdown, which Typst
# renders, so no engine-specific markup appears here.
side_by_side <- function(d, cols, caption, font_size = 7) {
  strong <- if (is.null(d[["strong"]])) rep(FALSE, nrow(d)) else d[["strong"]]
  blocks <- split(
    seq_len(nrow(d)),
    factor(d$measure, levels = unique(d$measure))
  )
  sizes <- lengths(blocks)
  cut <- which.min(abs(cumsum(sizes) - sum(sizes) / 2))
  left <- blocks[seq_len(cut)]
  right <- blocks[-seq_len(cut)]

  bold <- function(x, hit) {
    x <- as.character(x)
    ifelse(hit & nzchar(x), paste0("**", x, "**"), x)
  }

  # One block of rows per measure, with the measure named on its first row only,
  # and a blank spacer row after every block but the last.
  panel <- function(idx_list) {
    if (!length(idx_list)) {
      return(NULL)
    }
    parts <- imap(idx_list, function(idx, nm) {
      dd <- d[idx, , drop = FALSE]
      hit <- strong[idx]
      out <- data.frame(
        Measure = c(nm, rep("", nrow(dd) - 1)),
        lapply(dd[, cols, drop = FALSE], bold, hit = hit),
        check.names = FALSE,
        stringsAsFactors = FALSE
      )
      names(out) <- c("Measure", cols)
      out
    })
    blank <- parts[[1]][0, , drop = FALSE]
    blank[1, ] <- ""
    reduce(parts, function(a, b) rbind(a, blank, b))
  }

  L <- panel(left)
  R <- panel(right)
  if (is.null(R)) {
    return(kable(L, caption = caption, row.names = FALSE))
  }

  n <- max(nrow(L), nrow(R))
  pad <- function(x) {
    while (nrow(x) < n) {
      x[nrow(x) + 1, ] <- ""
    }
    x
  }
  both <- cbind(pad(L), pad(R))
  hdr <- c("Measure", "Term", "Est.", "95% CrI", "P(dir)")
  if (length(cols) == 5) {
    hdr <- c("Measure", "Condition", "Prime status", "N", "Mean", "SE")
  }
  kable(
    both,
    caption = caption,
    row.names = FALSE,
    col.names = rep(hdr, 2),
    align = rep(c("l", "l", rep("r", ncol(L) - 2)), 2)
  )
}

# `split` puts the measures in two panels side by side, which is what a figure
# caption's worth of space allows. The appendix has a whole page, so a single
# tall table reads better and no column has to be hyphenated.
model_table <- function(tbl, term_labels, caption, split = TRUE) {
  # Labels for the control terms, which are the same in every model.
  extra <- c(
    entityPirate = "Pirate vs Chef",
    entityWizard = "Wizard vs Chef",
    z_rate = "Articulation rate",
    z_freq = "Frequency",
    c_rprime = "Regularity priming"
  )
  d <- tbl %>%
    filter(!grepl("Intercept", .variable)) %>%
    mutate(measure = factor(measure, levels = unique(measure))) %>%
    arrange(measure) %>%
    mutate(
      Term = label_terms(.variable, term_labels),
      strong = pmax(p_greater_0, 1 - p_greater_0) >= pd_close,
      Estimate = sprintf("%.3f", estimate),
      `95% CrI` = sprintf("[%.3f, %.3f]", lower_95, upper_95),
      # P(direction): the posterior mass on the side of zero the estimate is on.
      `P(dir)` = fmt_p(pmax(p_greater_0, 1 - p_greater_0))
    ) %>%
    select(measure, Term, Estimate, `95% CrI`, `P(dir)`, strong)

  if (split && n_distinct(d$measure) > 2) {
    return(side_by_side(
      d,
      c("Term", "Estimate", "95% CrI", "P(dir)"),
      caption,
      font_size = 6
    ))
  }
  # A single block, with the measure named on its first row so the grouping
  # survives without engine-specific row packing, and a blank line between
  # measures so the blocks stay apart.
  out <- d %>%
    mutate(
      across(
        c(Term, Estimate, `95% CrI`, `P(dir)`),
        ~ ifelse(strong & nzchar(.x), paste0("**", .x, "**"), .x)
      ),
      Measure = ifelse(!duplicated(measure), as.character(measure), "")
    ) %>%
    select(measure, Measure, Term, Estimate, `95% CrI`, `P(dir)`) %>%
    # `measure` is a factor, and a blank spacer row cannot be written into one.
    mutate(across(everything(), as.character)) %>%
    as.data.frame()

  parts <- split(out, factor(out$measure, levels = unique(out$measure)))
  blank <- parts[[1]][0, , drop = FALSE]
  blank[1, ] <- ""
  out <- reduce(parts, function(a, b) rbind(a, blank, b))
  out <- out[, setdiff(names(out), "measure"), drop = FALSE]

  kable(out, caption = caption, row.names = FALSE, align = "llrrr")
}


report_ladder <- function(ladder, term_labels, exp_label, fig_prefix) {
  for (nm in names(ladder)) {
    models <- ladder[[nm]]
    if (!length(models)) {
      next
    }
    cat(sprintf("\n\n**%s -- model %s**\n\n", exp_label, nm))
    report_models(
      models,
      term_labels,
      sprintf(
        "%s, model %s. Positive = slower/longer. Bold marks terms with P(dir) at or above %.2f.",
        exp_label,
        nm,
        pd_close
      ),
      figure_name = sprintf(
        "%s-%s",
        fig_prefix,
        tolower(gsub("[^a-z0-9]+", "-", tolower(nm)))
      )
    )
  }
}

# Predictive comparison across the ladder, per measure.
compare_ladder <- function(ladder, caption) {
  ms <- unique(unlist(lapply(ladder, names)))
  rows <- map_dfr(ms, function(m) {
    fits <- compact(lapply(ladder, function(l) l[[m]]))
    if (length(fits) < 2) {
      return(NULL)
    }
    # loo_compare() dispatches on its first argument: a bare list of brmsfit
    # objects lands in the default method and errors. Pass the loo objects.
    out <- try(
      {
        fits <- lapply(fits, add_criterion, "loo")
        loos <- lapply(fits, function(f) f$criteria$loo)
        cmp <- loo_compare(loos)
        tibble(
          measure = m,
          model = rownames(cmp),
          elpd_diff = round(cmp[, "elpd_diff"], 1),
          se_diff = round(cmp[, "se_diff"], 1)
        )
      },
      silent = TRUE
    )
    # A failed comparison must not discard the fits that took hours to produce.
    if (inherits(out, "try-error")) {
      warning(
        "LOO comparison failed for ",
        m,
        ": ",
        conditionMessage(attr(out, "condition"))
      )
      NULL
    } else {
      out
    }
  })
  if (nrow(rows)) print(kable(rows, caption = caption))
}

fit_all_measures <- function(
  data,
  measures,
  fixed,
  subj_slopes,
  item_slopes,
  tag
) {
  set_names(measures, coalesce(measure_pretty[measures], measures)) %>%
    map(
      ~ fit_timing(
        data,
        .x,
        fixed,
        subj_slopes,
        item_slopes,
        tag,
        family = if (.x == "gap") hurdle_lognormal() else shifted_lognormal(),
        priors = if (.x == "gap") priors_hurdle else priors_lognormal
      )
    ) %>%
    compact()
}


# Readable table: one block per measure, terms spelled out, and the rows with
# most of the posterior on one side of zero in bold.

report_ladder <- function(ladder, term_labels, exp_label, fig_prefix) {
  for (nm in names(ladder)) {
    models <- ladder[[nm]]
    if (!length(models)) {
      next
    }
    cat(sprintf("\n\n**%s -- model %s**\n\n", exp_label, nm))
    report_models(
      models,
      term_labels,
      sprintf(
        "%s, model %s. Positive = slower/longer. Bold marks terms with P(dir) at or above %.2f.",
        exp_label,
        nm,
        pd_close
      ),
      figure_name = sprintf(
        "%s-%s",
        fig_prefix,
        tolower(gsub("[^a-z0-9]+", "-", tolower(nm)))
      )
    )
  }
}

# Predictive comparison across the ladder, per measure.
compare_ladder <- function(ladder, caption) {
  ms <- unique(unlist(lapply(ladder, names)))
  rows <- map_dfr(ms, function(m) {
    fits <- compact(lapply(ladder, function(l) l[[m]]))
    if (length(fits) < 2) {
      return(NULL)
    }
    # loo_compare() dispatches on its first argument: a bare list of brmsfit
    # objects lands in the default method and errors. Pass the loo objects.
    out <- try(
      {
        fits <- lapply(fits, add_criterion, "loo")
        loos <- lapply(fits, function(f) f$criteria$loo)
        cmp <- loo_compare(loos)
        tibble(
          measure = m,
          model = rownames(cmp),
          elpd_diff = round(cmp[, "elpd_diff"], 1),
          se_diff = round(cmp[, "se_diff"], 1)
        )
      },
      silent = TRUE
    )
    # A failed comparison must not discard the fits that took hours to produce.
    if (inherits(out, "try-error")) {
      warning(
        "LOO comparison failed for ",
        m,
        ": ",
        conditionMessage(attr(out, "condition"))
      )
      NULL
    } else {
      out
    }
  })
  if (nrow(rows)) print(kable(rows, caption = caption))
}

fit_all_measures <- function(
  data,
  measures,
  fixed,
  subj_slopes,
  item_slopes,
  tag
) {
  set_names(measures, coalesce(measure_pretty[measures], measures)) %>%
    map(
      ~ fit_timing(
        data,
        .x,
        fixed,
        subj_slopes,
        item_slopes,
        tag,
        family = if (.x == "gap") hurdle_lognormal() else shifted_lognormal(),
        priors = if (.x == "gap") priors_hurdle else priors_lognormal
      )
    ) %>%
    compact()
}

# One place where raw parameter names become readable labels.
label_terms <- function(v, term_labels) {
  extra <- c(
    entityPirate = "Pirate vs Chef",
    entityWizard = "Wizard vs Chef",
    z_rate = "Articulation rate",
    z_freq = "Frequency",
    c_rprime = "Regularity priming"
  )
  bare <- sub("^b_(hu_)?", "", v)
  out <- unname(term_labels[paste0("b_", bare)])
  out[is.na(out)] <- unname(extra[bare[is.na(out)]])
  out[is.na(out)] <- bare[is.na(out)]
  out
}

term_order <- function(lab, term_labels) {
  factor(
    lab,
    levels = rev(unique(c(
      unname(term_labels),
      "Regularity priming",
      "Frequency",
      "Articulation rate",
      "Pirate vs Chef",
      "Wizard vs Chef",
      sort(unique(lab))
    )))
  )
}

coef_table <- function(models) {
  if (!length(models)) {
    return(NULL)
  }
  imap_dfr(models, function(fit, nm) {
    fit %>%
      gather_draws(`b_.*`, regex = TRUE) %>%
      group_by(.variable) %>%
      summarize(
        estimate = median(.value),
        lower_95 = quantile(.value, 0.025),
        upper_95 = quantile(.value, 0.975),
        p_greater_0 = mean(.value > 0),
        .groups = "drop"
      ) %>%
      # The pause model has two parts: the duration of a pause when one occurs,
      # and the probability of producing none. They are separate panels.
      mutate(measure = if_else(grepl("^b_hu_", .variable), "No pause lik.", nm))
  })
}

coef_plot <- function(tbl, term_labels) {
  keep <- unname(term_labels)
  d <- tbl %>%
    filter(!grepl("Intercept", .variable)) %>%
    filter(label_terms(.variable, term_labels) %in% keep) %>%
    mutate(
      lab = label_terms(.variable, term_labels),
      lab = term_order(lab, term_labels),
      measure = factor(measure, levels = unique(measure)),
      pdir = pmax(p_greater_0, 1 - p_greater_0)
    ) %>%
    droplevels() %>%
    group_by(measure) %>%
    mutate(.text_x = max(upper_95) + 0.34 * (max(upper_95) - min(lower_95))) %>%
    ungroup()

  ggplot(d, aes(x = estimate, y = lab)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "gray50") +
    geom_errorbar(
      aes(xmin = lower_95, xmax = upper_95),
      orientation = "y",
      width = 0.15,
      linewidth = 0.7,
      colour = "black"
    ) +
    geom_point(size = 2.4, colour = "black") +
    geom_text(aes(x = .text_x, label = fmt_p(pdir)), hjust = 0.5, size = 2.6) +
    facet_wrap(~measure, nrow = 1, scales = "free_x") +
    # The right margin has to clear the P(dir) column, and four breaks in a
    # narrow facet collide, so the axis is given fewer.
    scale_x_continuous(expand = expansion(mult = c(0.12, 0.46)), n.breaks = 3) +
    coord_cartesian(clip = "off") +
    labs(x = "Estimate (log scale)", y = NULL) +
    theme_paper(legend_position = "none") +
    theme(
      axis.text.y = element_text(face = "bold", size = 8),
      # Six facets across one page leaves each axis about an inch wide, so the
      # tick labels have to be small enough not to run together.
      axis.text.x = element_text(size = 6.5),
      plot.margin = margin(t = 4, r = 10, b = 4, l = 4, unit = "pt")
    )
}


# `what` splits the two outputs: the body of the report carries the figures and
# the appendix carries the tables behind them, so a model is described once in
# each place rather than twice on the same page.
report_models <- function(
  models,
  term_labels,
  caption,
  figure_name = NULL,
  hide_terms = character(),
  what = c("both", "figure", "table"),
  split = TRUE
) {
  what <- match.arg(what)
  if (!length(models)) {
    cat("\n*Models not fitted. Set `fit_models <- TRUE` and re-render.*\n\n")
    return(invisible(NULL))
  }
  tbl <- coef_table(models)
  # Terms kept in the model as controls but left out of what is reported. The
  # coefficient plot drops them because they are absent from term_labels; the
  # table lists every fitted term, so it has to be told.
  if (length(hide_terms)) {
    tbl <- tbl %>% filter(!sub("^b_(hu_)?", "", .variable) %in% hide_terms)
  }
  # The table has to be the value the chunk ends on. Printing it from inside a
  # function makes knitr treat the markdown as console output, and it lands in
  # the document as literal pipes and asterisks.
  if (what == "table") {
    return(model_table(tbl, term_labels, caption, split = split))
  }
  if (what == "both") {
    print(model_table(tbl, term_labels, caption))
  }

  # The pause model is on a different scale from the timing measures, so it gets
  # its own panel row rather than being squeezed alongside them.
  gap_names <- c("Gap length", "No pause lik.")
  main <- tbl %>% filter(!measure %in% gap_names)
  gapp <- tbl %>% filter(measure %in% gap_names)

  p1 <- if (nrow(main)) coef_plot(main, term_labels) else NULL
  p2 <- if (nrow(gapp)) coef_plot(gapp, term_labels) else NULL
  p <- if (!is.null(p1) && !is.null(p2)) {
    p1 / p2 + patchwork::plot_layout(heights = c(1, 1))
  } else if (!is.null(p1)) {
    p1
  } else {
    p2
  }

  if (!is.null(figure_name) && !is.null(p)) {
    save_figure(
      p,
      figure_name,
      width = 7.2,
      height = if (!is.null(p1) && !is.null(p2)) 4.4 else 2.5
    )
  }
  print(p)
}


library(brms)
library(tidybayes)
dir.create("analysis/fits", recursive = TRUE, showWarnings = FALSE)

priors_lognormal <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 0.3), class = "b"),
  prior(normal(0, 0.3), class = "sd"),
  prior(lkj(2), class = "cor")
)

priors_hurdle <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 0.3), class = "b"),
  prior(normal(0, 0.3), class = "sd"),
  prior(lkj(2), class = "cor"),
  prior(normal(0, 1), class = "Intercept", dpar = "hu"),
  prior(normal(0, 1), class = "b", dpar = "hu"),
  prior(normal(0, 1), class = "sd", dpar = "hu")
)

priors_gaussian <- c(
  prior(normal(0, 500), class = "Intercept"),
  prior(normal(0, 100), class = "b"),
  prior(normal(0, 100), class = "sd"),
  prior(lkj(2), class = "cor")
)

code_half <- function(x, plus) {
  out <- ifelse(x == plus, 0.5, -0.5)
  if (length(unique(na.omit(out))) < 2) {
    stop(sprintf(
      "code_half(): '%s' matches nothing, so the contrast is constant. Levels present: %s",
      plus,
      paste(unique(x), collapse = ", ")
    ))
  }
  out
}

# Population-level terms, following the switches above. `cond` is the design
# factor for the experiment (c_tense or c_reg).
fixed_terms <- function(cond) {
  core <- if (fit_three_way) {
    sprintf("%s * c_prime * c_rprime", cond)
  } else if (fit_prime_interaction) {
    sprintf("%s * c_prime + c_prime * c_rprime", cond)
  } else {
    sprintf("%s * c_prime + c_rprime", cond)
  }
  paste(core, "+ z_freq")
}

# Every term the switches can produce, in display order. coef_plot() drops the
# ones a given fit does not contain.
prime_term_labels <- function(
  cond_label,
  cond_term,
  prime_label = "Tense priming",
  prime_short = "Priming",
  show_rprime = TRUE
) {
  # Only the terms that can appear in the reported model. Interactions with
  # regularity priming are never fitted; it is a control.
  #
  # `show_rprime` governs display, not the model. Where repetition of the
  # morphophonological form is a nuisance variable rather than the manipulation,
  # it is still estimated so that it cannot be absorbed by the terms of
  # interest, but it is left out of the coefficient table and figure, which
  # report the design.
  labels <- c(cond_label, prime_label)
  terms <- c(paste0("b_", cond_term), "b_c_prime")
  if (show_rprime) {
    labels <- c(labels, "Regularity priming")
    terms <- c(terms, "b_c_rprime")
  }
  setNames(
    c(labels, "Frequency", paste0(cond_label, " x ", prime_short)),
    c(terms, "b_z_freq", paste0("b_", cond_term, ":c_prime"))
  )
}

timing_measures <- c(
  "utt_onset",
  "det_dur",
  "np_onset",
  "subj_dur_pp",
  "verb_dur_pp",
  "gap"
)

phon_fit_data <- phon_analysis %>%
  filter(correct, prime_status != "burn") %>%
  mutate(
    c_reg = code_half(regularity, "IRREGULAR"),
    c_prime = code_half(prime_status, "primed"),
    sona_id = factor(sona_id),
    verb = factor(verb)
  ) %>%
  drop_na(utt_onset, z_freq, z_rate)

syn_fit_data <- syn_analysis %>%
  filter(correct, prime_status != "burn") %>%
  mutate(
    c_tense = code_half(tense, "future"),
    c_prime = code_half(prime_status, "primed"),
    c_rprime = code_half(prime_reg, "primed"),
    sona_id = factor(sona_id),
    verb = factor(verb)
  ) %>%
  drop_na(utt_onset, z_freq, z_rate)

concept_fit_data <- exp %>%
  filter(prime_status != "burn", rt_ok) %>%
  left_join(verb_freq %>% select(verb, zipf_past), by = "verb") %>%
  mutate(
    c_tense = code_half(tense, "future"),
    c_prime = code_half(prime_status, "primed"),
    c_rprime = code_half(prime_reg, "primed"),
    rank_freq = dense_rank(zipf_past),
    z_freq = as.numeric(scale(rank_freq)),
    z_prev = as.numeric(scale(prev_len)),
    subj = factor(subj),
    verb = factor(verb),
    entity = factor(entity)
  ) %>%
  drop_na(decision_rt, z_freq)
```

```{r}
# Helpers so every number in the prose is pulled from the fits and the data
# rather than typed in.
MOD <- "s5"

bb <- function(ex, dv, term, dpar = NULL) {
  f <- sprintf("analysis/fits/%s_%s_%s.rds", ex, MOD, dv)
  d <- posterior::as_draws_df(readRDS(f))
  v <- if (is.null(dpar)) paste0("b_", term) else paste0("b_", dpar, "_", term)
  x <- d[[v]]
  list(
    b = median(x),
    lo = quantile(x, .025),
    hi = quantile(x, .975),
    p = max(mean(x > 0), mean(x < 0)),
    pgt = mean(x > 0),
    pct = 100 * (exp(median(x)) - 1)
  )
}

# "b = 0.04, 95% CrI [0.01, 0.07], P(b > 0) = 0.99"
rep_b <- function(ex, dv, term, dpar = NULL, digits = 3) {
  z <- bb(ex, dv, term, dpar)
  sprintf(
    "β = %.*f, 95%% CrI [%.*f, %.*f], P(dir) = %s",
    digits,
    z$b,
    digits,
    z$lo,
    digits,
    z$hi,
    fmt_p(z$p)
  )
}

# "1090 ms (SE 15.6)"
ms <- function(df, v, ..., unit = " ms") {
  d <- df %>% filter(...)
  x <- d[[v]][!is.na(d[[v]])]
  sprintf("%.0f%s (SE %.1f)", mean(x), unit, sd(x) / sqrt(length(x)))
}
# ─── EFFECT DESCRIPTION ──────────────────────────────────────────────────────
# Turns a coefficient into a sentence: which way it runs, how big it is in
# milliseconds, and how much of the posterior stands behind it.

report_bf <- FALSE # Bayes factors: computed below, not printed, until the
# priors are settled

# Savage-Dickey BF10 against the point null. Needs only the draws and the
# analytic prior density at zero, so it runs on fits made without
# sample_prior = "yes". The value scales with the prior width, which is why it
# is switched off: the Normal(0, 100) prior on Experiment 3 reports evidence
# for the null almost whatever the data say, and would not be comparable with
# the Normal(0, 0.3) used on the log scale elsewhere.
bf10_savage_dickey <- function(fit, term) {
  ps <- brms::prior_summary(fit)
  pr <- ps$prior[ps$class == "b" & ps$coef == ""][1]
  sdp <- suppressWarnings(as.numeric(sub(
    ".*normal\\(0, *([0-9.eE+-]+)\\).*",
    "\\1",
    pr
  )))
  if (is.na(sdp)) {
    return(NA_real_)
  }
  x <- posterior::as_draws_df(fit)[[paste0("b_", term)]]
  if (is.null(x)) {
    return(NA_real_)
  }
  d0 <- stats::density(x, n = 4096)
  stats::dnorm(0, 0, sdp) / stats::approx(d0$x, d0$y, xout = 0)$y
}

jeffreys <- function(bf) {
  if (is.na(bf)) {
    return(NA_character_)
  }
  r <- if (bf < 1) 1 / bf else bf
  band <- if (r < 3) {
    "anecdotal"
  } else if (r < 10) {
    "moderate"
  } else if (r < 30) {
    "strong"
  } else if (r < 100) {
    "very strong"
  } else {
    "extreme"
  }
  paste0(band, if (bf < 1) " evidence against" else " evidence for")
}

# c_prime is tense repetition in Experiments 1 and 3 but form repetition in
# Experiment 2, and Experiment 2 codes regularity where the others code tense,
# so the level names are per experiment rather than global.
contrast_levels_for <- function(ex) {
  switch(
    ex,
    syn = list(
      c_tense = c(nice = "Tense", plus = "future", minus = "past"),
      c_prime = c(
        nice = "Tense priming",
        plus = "tense-primed",
        minus = "tense-unprimed"
      ),
      c_rprime = c(
        nice = "Regularity priming",
        plus = "form-primed",
        minus = "form-unprimed"
      )
    ),
    phon = list(
      c_reg = c(nice = "Regularity", plus = "irregular", minus = "regular"),
      c_prime = c(
        nice = "Regularity priming",
        plus = "form-primed",
        minus = "form-unprimed"
      ),
      c_rprime = c(
        nice = "Regularity priming",
        plus = "form-primed",
        minus = "form-unprimed"
      )
    ),
    concept = list(
      c_tense = c(nice = "Tense", plus = "future", minus = "past"),
      c_prime = c(
        nice = "Tense priming",
        plus = "tense-primed",
        minus = "tense-unprimed"
      ),
      c_rprime = c(
        nice = "Regularity priming",
        plus = "form-primed",
        minus = "form-unprimed"
      )
    ),
    stop("unknown experiment: ", ex)
  )
}

# The reminder that goes above every coefficient figure, built from the same
# coding the models use so it cannot drift from them.
sign_note <- function(
  ex,
  outcome = "onsets and durations",
  hide = character()
) {
  lv <- contrast_levels_for(ex)
  lv <- lv[!names(lv) %in% hide]
  # phon lists the same control twice; show each label once.
  keys <- names(lv)[!duplicated(vapply(lv, `[[`, character(1), "nice"))]
  parts <- vapply(
    keys,
    function(k) {
      sprintf(
        "%s +0.5 = %s, -0.5 = %s",
        lv[[k]][["nice"]],
        lv[[k]][["plus"]],
        lv[[k]][["minus"]]
      )
    },
    character(1)
  )
  sprintf(
    paste0(
      "*How to read the signs.* A positive coefficient means longer or ",
      "later %s; a negative one means shorter or earlier. Predictors are ",
      "sum coded at ±0.5, so each estimate is the difference between ",
      "its two levels: %s. An interaction is the difference between two ",
      "such differences."
    ),
    outcome,
    paste(parts, collapse = "; ")
  )
}

draws_of <- function(fit, term) {
  posterior::as_draws_df(fit)[[paste0("b_", term)]]
}

effect_stats <- function(fit, term) {
  x <- draws_of(fit, term)
  if (is.null(x)) {
    return(NULL)
  }
  q <- stats::quantile(x, c(.025, .975))
  list(
    m = mean(x),
    lo = q[[1]],
    hi = q[[2]],
    pd = max(mean(x > 0), mean(x < 0)),
    credible = q[[1]] > 0 || q[[2]] < 0
  )
}

# Coefficients are on the log scale, so exp(b) is a ratio; against the measure's
# observed mean that becomes a millisecond change. Approximate by construction.
as_ms <- function(b, baseline) baseline * (exp(b) - 1)
as_pct <- function(b) (exp(b) - 1) * 100
fmt_ci <- function(z) {
  sprintf("β = %.3f, 95%% CrI [%.3f, %.3f]", z$m, z$lo, z$hi)
}
verdict_of <- function(z) {
  if (z$credible) {
    "credible"
  } else if (z$pd >= pd_close) {
    "leaning"
  } else {
    "flat"
  }
}

bf_clause <- function(fit, term) {
  if (!report_bf) {
    return("")
  }
  b <- bf10_savage_dickey(fit, term)
  if (is.na(b)) {
    return("")
  }
  sprintf(" BF~10~ = %.2f, %s.", b, jeffreys(b))
}

# One sentence for a main effect.
describe_main <- function(fit, term, measure, baseline, ex) {
  z <- effect_stats(fit, term)
  if (is.null(z)) {
    return("")
  }
  lv <- contrast_levels_for(ex)[[term]]
  dir <- if (z$m > 0) "longer or later" else "shorter or earlier"
  if (verdict_of(z) == "flat") {
    return(sprintf(
      paste0(
        "%s separates %s from %s by at most about %.0f ms ",
        "(%.1f%%), and the posterior has no clear direction ",
        "(P(dir) = %.2f).%s"
      ),
      measure,
      lv[["plus"]],
      lv[["minus"]],
      abs(as_ms(z$m, baseline)),
      abs(as_pct(z$m)),
      z$pd,
      bf_clause(fit, term)
    ))
  }
  tail <- if (z$credible) {
    sprintf("The interval excludes zero (P(dir) = %.2f).", z$pd)
  } else {
    sprintf(
      "The interval still spans zero, but %.0f%% of the posterior is on one side.",
      100 * z$pd
    )
  }
  sprintf(
    "%s is %s in %s than in %s by about %.0f ms (%.1f%%). %s%s",
    measure,
    dir,
    lv[["plus"]],
    lv[["minus"]],
    abs(as_ms(z$m, baseline)),
    abs(as_pct(z$m)),
    tail,
    bf_clause(fit, term)
  )
}

# An interaction is only readable as two simple effects, so report those in
# milliseconds rather than the product term on its own.
describe_interaction <- function(
  fit,
  term,
  inner,
  outer,
  measure,
  baseline,
  ex
) {
  z <- effect_stats(fit, term)
  if (is.null(z)) {
    return("")
  }
  bi <- draws_of(fit, inner)
  x <- draws_of(fit, term)
  if (is.null(bi) || is.null(x)) {
    return("")
  }
  hi <- mean(bi + 0.5 * x)
  lo <- mean(bi - 0.5 * x)
  ol <- contrast_levels_for(ex)[[outer]]
  il <- contrast_levels_for(ex)[[inner]]
  bigger <- if (abs(hi) > abs(lo)) ol[["plus"]] else ol[["minus"]]
  stem <- sprintf(
    paste0(
      "The %s-versus-%s effect on %s is about %.0f ms when %s ",
      "and %.0f ms when %s"
    ),
    il[["plus"]],
    il[["minus"]],
    measure,
    as_ms(hi, baseline),
    ol[["plus"]],
    as_ms(lo, baseline),
    ol[["minus"]]
  )
  # Only claim amplification when the two halves are actually distinguishable;
  # otherwise the phrase dresses up a difference of a millisecond or two.
  if (verdict_of(z) == "flat") {
    return(sprintf(
      "%s, which is not a reliable difference (P(dir) = %.2f).%s",
      stem,
      z$pd,
      bf_clause(fit, term)
    ))
  }
  tail <- if (z$credible) {
    sprintf(
      "The difference between the two is reliable (P(dir) = %.2f).",
      z$pd
    )
  } else {
    sprintf(
      "%.0f%% of the posterior favours this pattern, though the interval spans zero.",
      100 * z$pd
    )
  }
  sprintf(
    "%s, so it is amplified in the %s. %s%s",
    stem,
    bigger,
    tail,
    bf_clause(fit, term)
  )
}

# Mean of a measure, for turning log coefficients into milliseconds.
baseline_of <- function(df, v) mean(df[[v]], na.rm = TRUE)

# Reads the reported model straight off disk, like bb() above.
fit_of <- function(ex, dv) {
  readRDS(sprintf("analysis/fits/%s_%s_%s.rds", ex, MOD, dv))
}

e1 <- exp %>% filter(prime_status != "burn", rt_ok)
e2 <- phon_analysis %>% filter(correct, prime_status != "burn")
e3 <- syn_analysis %>% filter(correct, prime_status != "burn")
prate <- function(df, ...) {
  d <- df %>% filter(...)
  sprintf(
    "%.1f%% (SE %.1f)",
    100 * mean(d$gap > 0),
    100 * sqrt(mean(d$gap > 0) * (1 - mean(d$gap > 0)) / nrow(d))
  )
}
```


Models of language production treat morphological features with a good deal of
uniformity. In the lemma model, morphosyntactic information is carried by
diacritics on lemma nodes, and the mechanism that assigns a diacritic is much the
same whether it encodes number, tense or case. Cue-based models are
architecturally different but inherit the same lexicalist premise: features are
properties of chunks retrieved from memory, and nothing in the architecture
distinguishes one kind of feature from another in its timing. Both make
morphological encoding dependent on lexical selection, and therefore subsequent
to it. A feature whose information source is available early is nonetheless held
behind the same bottleneck as one that is not.

Tense is a good place to test that premise, because its information source is
unusually early. Agreement reaches the verb through a linguistic mediator: the
discourse fixes the number of the subject, and the subject fixes the number of
the verb, so the verb's number cannot be specified until another element has been
planned. Tense has no such mediator. The speaker's communicative intention fixes
whether the event is past or future, and that is known from the very start of
planning. If the production system can specify a morphosyntactic feature without
first retrieving the word that carries it, tense is where it should be visible.

That makes it worth separating two things usually spoken about in one breath.
One is specification of the tense feature itself, marking the relevant node as
past or future as a function of the discourse. The other is retrieval of the
form that realises it, *spun* against *will spin*, which cannot be settled
without knowing which verb is being produced, since the shape of the past depends
on the verb's lexical identity. The second cannot precede lexical retrieval. The
first, on the account pursued here, can.

The three experiments isolate the levels in turn. Experiment 1 holds the task
constant and repeats the tense, so an effect early in the utterance would show
the feature being specified before the verb has been reached. Experiment 2 holds
tense constant, every event being past, and repeats the way the past is formed
instead; here an effect is expected late, around the verb. The contrast with
agreement is instructive: in agreement two lexical items bear on one closed-class
element, whereas here a single piece of information, past, meets a verb whose
lexical identity refracts it into a regular or an irregular surface form.
Experiment 3 repeats the same decision about time with no utterance at all, and
so tests whether a repetition effect needs linguistic material to act on.

The two accounts differ in where they place these operations.
`@fig-predictions`{=typst} lays them over the utterance. Under the first, the
tense feature is settled before articulation begins and leaves its trace at the
left, while the form is retrieved only as the verb is approached and leaves its
trace at the right. Under the second, the diacritic waits until the word carrying
it is about to be spoken, so both fall together in the pause before the verb and
neither leaves a trace earlier. The accounts agree from the verb onwards, so the
determiner and the subject noun are where they can be told apart.

```{=typst}
#figure(
  block(width: 100%, inset: (y: 4pt))[
    // The utterance, as areas along the timeline.
    #grid(
      columns: tl, column-gutter: 7pt,
      region(tone: 99%)[#text(fill: luma(45%), style: "italic")[before speech]],
      grid.cell(colspan: 2, region[*The*]),
      region[*Pirate*],
      region(tone: 99%)[#text(fill: luma(45%), size: 8pt)[pause]],
      grid.cell(colspan: 2, region[*will spin* / *spun*]),
      region[*a top*],
    )

    // Time, padded away from the words above and the brackets below.
    #v(7pt)
    #timearrow
    #v(9pt)

    // The two accounts, directly under the axis.
    #grid(
      columns: (2.4em,) + tl, column-gutter: 7pt, row-gutter: 9pt,
      align: horizon,

      hyp[H1:],
      grid.cell(colspan: 2, span(tone: 93%)[conceptual planning]),
      grid.cell(colspan: 2, span[*diacritic selection*]),
      grid.cell(colspan: 2, span[*form selection*]),
      blank, blank,

      hyp[H2:],
      grid.cell(colspan: 2, span(tone: 93%)[conceptual planning]),
      blank, blank,
      grid.cell(colspan: 2, span[*diacritic + form selection*]),
      blank, blank,
    )
  ],
  caption: [Two accounts of when tense is planned, laid over the utterance they
  predict about. Time runs left to right. Under H1 the tense feature is fixed
  early, while the sentence is still being begun, and the form that spells it out
  is retrieved only as the verb is approached. Under H2 both are resolved
  together, late, in the pause before the verb and just into it. The accounts
  make the same predictions from the verb onwards and differ only over the
  determiner and the subject noun, which is where they can be told apart.],
  kind: "quarto-float-fig",
  supplement: "Figure",
) <fig-predictions>
```

All timing measures are in milliseconds. Error bars are 95% intervals built on
Morey (2008) within-subject standard errors. Instead of reporting statistical significance, we report the posterior mean, the 95% credible interval, and the proportion of the posterior falling on one side of zero. An effect is called *credible* when the interval excludes zero, *leaning* when at least `r round(100 * pd_close)`% of the posterior is on one side but the interval still spans zero, and *flat* otherwise.

# Data and preprocessing

Experiments were carried out in PcIbex, and spoken responses were uploaded per block as `.webm`, converted to 16 kHz mono
wav, transcribed with AssemblyAI, and force-aligned against those transcripts
with the Montreal Forced Aligner (`english_us_arpa` acoustic model and
dictionary, out-of-vocabulary items passed through G2P). Word and phone¡™
boundaries come from the resulting TextGrids. The commands are listed in the
appendix. Recordings that never reached the server, and sessions with no usable audio, are
excluded and will not reported in any analysis below.

```{r, include=TRUE, results='asis'}
# 4c. Participant filtering write-up. Reflects whichever switches are set above.
exp_label <- function(e) c(phon = "morphophonology", syntax = "morphosyntax")[e]

n_start <- subject_stats %>% count(exp, name = "n")
n_end <- kept_subjects %>% count(exp, name = "n")

phrase <- function(df, unit = "participants") {
  paste(sprintf("%d %s %s", df$n, exp_label(df$exp), unit), collapse = " and ")
}

cat("## Participant filtering\n\n")

if (!filter_participants || nrow(filter_log) == 0) {
  cat(sprintf(
    paste(
      "No participant-level exclusions were applied. All %s with",
      "aligned recordings enter the analyses below.\n\n"
    ),
    phrase(n_start)
  ))
} else {
  cat(sprintf(
    paste(
      "Forced alignment yielded %s. "
    ),
    phrase(n_start)
  ))

  for (i in seq_len(nrow(distinct(filter_log, step)))) {
    s <- distinct(filter_log, step, desc)[i, ]
    rows <- filter_log %>% filter(step == s$step)
    counts <- paste(
      sprintf(
        "%d of %d in %s (%d remaining)",
        rows$n_removed,
        rows$n_before,
        exp_label(rows$exp),
        rows$n_after
      ),
      collapse = ", and "
    )
    cat(sprintf(
      # "**Step %d, %s.** Participants who %s were excluded: %s.\n\n",
      "Participants who %s were excluded: %s. ",
      # i,
      # tolower(s$step),
      s$desc,
      counts
    ))
  }

  cat(sprintf(
    "After all participant-level exclusions, %s remain.\n\n",
    phrase(n_end)
  ))
}

if (filter_trials) {
  trial_counts <- response_flags %>%
    semi_join(kept_subjects, by = c("exp", "sona_id")) %>%
    group_by(exp) %>%
    summarize(n = n(), kept = sum(utt_ok, na.rm = TRUE), .groups = "drop")

  cat(sprintf(
    paste(
      "Within the retained participants, individual trials were excluded",
      "when the response did not name the correct subject entity, did not",
      "use the target verb in the tense the trial called for, did not match",
      "the target sentence in length, was empty, or marked tense with an",
      "adverbial instead of on the verb. This retained %s.\n\n"
    ),
    paste(
      sprintf(
        "%d of %d trials (%.0f%%) in %s",
        trial_counts$kept,
        trial_counts$n,
        100 * trial_counts$kept / trial_counts$n,
        exp_label(trial_counts$exp)
      ),
      collapse = " and "
    )
  ))
} else {
  cat(
    "No trial-level exclusions were applied; all trials from retained participants enter the analyses.\n\n"
  )
}
```


# Descriptive results

The condition means for all three experiments are in
@fig-all-means. Experiments 1 and 2 share their five measures, so their
rows are directly comparable; Experiment 3 has only the keypress latency, and its
single panel is drawn to the width of one panel in the rows above it. The tables
behind these figures are in the appendix.

```{r fig-all-means, fig.height=8, fig.width=12, out.width="100%", fig.cap="Condition means for all three experiments. Top row, Experiment 1: timing by tense and tense repetition. Middle row, Experiment 2: timing by regularity and regularity repetition. Bottom row, Experiment 3: decision latency by tense and tense repetition, drawn at the width of a single panel above. Onsets and the gap are in milliseconds, the two durations in milliseconds per phoneme, and the decision latency in milliseconds. Filled markers are the primed level, open markers its baseline. Error bars are 95% intervals on Morey (2008) within-subject standard errors."}
p_syn <- condition_means_plot(
  syn_plot_data,
  tense,
  "Tense",
  c("past", "future")
)
p_phon <- condition_means_plot(
  phon_plot_data,
  regularity,
  "Regularity",
  c("IRREGULAR", "REGULAR")
)

p_concept <- exp %>%
  filter(prime_status != "burn", rt_ok) %>%
  mutate(measure = "Decision RT") %>%
  within_se(decision_rt, subj, tense, prime_status, strata = measure) %>%
  ggplot(aes(
    x = prime_status,
    y = m,
    shape = tense,
    fill = tense,
    linetype = tense,
    group = tense
  )) +
  facet_wrap(~measure, scales = "free_y", nrow = 1) +
  geom_line(position = pd, linewidth = 0.7, colour = "black") +
  geom_errorbar(
    aes(ymin = lwr, ymax = upr),
    position = pd,
    width = 0,
    linewidth = 0.7,
    colour = "black",
    linetype = "solid"
  ) +
  geom_point(position = pd, size = 2.5, colour = "black") +
  cond_shapes("Tense", c("past", "future")) +
  cond_fills("Tense", c("past", "future")) +
  cond_linetypes("Tense", c("past", "future")) +
  labs(x = "Prime status", y = "Time (ms)") +
  theme_paper(base_size = 9, legend_position = "bottom")

# Experiment 3 has one measure against six above it, so it is padded out to the
# width of a single panel rather than stretched across the page.
row_label <- function(p, txt) {
  p +
    labs(title = txt) +
    theme(plot.title = element_text(face = "bold", size = 9, hjust = 0))
}

save_figure(p_syn, "fig-exp1-syn-means", width = 7.2, height = 2.2)
save_figure(p_phon, "fig-exp2-phon-means", width = 7.2, height = 2.2)
save_figure(p_concept, "fig-exp3-concept-means", width = 3.2, height = 2.0)

(row_label(p_syn, "Experiment 1: Morphosyntax") /
  row_label(p_phon, "Experiment 2: Morphophonology") /
  (row_label(p_concept, "Experiment 3: Conceptual task") +
    patchwork::plot_spacer() +
    patchwork::plot_layout(widths = c(1, 5)))) +
  patchwork::plot_layout(heights = c(1, 1, 1), guides = "collect") &
  theme(legend.position = "bottom")
```


---



# Experiment 1: Morphosyntax

Producing a tensed sentence requires two decisions that can come apart. One is
which tense the sentence carries, a choice about its structure. The other is
which form realises that tense on the verb. If the two are separate operations
they should be able to leave their traces at different points in the utterance. This
experiment isolates the first: participants describe the same scenes in the past
or the future, and what is modulated whether the tense of the trial is being repeated from trial to trial.

## Prediction

The effects related to the repetition should surface early in the sentence.
Choosing between past and future is a choice about
the structure of the sentence, possibly made before articulation begins, so its effects
should appear at the start of the utterance rather than around the verb. The
interaction between tense and tense repetition at the onset of the subject noun is the
measure this predicts, and the later measures should show little.

## Participants

`r syn_c$n_subj` participants are analysed here. All were recruited through the
University of Maryland SONA subject pool and received course credit. Age, gender
and language background were collected at the start of the session.

The study was approved by the University of Maryland Institutional Review Board.
Participants read an information sheet describing the task, the recording of
their voice, and their right to stop at any point without penalty, and gave
informed consent before beginning. Responses are identified only by the pool
identifier needed to award credit and by a per-session token; no names or contact
details were collected, and the recordings and results analysed here carry no
information that identifies a participant directly.

## Materials

Three characters appear throughout: the Pirate, the Chef and the Wizard.
Eighteen verbs are divided into three sets of six, one set per block:

| Block | Verbs |
|:--|:--|
| 1 | drink, read, eat, paint, wash, push |
| 2 | build, sweep, ride, climb, stir, peel |
| 3 | blow, dig, shake, carry, play, smell |

Two further verbs, *spin* and *drag*, are held back for practice and never used in the main experiment.

Each block, first, familiarizes the event with the participant without the exact sentence to utter and without either the past or future tense marking:

```{=typst}
#example[The Pirate's spinning a top is in the past.]
```

Depending on the experimental condition, participants either produced past or the future marked information.

```{=typst}
#example[
  + The Pirate spun a top. #h(0.6em) #text(size: 9pt)[(past)]
  + The Pirate will spin a top. #h(0.6em) #text(size: 9pt)[(future)]
]
```

## Procedure

After consent, demographics and a microphone check, participants read the
instructions, learned two practice verbs and produced two practice sentences.

The main task is built from nine blocks, arranged in three meta-blocks of three.
A block is the unit that pairs learning with production. It opens with a
familiarisation phase: the six verbs of that block are presented one at a time
with audio, and then each is placed in time, so the participant learns both the
event and whether it is being described as past or future. Only after that phase
does the participant produce anything. The twelve production trials that follow
show a picture and ask for a spoken description, with no cue or reminder of the
tense on the screen; what the participant produces comes from what they learned
at the start of the block. Breaks fall between blocks, and the second and third
meta-blocks open with a screen announcing that the events now belong to a new
situation, because the same verbs recur with different characters and different
times.

Nine blocks of twelve trials give 108 recorded sentences per participant. The
eighteen verbs are covered once per meta-block, so each verb is produced six
times over the session.

`@fig-procedure`{=typst} lays out the session, one block, and one trial.

```{=typst}

#figure(
  block[
    // The session: three meta-blocks of three blocks each.
    #align(left, text(size: 8pt, weight: "bold")[The session])
    #v(3pt)
    #grid(
      columns: 9, column-gutter: 3pt, align: center + horizon,
      stage[#sub[Consent, demographics,] \ #sub[mic check, practice]], arw,
      stage(fill: luma(88%))[#sub[Meta-block 1] \ #sub[3 blocks]], arw,
      stage(fill: luma(88%))[#sub[Meta-block 2] \ #sub[3 blocks]], arw,
      stage(fill: luma(88%))[#sub[Meta-block 3] \ #sub[3 blocks]], arw,
      stage[#sub[Upload,] \ #sub[exit page]],
    )
    #v(4pt)
    #note[nine blocks of twelve trials, 108 recorded sentences]
    #v(12pt)

    // One block: learning first, production after.
    #align(center, block(
      stroke: 0.6pt + luma(55%), radius: 3pt, inset: (x: 9pt, y: 7pt), width: 100%,
      [
        #align(left, text(size: 8pt, weight: "bold")[One block: learn six verbs, then produce twelve sentences])
        #v(4pt)
        #grid(
          columns: 7, column-gutter: 3pt, align: center + horizon,
          stage(fill: luma(94%))[#sub[*Familiarisation*] \ #sub[6 verbs, one at a time, \ with audio]],
          arw,
          stage(fill: luma(94%))[#sub[*Times taught*] \ #sub[each event placed in \ past or future]],
          arw,
          stage[#sub[Ready screen]],
          arw,
          stage(fill: luma(78%))[#sub[*12 production trials*] \ #sub[picture only, no cue]],
        )
        #v(8pt)

        // The twelve trials, and which are analysed.
        #grid(
          columns: (3.6em,) + (2.5em,) * 12, column-gutter: 2pt, row-gutter: 2pt,
          align: center + horizon,
          text(size: 7.5pt)[Position], ..range(1, 13).map(i => text(size: 7.5pt)[#i]),
          text(size: 7.5pt)[Pattern 1],
          ..("PAST", "FUT", "FUT", "PAST", "PAST", "FUT",
             "FUT", "PAST", "PAST", "FUT", "FUT", "PAST").enumerate().map(((i, t)) => {
            cell(fill: if i in (0, 5, 6, 11) { luma(80%) } else { white })[#t]
          }),
          text(size: 7.5pt)[Pattern 2],
          ..("FUT", "PAST", "PAST", "FUT", "FUT", "PAST",
             "PAST", "FUT", "FUT", "PAST", "PAST", "FUT").enumerate().map(((i, t)) => {
            cell(fill: if i in (0, 5, 6, 11) { luma(80%) } else { white })[#t]
          }),
        )
      ]
    ))
    #v(12pt)

    // One production trial.
    #align(center, block(
      stroke: 0.6pt + luma(55%), radius: 3pt, inset: (x: 9pt, y: 7pt),
      [
        #align(left, text(size: 8pt, weight: "bold")[One production trial])
        #v(4pt)
        #grid(
          columns: 7, rows: (auto, auto), column-gutter: 4pt, row-gutter: 3pt,
          align: center + horizon,
          stage[Blank screen], arw,
          stage(fill: luma(88%))[Fixation cross], arw,
          stage(fill: luma(78%))[*Picture* \ #sub[recording runs]], arw,
          stage[Continue],

          note[400 ms], [],
          note[600 ms], [],
          note[4500 ms, stops itself], [],
          note[self-paced],
        )
      ]
    ))
  ],
  caption: [Procedure for Experiment 1. Learning and production are separate
  phases of a block, not steps within a trial: the six verbs of a block are
  taught first, with audio and with each event placed in past or future, and the
  twelve production trials that follow show only a picture, with nothing on the
  screen naming the tense. Those twelve trials are two passes over the block's
  six verbs, one under each tense order, with the order of the passes randomised;
  shaded positions are boundary trials, dropped because they have no comparable
  predecessor, and every other trial is primed when it repeats the tense of the
  trial before it and unprimed when it does not. Experiment 2 has the same shape
  with regularity in place of tense and no time-teaching step, and Experiment 3
  replaces the spoken response with a keypress.],
  kind: "quarto-float-fig",
  supplement: "Figure",
) <fig-procedure>
```

Which pattern runs first is randomised per block, block order is shuffled within
each meta-block, the assignment of characters to verbs rotates across meta-blocks,
and the ordering avoids placing the same character on consecutive trials.

## Analysis

Five measures are taken from the forced alignment of each recording: utterance
onset, the onset of the subject noun, the per-phoneme duration of the subject
noun, the per-phoneme duration of the verb, and the gap running from the offset
of the subject noun to the tense-bearing word, which is *will* in the future and
the past-marked verb in the past. Experiment 2 uses the same five measures;
Experiment 3 has no audio and measures keypress latency instead.

Each measure is modelled separately. The terms of interest are tense, tense
repetition, and their interaction. Repetition of the morphophonological form is
carried alongside as a control, because a trial that repeats the tense often
repeats the form as well. Frequency, the subject noun and articulation rate are
also controlled. The appendix gives the contrast coding, the priors, the full
random-effect structure and the convergence checks.

Of `r syn_c$total-syn_c$wrong` trials, `r syn_c$burn` were boundary trials that was not included in the data, leaving
**`r syn_c$included` trials** from `r syn_c$n_subj` participants.

The model estimates each measure separately. The measures are the time from the
start of the trial to the first word, the time to the subject noun, the duration
of the subject noun per phone, the duration of the verb per phone, and the
interval between the subject noun and the verb.

`r sign_note("syn", hide = "c_rprime")`

Speakers reach the verb without pausing on most trials, so the interval is
modelled in two parts. **Gap length** is how long the pause lasts on the trials where
there is one. **No pause lik.** is how likely the speaker is to run
straight into the verb; a positive coefficient there means fewer pauses. The
first part is on the log scale in milliseconds, the second on the log-odds scale.


## Results



```{r, results='asis', fig.height=4, fig.width=8, fig.cap="Experiment 1: population-level coefficients of the reported model. Points are posterior medians, bars 95% credible intervals; the dashed line marks zero. The figure at the right of each panel is the proportion of the posterior on one side of zero."}
syn_ladder <- fit_ladder(syn_fit_data, timing_measures, specs_tense, "syn")

# ── Reported model: spec 5 (design x prime + freq + reg-prime + entity + rate)
syn_ladder_main <- pick_headline(syn_ladder)

report_models(
  syn_ladder_main,
  # The design is tense and tense repetition; form repetition is a control and
  # is estimated but not reported here.
  prime_term_labels("Tense", "c_tense", show_rprime = FALSE),
  sprintf(
    paste0(
      "Experiment 1, reported model (spec 5). Positive = slower/longer. ",
      "Bold marks terms with P(dir) at or above %.2f."
    ),
    pd_close
  ),
  figure_name = "fig-exp1-coefs-main",
  hide_terms = c("c_rprime", "entityPirate", "entityWizard"),
  what = "figure"
)
```


Tense itself changed the shape of the utterance from the determiner onwards
(@fig-all-means, top row). Speakers took longer over the determiner before a future sentence than before a
past one, `r ms(e3, "det_dur", tense == "future")` against
`r ms(e3, "det_dur", tense == "past")`, and the model confirms the difference
(`r rep_b("syn", "det_dur", "c_tense")`). This is the earliest point at which
tense could show up at all, and *the* is the same word in both conditions, so
neither the lexical identity of the word nor its frequency can be responsible.
The subject noun was then reached later in the future,
`r ms(e3, "np_onset", tense == "future")` against
`r ms(e3, "np_onset", tense == "past")`
(`r rep_b("syn", "np_onset", "c_tense")`), was itself produced faster,
`r ms(e3, "subj_dur_pp", tense == "future", unit = " ms/phone")` against
`r ms(e3, "subj_dur_pp", tense == "past", unit = " ms/phone")`
(`r rep_b("syn", "subj_dur_pp", "c_tense")`), and the verb was drawn out,
`r ms(e3, "verb_dur_pp", tense == "future", unit = " ms/phone")` against
`r ms(e3, "verb_dur_pp", tense == "past", unit = " ms/phone")`
(`r rep_b("syn", "verb_dur_pp", "c_tense")`). The raw means for verb duration
are almost identical across the two tenses, and the model separates them only
once frequency is controlled: future trials carry the bare form, which is far
more frequent than the past form and therefore shorter, and that difference
masks the effect of tense in the unadjusted means.

Repeating the tense from the previous trial did not speed anything up. If
anything the numerically primed trials were slower rather than faster, and no
main effect of tense repetition was reliable on any measure, including utterance
onset (`r rep_b("syn", "utt_onset", "c_prime")`) and the onset of the subject
noun (`r rep_b("syn", "np_onset", "c_prime")`).

Repetition did, however, interact with tense, and it did so early. At the onset
of the subject noun, repeating the tense delayed the future
(`r ms(e3, "np_onset", tense == "future", prime_status == "primed")` primed
against `r ms(e3, "np_onset", tense == "future", prime_status == "unprimed")`
unprimed) while the past ran in the opposite direction
(`r ms(e3, "np_onset", tense == "past", prime_status == "primed")` primed against
`r ms(e3, "np_onset", tense == "past", prime_status == "unprimed")` unprimed).
The interaction is in the predicted direction and most of the posterior supports
it, though the interval still includes zero
(`r rep_b("syn", "np_onset", "c_tense:c_prime")`). The same asymmetry appears in
whether speakers paused before the tense-bearing word at all: tense repetition
made a pause less likely in the future than in the past
(`r rep_b("syn", "gap", "c_tense:c_prime", dpar = "hu")`), which is the strongest
interaction anywhere in the experiment.

Neither interaction survives past the subject noun. The duration of the noun
(`r rep_b("syn", "subj_dur_pp", "c_tense:c_prime")`) and the duration of the verb
(`r rep_b("syn", "verb_dur_pp", "c_tense:c_prime")`) separate the conditions by
about a millisecond, with the posterior split almost evenly around zero on the
verb. Repeating a tense therefore leaves its trace before the verb is reached and
nothing at the verb itself, which is where an effect of selecting the tense
feature, rather than of retrieving the form that spells it out, should appear.
The effect is a lean rather than an established result: with
`r n_distinct(e3$sona_id)` participants the sample is no longer the limiting
factor, so the width of the interval reflects the size of the effect.


# Experiment 2: Morphophonology

The companion to Experiment 1. Here the tense is held constant and the form is
varied instead, so that what repeats across trials is not the choice of tense
but the way the past is spelled out on the verb. If selecting a tense feature
and retrieving its form are separate operations, repetition of the form should
act at a different point in the utterance from repetition of the tense.

## Prediction

Repetition should act near the verb. The form of the past tense is not needed
until the verb is articulated, so preparing it should leave the time to begin
speaking unchanged and appear on the words immediately around the verb. The
effect should be larger for irregular verbs, whose past forms have to be
retrieved individually, than for regular verbs, which share one ending. The
clearest place to see this is the interaction between regularity and repetition,
on the duration of the subject noun and on whether the speaker pauses before
reaching the verb.

## Participants

`r phon_c$n_subj` participants are analysed here, recruited through the same
University of Maryland SONA subject pool as Experiment 1 and credited for
participation. No participant took part in more than one experiment. Approval,
consent and anonymisation were as described for Experiment 1.

## Materials

The characters and the eighteen verbs are those of Experiment 1, in the same
three blocks. Ten of the twenty verbs form their past irregularly (*blow, build,
dig, drink, eat, read, ride, shake, spin, sweep*) and the rest are regular. Each
block holds exactly three of each, so regularity can repeat or switch from trial
to trial without being confounded with block.

Every response is in the past tense:

```{=typst}
#example[The Pirate spun a top.]
```

## Procedure

The session ran as in Experiment 1, with the same nine blocks in three
meta-blocks and the same split between a familiarisation phase and the twelve
production trials that follow it. The familiarisation is shorter here: the six
verbs of the block are presented one at a time with audio, but nothing places
them in time, because every event is past. Blocks are separated by breaks, and
the session again yields 108 recorded sentences per participant.
`@fig-procedure-phon`{=typst} gives the block.

```{=typst}
#figure(
  block(width: 100%)[
    #nests
    #v(9pt)
    #grid(
      columns: 5, column-gutter: 3pt, align: center + horizon,
      stage(fill: luma(94%))[#sub[*Familiarisation*] \ #sub[6 verbs with audio, \ no times taught]],
      arw,
      stage[#sub[Ready screen]],
      arw,
      stage(fill: luma(78%))[#sub[*12 production trials*] \ #sub[picture only]],
    )
    #v(7pt)
    #grid(
      columns: (3.6em,) + (2.5em,) * 12, column-gutter: 2pt, row-gutter: 2pt,
      align: center + horizon,
      text(size: 7.5pt)[Position], ..range(1, 13).map(i => text(size: 7.5pt)[#i]),
      text(size: 7.5pt)[Pattern 1],
      ..("I", "R", "R", "I", "I", "R", "R", "I", "I", "R", "R", "I").enumerate().map(((i, t)) => {
        cell(fill: if i in (0, 5, 6, 11) { luma(80%) } else { white })[#t]
      }),
      text(size: 7.5pt)[Pattern 2],
      ..("R", "I", "I", "R", "R", "I", "I", "R", "R", "I", "I", "R").enumerate().map(((i, t)) => {
        cell(fill: if i in (0, 5, 6, 11) { luma(80%) } else { white })[#t]
      }),
    )
  ],
  caption: [One block of Experiment 2, which is one ninth of the session:
  three blocks make a meta-block and three meta-blocks make the session.
  I is an irregular verb and R a regular
  one. Two passes over the same six verbs, one under each order, with the order
  of the passes randomised; shaded positions are boundary trials. Tense is never
  taught, since every event is past. The session, the trial timings and the
  randomisation are otherwise those of Experiment 1.],
  kind: "quarto-float-fig",
  supplement: "Figure",
) <fig-procedure-phon>
```

Which pattern runs first is randomised per block, block order is shuffled within
each meta-block, the assignment of characters to verbs rotates across
meta-blocks, and the ordering avoids placing the same character on consecutive
trials.

A trial is **primed** when its verb forms the past in the same way as the verb on
the trial before it and **unprimed** when it does not. As in Experiment 1, the
first and last position of each six-trial pattern have no comparable predecessor
and are excluded as boundary trials.

## Analysis

The five measures are those of Experiment 1. Each is modelled separately, and
the terms of interest are regularity, regularity repetition, and their
interaction. Tense plays no role here, since every trial is past. Frequency, the
subject noun and articulation rate are controlled. The appendix gives the
contrast coding, the priors, the random-effect structure and the convergence
checks.

Of `r phon_c$total` trials, `r phon_c$wrong` (`r phon_c$pct_wrong`%) failed the
utterance screen and `r phon_c$burn` were boundary trials, leaving
**`r phon_c$included` trials** from `r phon_c$n_subj` participants.

The model estimates each measure separately. The measures are the time from the
start of the trial to the first word, the time to the subject noun, the duration
of the subject noun per phone, the duration of the verb per phone, and the
interval between the subject noun and the verb.

`r sign_note("phon")`

Speakers reach the verb without pausing on most trials, so the interval is
modelled in two parts. **Gap length** is how long the pause lasts on the trials where
there is one. **No pause lik.** is how likely the speaker is to run
straight into the verb; a positive coefficient there means fewer pauses. The
first part is on the log scale in milliseconds, the second on the log-odds scale.


## Results



```{r, results='asis', fig.height=4, fig.width=8, fig.cap="Experiment 2: population-level coefficients of the reported model. Points are posterior medians, bars 95% credible intervals; the dashed line marks zero. The figure at the right of each panel is the proportion of the posterior on one side of zero."}
phon_ladder <- fit_ladder(phon_fit_data, timing_measures, specs_reg, "phon")

# ── Reported model: spec 5 (design x prime + freq + reg-prime + entity + rate)
phon_ladder_main <- pick_headline(phon_ladder)

report_models(
  phon_ladder_main,
  # c_prime is already regularity repetition here, so there is no separate
  # form-repetition control to report.
  prime_term_labels(
    "Regularity",
    "c_reg",
    prime_label = "Regularity priming",
    prime_short = "Priming",
    show_rprime = FALSE
  ),
  sprintf(
    paste0(
      "Experiment 2, reported model (spec 5). Positive = slower/longer. ",
      "Bold marks terms with P(dir) at or above %.2f."
    ),
    pd_close
  ),
  figure_name = "fig-exp2-coefs-main",
  hide_terms = c("entityPirate", "entityWizard"),
  what = "figure"
)
```


Regularity on its own did almost nothing (@fig-all-means, middle row).
Irregular and regular verbs did not
differ at utterance onset (`r rep_b("phon", "utt_onset", "c_reg")`), at the onset
of the subject noun (`r rep_b("phon", "np_onset", "c_reg")`), or in the duration
of the subject noun (`r rep_b("phon", "subj_dur_pp", "c_reg")`). The verb itself
was numerically longer when irregular, but the interval is wide and includes zero
(`r rep_b("phon", "verb_dur_pp", "c_reg")`).

Repeating the way the past is formed did leave a trace, and it appeared at the
verb. Verbs were drawn out when the preceding trial formed its past the same way,
`r ms(e2, "verb_dur_pp", prime_status == "primed", unit = " ms/phone")` against
`r ms(e2, "verb_dur_pp", prime_status == "unprimed", unit = " ms/phone")`, with
most of the posterior on that side
(`r rep_b("phon", "verb_dur_pp", "c_prime")`). This is where the account expects
form repetition to act: the shape of the past is not needed until the verb is
spoken.

The same repetition also delayed the start of the utterance,
`r ms(e2, "utt_onset", prime_status == "primed")` against
`r ms(e2, "utt_onset", prime_status == "unprimed")`
(`r rep_b("phon", "utt_onset", "c_prime")`), which the account does not predict.
Nothing about the form of a verb that has not been reached should govern when a
speaker begins to talk, so this effect sits awkwardly with the claim that form
repetition acts late. Both effects are leans rather than established results, and
they are the same size, so the experiment does not on its own separate an early
locus from a late one.

The interaction between regularity and repetition was expected to be clearest on
the words leading up to the verb, and the largest one in the experiment is indeed
on the duration of the subject noun
(`r rep_b("phon", "subj_dur_pp", "c_reg:c_prime")`). Repetition lengthened the
noun more before an irregular verb than before a regular one, which is the
direction predicted if an irregular past has to be retrieved individually while
regulars share one ending. It does not reach conventional credibility, and no
comparable interaction appears at utterance onset
(`r rep_b("phon", "utt_onset", "c_reg:c_prime")`) or on the verb itself
(`r rep_b("phon", "verb_dur_pp", "c_reg:c_prime")`).


# Experiment 3: Conceptual task

The control for the other two. Experiments 1 and 2 both find their effects by
repeating something across trials, and both require speech. This experiment
repeats the same decision about time without asking for any utterance, so it
tests whether a repetition effect needs linguistic material to act on at all.

## Prediction

Repetition should have no effect here. If the repetition effects in the
production experiments come from reusing linguistic material, then a task that
requires no verb form to be produced gives nothing to reuse. Participants decide
which time a picture refers to and press a key. Deciding about the future may
take longer than deciding about the past, but repeating a decision about the same
time should not speed up the next one, and the effect of repetition should not
differ between the two tenses. A repetition effect here would mean the effects in
the other two experiments need not be linguistic at all.

## Participants

`r concept_n_subj` participants are analysed here, recruited through the same
University of Maryland SONA subject pool as the production experiments. Approval,
consent and anonymisation were as described for Experiment 1; nothing was
recorded here, so the session produced no audio.

## Materials

The same three characters and the same twenty verbs as Experiments 1 and 2, with
each character paired with each verb and a time assigned to that pairing. Because
nothing is spoken, the verb enters only through the picture and the learned
pairing, not through a produced form.

## Procedure

Participants learned the character, verb and time pairings, then made speeded
keypress decisions about which time a picture referred to, pressing **C** for past
and **M** for future. Trials were arranged in twelve-trial blocks on the same
scheme as the production experiments, so that tense repeats or switches from one
trial to the next in the same way. A trial is **primed** when its tense matches
the previous trial's and **unprimed** when it switches, and positions 1, 6, 7 and
12 of each block are boundary trials, as in `@fig-procedure`{=typst}.
`@fig-procedure-concept`{=typst} gives the trial.

```{=typst}
#figure(
  block(width: 100%)[
    #nests
    #v(9pt)
    #grid(
      columns: 5, column-gutter: 4pt, rows: (auto, auto),
      align: center + horizon,
      stage(fill: luma(94%))[#sub[*Learning*] \ #sub[character, verb and time]],
      arw,
      stage(fill: luma(88%))[#sub[Picture]],
      arw,
      stage(fill: luma(78%))[#sub[*Keypress*] \ #sub[C = past, M = future]],

      note[once per block], [], note[no speech], [], note[latency measured],
    )
  ],
  caption: [One trial of Experiment 3, and the learning that precedes it. A
  block teaches its pairings once and then runs twelve trials; three blocks
  make a meta-block and three meta-blocks make the session, as in the
  production experiments. The pairings are learned first, as in the
  production experiments, and the trial itself asks only for a keypress. Nothing
  is spoken, so the articulatory measures of Experiments 1 and 2 do not apply.],
  kind: "quarto-float-fig",
  supplement: "Figure",
) <fig-procedure-concept>
```

## Analysis

The single measure is the latency of the keypress reporting the decision. The
terms of interest are tense, tense repetition, and their interaction. No speech
is recorded, so the utterance screen and the articulatory controls used in the
other two experiments do not apply.

Of `r concept_total` trials, `r concept_burn` boundary trials (positions 1, 6, 7
and 12 of each 12-trial block) were excluded. A further `r concept_rt_trimmed`
trials fell outside the `r rt_min`–`r rt_max` ms response window and were dropped
as anticipations or disengagement: the raw responses run from 0 ms to 173 s, and
the upper tail distorts both the means and the lognormal fit. That leaves
**`r concept_included` trials**.

Experiment 3 has a single dependent measure, the latency of the keypress
reporting the decision, so one model is fitted rather than a set over measures.
The task involves no speech, so none of the articulatory measures used in
Experiments 1 and 2 apply here.

`r sign_note("concept", "decision latencies")`


## Results

Regularity repetition is also tracked. The two are not independent: regularity
repeats on `r round(100 * concept_rrep_given_tprime)`% of trials where tense
repeats but `r round(100 * concept_rrep_given_tswitch)`% of trials where it
switches, so the two priming effects are estimated jointly in one model rather
than separately; the cell means are in the appendix.


```{r, results='asis', fig.height=1.8, fig.width=4.4, out.width="62%", fig.cap="Experiment 3: population-level coefficients of the reported model for decision latency. Points are posterior medians, bars 95% credible intervals. The figure at the right is the proportion of the posterior on one side of zero."}
concept_ladder <- fit_ladder(
  concept_fit_data,
  "decision_rt",
  specs_tense_norate,
  "concept",
  subject = "subj"
)

# ── Reported model: spec 5 (design x prime + freq + reg-prime + entity + rate)
concept_ladder_main <- pick_headline(concept_ladder)

report_models(
  concept_ladder_main,
  prime_term_labels("Tense", "c_tense"),
  sprintf(
    paste0(
      "Experiment 3, reported model (spec 5). Positive = slower/longer. ",
      "Bold marks terms with P(dir) at or above %.2f."
    ),
    pd_close
  ),
  figure_name = "fig-exp3-coefs-main",
  hide_terms = c("entityPirate", "entityWizard"),
  what = "figure"
)
```


Decisions were slower for future than for past (@fig-all-means, bottom row).
Future trials averaged
`r ms(e1, "decision_rt", tense == "future")` and past trials
`r ms(e1, "decision_rt", tense == "past")`. The model supports this difference
(`r rep_b("concept", "decision_rt", "c_tense")`).

Tense repetition did not affect decision times. Primed trials averaged
`r ms(e1, "decision_rt", prime_status == "primed")` and unprimed trials
`r ms(e1, "decision_rt", prime_status == "unprimed")`, and the model gives
`r rep_b("concept", "decision_rt", "c_prime")`. The interaction between tense and
tense repetition is absent
(`r rep_b("concept", "decision_rt", "c_tense:c_prime")`), with the posterior
centred on zero.

This is what we expected. The conceptual task requires participants to retrieve
which time a picture refers to and press a key, with no verb form to produce.
Deciding about the future takes longer than deciding about the past, but
repeating a decision about the same time does not make the next one faster.


---


# What the three experiments show

Repeating a tense left its mark before the verb was reached and nothing at the
verb itself. The interaction between tense and tense repetition ran in the
predicted direction at the onset of the subject noun, and repetition made a pause
before the tense-bearing word less likely in the future than in the past; neither
the duration of the subject noun nor the duration of the verb separated the
conditions. Tense itself was already visible on the determiner, a word identical
in both conditions, which is the earliest point at which it could show up at all.
That is the pattern H1 predicts and H2 does not.

Repeating the form gave a weaker and less well-placed result. The interaction
between regularity and repetition, which is the term that separates a form that
has to be retrieved individually from one that shares an ending, fell on the
duration of the subject noun and not on the verb. Nothing comparable appeared at
the verb itself. Experiment 3 produced no interaction anywhere, so whatever the
production experiments are picking up is not a generic effect of repeating a
decision.

Main effects of repetition are left out of this summary. Both experiments show
some, but a main effect says only that repetition changed a measure, not when the
information was planned, and the two accounts make the same prediction about it.
Only the interactions distinguish them.

`@fig-evidence`{=typst} places these results on the same timeline as the
predictions.

```{=typst}
#let found(body, tone: 82%) = box(
  width: 100%, fill: luma(tone), radius: 2pt,
  stroke: 0.8pt + luma(30%), inset: (x: 3pt, y: 6pt),
  align(center, text(size: 7.5pt, body))
)

#figure(
  block(width: 100%, inset: (y: 4pt))[
    #grid(
      columns: tl, column-gutter: 7pt,
      region(tone: 99%)[#text(fill: luma(45%), style: "italic")[before speech]],
      grid.cell(colspan: 2, region[*The*]),
      region[*Pirate*],
      region(tone: 99%)[#text(fill: luma(45%), size: 8pt)[pause]],
      grid.cell(colspan: 2, region[*will spin* / *spun*]),
      region[*a top*],
    )
    #v(7pt)
    #timearrow
    #v(9pt)

    #grid(
      columns: (5.2em,) + tl, column-gutter: 7pt, row-gutter: 9pt,
      align: horizon,

      align(right + horizon, text(size: 8pt, weight: "bold")[Tense × \ repetition]),
      grid.cell(colspan: 3, found[NP onset \ #text(size: 7pt)[$P_"dir"$ = .87]]),
      blank,
      found[pause less likely \ #text(size: 7pt)[$P_"dir"$ = .93]],
      blank, blank, blank,

      align(right + horizon, text(size: 8pt, weight: "bold")[Regularity × \ repetition]),
      blank, blank, blank,
      found[subject duration \ #text(size: 7pt)[$P_"dir"$ = .94]],
      blank, blank, blank, blank,
    )
    #v(7pt)
    #align(left, text(size: 7.5pt, fill: luma(30%))[
      Only interactions are shown: a main effect of repetition, on its own, does
      not tell the two accounts apart. Tense repetition acts to the left of the
      verb and not on it, as H1 predicts.
    ])
  ],
  caption: [Where the effects actually fell, on the timeline of
  `@fig-predictions`{=typst}. Only the interactions are shown, since these are
  what the two accounts disagree about; each band gives the proportion of the
  posterior on one side of zero. Every one is a lean rather than a credible
  effect, as no interval excluded zero. Regions left blank are ones where the
  measure was taken and no interaction was found. Experiment 3 is not shown: it
  produced no interaction anywhere.],
  kind: "quarto-float-fig",
  supplement: "Figure",
) <fig-evidence>
```


---



# A1. Model coefficients {.unnumbered}

The tables behind the coefficient figures in each Results section. Estimates are
on the log scale, so a coefficient is a proportional change; positive means
slower or longer. Bold marks terms with P(dir) at or above the threshold set in
the header of this document. Terms held in the models as controls but left out of
the figures are also left out here.

### Experiment 1: Morphosyntax {.unnumbered}

```{r}
#| label: tbl-coefs-exp1
#| tbl-cap: !expr sprintf("Experiment 1, reported model (spec 5). Positive = slower/longer. Bold marks terms with P(dir) at or above %.2f.", pd_close)
report_models(
  syn_ladder_main,
  prime_term_labels("Tense", "c_tense", show_rprime = FALSE),
  caption = NULL,
  hide_terms = c("c_rprime", "entityPirate", "entityWizard"),
  what = "table",
  split = FALSE
)
```

### Experiment 2: Morphophonology {.unnumbered}

```{r}
#| label: tbl-coefs-exp2
#| tbl-cap: !expr sprintf("Experiment 2, reported model (spec 5). Positive = slower/longer. Bold marks terms with P(dir) at or above %.2f.", pd_close)
report_models(
  phon_ladder_main,
  prime_term_labels(
    "Regularity",
    "c_reg",
    prime_label = "Regularity priming",
    prime_short = "Priming",
    show_rprime = FALSE
  ),
  caption = NULL,
  hide_terms = c("entityPirate", "entityWizard"),
  what = "table",
  split = FALSE
)
```

### Experiment 3: Conceptual task {.unnumbered}

```{r}
#| label: tbl-coefs-exp3
#| tbl-cap: !expr sprintf("Experiment 3, reported model (spec 5). Positive = slower/longer. Bold marks terms with P(dir) at or above %.2f.", pd_close)
report_models(
  concept_ladder_main,
  prime_term_labels("Tense", "c_tense"),
  caption = NULL,
  hide_terms = c("entityPirate", "entityWizard"),
  what = "table",
  split = FALSE
)
```


---



# A2. Condition means {.unnumbered}

The tables behind the figures in the Descriptive results section.

### Experiment 1: Morphosyntax {.unnumbered}

```{r}
#| label: tbl-means-exp1
#| tbl-cap: "Experiment 1: timing measures by tense and tense repetition. Onsets and the gap are in milliseconds; subject-noun and verb durations are milliseconds per phoneme. SE is the Morey (2008) within-subject standard error."
means_table(
  syn_plot_data,
  tense,
  NULL
)
```

### Experiment 2: Morphophonology {.unnumbered}

```{r}
#| label: tbl-means-exp2
#| tbl-cap: "Experiment 2: timing measures by regularity and regularity repetition. Onsets and the gap are in milliseconds; subject-noun and verb durations are milliseconds per phoneme. SE is the Morey (2008) within-subject standard error."
means_table(
  phon_plot_data,
  regularity,
  NULL
)
```

### Experiment 3: Conceptual task {.unnumbered}

```{r}
#| label: tbl-means-exp3
#| tbl-cap: "Experiment 3: decision RT by tense and tense repetition. SE is the standard error of the mean."
exp %>%
  filter(prime_status != "burn", rt_ok) %>%
  group_by(Tense = tense, `Prime status` = prime_status) %>%
  summarize(
    N = n(),
    `Mean RT (ms)` = round(mean(decision_rt, na.rm = TRUE)),
    SE = round(sd(decision_rt, na.rm = TRUE) / sqrt(n()), 1),
    .groups = "drop"
  ) %>%
  kable()
```

```{r}
#| label: tbl-means-exp3-reg
#| tbl-cap: "Experiment 3: decision RT by regularity repetition and tense repetition, showing how far the two priming variables co-vary."
exp %>%
  filter(prime_status != "burn", prime_reg != "burn", rt_ok) %>%
  group_by(`Regularity prime` = prime_reg, `Tense prime` = prime_status) %>%
  summarize(
    N = n(),
    `Mean RT (ms)` = round(mean(decision_rt, na.rm = TRUE)),
    SE = round(sd(decision_rt, na.rm = TRUE) / sqrt(n()), 1),
    .groups = "drop"
  ) %>%
  kable()
```


---


# A3. By-item effects {.unnumbered}

By-verb deviations from the intercept in the reported model of each
experiment. A verb whose interval excludes zero behaves differently from the
rest of the item set.

### Experiment 1: Morphosyntax {.unnumbered}

```{r, fig.height=2.5, fig.width=8, fig.cap="Experiment 1: by-verb deviations from the population intercept in the reported model. A verb whose interval excludes zero behaves differently from the rest of the item set."}
p_items <- item_forest(syn_ladder_main, "Experiment 1")
if (!is.null(p_items)) {
  save_figure(p_items, "fig-exp1-item-forest", width = 7.2, height = 4)
  p_items
}
```


### Experiment 2: Morphophonology {.unnumbered}

```{r, fig.height=2.5, fig.width=8, fig.cap="Experiment 2: by-verb deviations from the population intercept in the reported model."}
p_items <- item_forest(phon_ladder_main, "Experiment 2")
if (!is.null(p_items)) {
  save_figure(p_items, "fig-exp2-item-forest", width = 7.2, height = 4)
  p_items
}
```


### Experiment 3: Conceptual task {.unnumbered}

```{r, fig.height=2.2, fig.width=3.6, out.width="55%", fig.cap="Experiment 3: by-verb deviations from the population intercept in the reported model."}
p_items <- item_forest(concept_ladder_main, "Experiment 3")
if (!is.null(p_items)) {
  save_figure(p_items, "fig-exp3-item-forest", width = 4.4, height = 2.6)
  p_items
}
```


---


# A4. Model comparison and checks {.unnumbered}

### Model comparison {.unnumbered}

Models were compared by leave-one-out cross-validation. Differences are relative
to the best model for that measure; a difference smaller than about twice its
standard error is not meaningful.

```{r, results='asis'}
if (run_loo) {
  compare_ladder(concept_ladder, "Experiment 3: model comparison")
  compare_ladder(phon_ladder, "Experiment 2: model comparison")
  compare_ladder(syn_ladder, "Experiment 1: model comparison")
} else {
  cat(
    "\n*Model comparison is switched off. Set `run_loo <- TRUE` to compute it.*\n\n"
  )
}
```

### Convergence {.unnumbered}

No divergent transitions, R-hat at or below 1.01, and effective sample sizes
in the thousands for every model reported.

```{r, results='asis'}
convergence_table(concept_ladder_main, "Experiment 3: convergence")
convergence_table(phon_ladder_main, "Experiment 2: convergence")
convergence_table(syn_ladder_main, "Experiment 1: convergence")
```

### Posterior predictive checks {.unnumbered}

Each panel compares the observed distribution of a measure (dark line) against
data simulated from the fitted model (light lines).

```{r, fig.height=2, fig.width=8, fig.cap="Experiment 2: posterior predictive checks. Simulated datasets from the fitted model against the observed distribution of each measure."}
pp_grid(phon_ladder_main, "Experiment 2")
```

```{r, fig.height=2, fig.width=7.2, fig.cap="Experiment 1: posterior predictive checks. Simulated datasets from the fitted model against the observed distribution of each measure."}
pp_grid(syn_ladder_main, "Experiment 1")
```

```{r, fig.height=1.8, fig.width=4.4, out.width="62%", fig.cap="Experiment 3: posterior predictive checks for decision latency."}
pp_grid(concept_ladder_main, "Experiment 3")
```


# A5. Statistical models {.unnumbered}

All models were fitted with brms (Stan), four chains of 4000 iterations with the
first 1000 discarded, `adapt_delta` = 0.95.

### Contrast coding {.unnumbered}

All categorical predictors are sum coded at $\pm 0.5$, so each coefficient is the
difference between the two levels and the intercept is the grand mean.

| predictor | +0.5 | -0.5 | used in |
|---|---|---|---|
| Tense | future | past | Experiments 1 and 3 |
| Regularity | irregular | regular | Experiment 2 |
| Tense priming | primed | unprimed | Experiments 1 and 3 |
| Regularity priming | primed | unprimed | all three |

In Experiment 2 regularity priming is the manipulation and appears as *Priming*
in the tables. In Experiments 1 and 3 the manipulation is tense priming, and
regularity priming enters only as a control.

The subject noun (Chef, Pirate, Wizard) is treated as a factor with Chef as the
reference level. Frequency is the rank of the produced word form in the SUBTLEX
Zipf scale, from least to most frequent, centred and divided by its standard
deviation. Ranking rather than the Zipf value itself keeps a small number of very
frequent verbs from dominating an otherwise flat predictor. Articulation rate is
the number of phones per second on that trial, computed over the words that are
not being measured, and is also centred and scaled.

### Response distributions {.unnumbered}

Onsets and durations are positive and right skewed and were fitted with a shifted
lognormal distribution. The shift estimates a floor below which no response
occurs.

The interval between the subject noun and the verb is zero on
`r sprintf("%.0f%%", 100*mean(e2$gap == 0))` of trials in Experiment 2 and
`r sprintf("%.0f%%", 100*mean(e3$gap == 0))` in Experiment 3, because speakers
often move into the verb without pausing. A single continuous distribution
describes this badly, so it was fitted with a hurdle lognormal: one part gives the
probability of producing no pause, the other the duration of the pause when there
is one. Both parts take the same predictors, which allows repetition to change how
often speakers pause separately from how long they pause. Coefficients labelled as
being on the probability of no pause come from the first part, where a positive
value means fewer pauses.

### Model structure {.unnumbered}

Every measure was fitted with the same structure. For Experiments 1 and 3,

$$
\text{measure} \sim \text{Tense} \times \text{Tense priming} + \text{Regularity priming} + \text{Frequency} + \text{Character} + \text{Rate}
$$
$$
+ \; (1 + \text{Tense} \times \text{Tense priming} + \text{Regularity priming} \mid \text{participant}) + (1 + \text{Tense priming} \mid \text{verb})
$$

and for Experiment 2,

$$
\text{measure} \sim \text{Regularity} \times \text{Priming} + \text{Frequency} + \text{Character} + \text{Rate}
$$
$$
+ \; (1 + \text{Regularity} \times \text{Priming} \mid \text{participant}) + (1 + \text{Priming} \mid \text{verb})
$$

Experiment 1 has no recordings and therefore no rate term.


Each measure was fitted with the same population-level terms: the design factor
(tense or regularity), repetition, their interaction, frequency, the subject noun,
articulation rate, and, in Experiments 1 and 3, form repetition as a control.
Random effects are by participant, with slopes for the design factor, repetition
and their interaction, and by verb, with a slope for repetition. Frequency, the
subject noun and articulation rate vary between items or within a trial and take
no by-participant slope.

Regularity priming is included because it is not independent of tense priming in
these materials. The form repeats on
`r sprintf("%.0f%%", 100*mean((e3 %>% filter(prime_status=="primed", prime_reg!="burn"))$prime_reg == "primed"))`
of trials where tense repeats and
`r sprintf("%.0f%%", 100*mean((e3 %>% filter(prime_status=="unprimed", prime_reg!="burn"))$prime_reg == "primed"))`
of trials where it switches, so leaving it out would let part of a form-repetition
effect appear as a tense priming effect. It is never entered into an
interaction.

### Priors {.unnumbered}

| parameter | prior |
|---|---|
| intercept | Normal(0, 1) |
| population-level coefficients | Normal(0, 0.3) |
| random-effect standard deviations | Normal(0, 0.3) |
| random-effect correlations | LKJ(2) |
| hurdle intercept and coefficients | Normal(0, 1) |

Coefficients are on the log scale, so Normal(0, 0.3) places most of its mass on
effects smaller than a factor of two and rules out nothing the data could show.

### Reporting {.unnumbered}

Each coefficient is reported as its posterior median, a 95% credible interval,
and P(dir), the proportion of the posterior lying on the same side of zero as
the median. P(dir) runs from 0.5, meaning the posterior is split evenly and
the sign is undetermined, to 1, meaning the whole posterior falls on one side.
It is not a significance test and has no threshold attached; we describe an
effect as supported when most of the posterior is on one side and the interval is
narrow enough to be informative. Values are never printed as 0 or 1, since a
finite posterior sample cannot establish either.

### Convergence and model checking {.unnumbered}

Every model reported here converged: no divergent transitions, R-hat at or below 1.01,
and bulk effective sample sizes above 400. Posterior predictive checks compare
simulated data against the observed distribution for each measure. Posterior
contraction, the reduction in variance from prior to posterior, is above 0.94 for
the frequency, subject noun and articulation rate terms and between 0.56 and 0.99
for the interactions, so the posteriors are determined by the data rather than by
the priors.

Models were compared by leave-one-out cross-validation. Adding the subject noun
and articulation rate improves prediction on the duration measures by a wide
margin, which is why they are in the reported model: the three subject nouns
differ by more than 50 ms per phone, far more than any effect of the manipulation.


# A6. Recordings and measurement {.unnumbered}

Spoken responses were recorded in the browser, one file per trial, and uploaded
in blocks. Recordings were converted to 16 kHz mono audio and transcribed
automatically. Each transcript was then aligned to its recording with the
Montreal Forced Aligner using its English acoustic model and pronunciation
dictionary, with pronunciations generated for words outside the dictionary. The
alignment gives the start and end time of every word and every phone, and all
timing measures are read from it.

Speech onset is the start of the first word. The onset of the subject noun and
its duration are taken from the second word. The interval before the verb runs
from the end of the subject noun to the start of the verb, and is zero when the
two are adjacent. The verb is located by matching the transcript against the
form the trial called for rather than by position, since the future adds a word
before it.

Durations of the subject noun and the verb are divided by the number of phones
the aligner placed inside them. Words differ in length, and without this a longer
word would count as slower speech. Articulation rate is the number of phones per
second over the remaining words of the sentence, which leaves the two measured
words out of their own control variable.

Transcripts were scored against the sentence the trial called for. A response was
kept if it named the right character and used the right verb in the right tense.
Material after the verb was not scored, since every measure lies at or before the
verb. Two adjustments were made after inspecting the transcripts. The past tense
of *blow* is a homophone of *blue* and was transcribed as the colour on most
trials, so both spellings were accepted. *Blowed* was kept as an error, being a
real regularisation rather than a transcription problem. Responses that marked
time with an adverbial instead of on the verb, such as *the wizard climbs a
ladder in the past*, were excluded, since they leave nothing for the verb to
carry.

Participants were excluded if fewer than half their responses met these criteria.
