---
title: "Tense Timing Experiments: Analysis & Report"
format:
  pdf:
    toc: true
    toc-depth: 3
    fontsize: 10pt
    geometry: margin=0.5in
    pdf-engine: xelatex
    mainfont: "Helvetica Neue"
    sansfont: "Helvetica Neue"
execute:
  echo: false
  warning: false
  message: false
  fig-align: center
  fig-width: 7
  fig-height: 2.5
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(include = FALSE)
# The default pdf device cannot embed Helvetica Neue and errors on any figure
# that draws text; cairo_pdf handles it.
knitr::opts_chunk$set(dev = "cairo_pdf")
library(ggplot2)
theme_set(theme_classic(base_size = 13, base_family = "Helvetica Neue"))
```

```{r}
# ─── FILTERING SWITCHES ──────────────────────────────────────────────────────
# Every filter below is optional. Set a switch to FALSE to skip that step; the
# reported participant counts and the write-up in "Participant filtering"
# follow whatever is set here.

filter_participants <- TRUE # master switch for all participant-level exclusion
filter_low_accuracy <- TRUE # drop participants below min_subject_accuracy
filter_adverbial <- FALSE # drop participants over max_adverbial_rate
filter_incomplete <- FALSE # drop participants with a failed upload chunk

filter_trials <- TRUE # drop individual trials failing the utterance screen

fit_models <- TRUE # TRUE fits the brms models (slow; cached to
# analysis/fits/*.rds, so it is a one-off cost)

min_subject_accuracy <- 0.50 # min proportion of trials passing all trial-level flags
max_adverbial_rate <- 0.20 # max proportion of "... in the past/future" responses
n_trials_full <- 108 # trials per participant in the CSV (all participants)

# Experiment 1 is a speeded keypress task; RTs outside this window are either
# anticipations or disengagement. The raw range runs from 0 ms to 173 s.
rt_min <- 200 # ms
rt_max <- 5000 # ms

# Which priming interactions to estimate. The two priming variables are
# correlated, so their interaction asks whether tense priming depends on the
# morphophonological form also repeating. The three-way term additionally asks
# whether that dependency differs by tense, which the smaller experiments cannot
# support: Exp 3 has 771 analysed trials over 15 participants, so a three-way
# split leaves under 100 trials per cell.
fit_prime_interaction <- FALSE # tense prime x regularity prime
fit_three_way <- FALSE # ... additionally crossed with tense
# ─────────────────────────────────────────────────────────────────────────────

library(tidyverse)
library(janitor)
library(scales)
library(patchwork)

# Todo: make sure that it saves the subject id in the recording name.
# Todo: check if you can burn just the first 1 and 12.
#  in addition to primed and unprimed, can you make primed to be F->F Prime/ T->T Prime, and also T->F Switch, F->T Switch. And also same thing with the irregulars. But do the prime x tense first, and then do prime type x tense later in the plots.
```


```{r}
# User-defined function to read in PCIbex Farm results files
read.pcibex <- function(
  filepath,
  auto.colnames = TRUE,
  fun.col = function(col, cols) {
    cols[cols == col] <- paste(col, "Ibex", sep = ".")
    return(cols)
  }
) {
  n.cols <- max(count.fields(filepath, sep = ",", quote = NULL), na.rm = TRUE)
  if (auto.colnames) {
    cols <- c()
    con <- file(filepath, "r")
    while (TRUE) {
      line <- readLines(con, n = 1, warn = FALSE)
      if (length(line) == 0) {
        break
      }
      m <- regmatches(line, regexec("^# (\\d+)\\. (.+)\\.$", line))[[1]]
      if (length(m) == 3) {
        index <- as.numeric(m[2])
        value <- m[3]
        if (is.function(fun.col)) {
          cols <- fun.col(value, cols)
        }
        cols[index] <- value
        if (index == n.cols) {
          break
        }
      }
    }
    close(con)
    return(read.csv(
      filepath,
      comment.char = "#",
      header = FALSE,
      col.names = cols
    ))
  } else {
    return(read.csv(
      filepath,
      comment.char = "#",
      header = FALSE,
      col.names = seq(1:n.cols)
    ))
  }
}


subj_id <- function(
  df,
  cols = c("results_reception_time", "md5_hash_of_participant_s_ip_address"),
  label = "subj"
) {
  lookup <- df %>%
    distinct(across(all_of(cols))) %>%
    mutate(!!label := paste0("S[", row_number(), "]"))

  df %>% left_join(lookup, by = cols)
}

# ─── PLOTTING STYLE ──────────────────────────────────────────────────────────
# Grayscale print figures, matching the insensitivity/SPR paper: conditions are
# carried by shape + fill + linetype rather than colour, all ink is black, error
# bars are capless, and the minor grid is dropped.

theme_paper <- function(base_size = 13, legend_position = "right") {
  theme_classic(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.line = element_line(colour = "black", linewidth = 0.6),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", size = base_size - 2),
      legend.position = legend_position,
      legend.key.width = unit(1.2, "cm")
    )
}

pd <- position_dodge(width = 0.3)

# Filled marker = the marked/primed level, open marker = its baseline.
cond_shapes <- function(name, levels) {
  scale_shape_manual(name = name, values = setNames(c(16, 21), levels))
}
cond_fills <- function(name, levels) {
  scale_fill_manual(name = name, values = setNames(c("black", "white"), levels))
}
cond_linetypes <- function(name, levels) {
  scale_linetype_manual(
    name = name,
    values = setNames(c("solid", "22"), levels)
  )
}

# Morey (2008) within-subject SE: remove each subject's own mean, add the group
# mean back, then apply the sqrt(M/(M-1)) correction for M within-subject cells.
#
# `strata` names a variable whose levels are on different scales and must not be
# pooled (here `measure`: an onset in seconds and a duration in seconds are not
# comparable quantities). Normalisation, the cell count M, and the correction are
# all computed separately within each stratum. `...` gives the condition
# variables that define the cells.
within_se <- function(df, value, subject, ..., strata = NULL) {
  df %>%
    group_by({{ strata }}) %>%
    mutate(
      .grand = mean({{ value }}, na.rm = TRUE),
      .cells = n_distinct(paste(!!!enquos(...)))
    ) %>%
    group_by({{ strata }}, {{ subject }}) %>%
    mutate(.norm = {{ value }} - mean({{ value }}, na.rm = TRUE) + .grand) %>%
    group_by({{ strata }}, ...) %>%
    summarize(
      m = mean({{ value }}, na.rm = TRUE),
      n = sum(!is.na(.norm)),
      .cells = first(.cells),
      se = sd(.norm, na.rm = TRUE) /
        sqrt(n) *
        sqrt(.cells / max(.cells - 1, 1)),
      .groups = "drop"
    ) %>%
    mutate(lwr = m - 1.96 * se, upr = m + 1.96 * se) %>%
    select(-.cells)
}

save_figure <- function(
  plot,
  name,
  width = 6.2,
  height = 2.6,
  dir = "analysis/fig_output"
) {
  dir.create(dir, recursive = TRUE, showWarnings = FALSE)
  ggsave(
    file.path(dir, paste0(name, ".pdf")),
    plot,
    width = width,
    height = height,
    device = cairo_pdf
  )
  invisible(plot)
}
# ─────────────────────────────────────────────────────────────────────────────

add_trial_info <- function(
  df,
  subject_col = "subj",
  block_col = "block",
  tense_col = "tense",
  n_cycle = 12,
  burn_pos = c(1, 6, 7, 12)
) {
  df %>%
    group_by(.data[[subject_col]]) %>%
    mutate(trial_n = row_number()) %>%
    group_by(.data[[subject_col]], .data[[block_col]]) %>%
    mutate(
      block_trial_n = (row_number() - 1) %% n_cycle + 1,
      prime_status = case_when(
        block_trial_n %in% burn_pos ~ "burn",
        .data[[tense_col]] == lag(.data[[tense_col]]) ~ "primed",
        TRUE ~ "unprimed"
      )
    ) %>%
    ungroup()
}
```


```{r}
concept <- read.pcibex("data/concept_june6.csv") %>% clean_names() %>% subj_id()
exp <- concept %>%
  filter(penn_element_name == "tenseChoice") %>%
  select(
    subj,
    label,
    choice = value,
    assigned_list,
    block,
    verb,
    regularity,
    tense,
    entity,
    on_left = right_opt,
    on_right = practice_correct,
    decision_rt
  ) %>%
  # Tense repetition is the primary priming variable; regularity repetition is
  # tracked alongside it because the two are correlated in this sequence.
  add_trial_info() %>%
  rename(prime_status_main = prime_status) %>%
  add_trial_info(tense_col = "regularity") %>%
  rename(prime_reg = prime_status, prime_status = prime_status_main) %>%
  # Case differs across the three CSVs: concept logs "PAST"/"FUTURE" and
  # "Irregular"/"Regular", the alignments use lowercase tense from the file
  # stem, and phon logs "IRREGULAR"/"REGULAR". Normalise here so that scales,
  # contrast coding and tables match across experiments.
  mutate(
    tense = str_to_lower(tense),
    regularity = str_to_upper(regularity),
    decision_rt = as.double(decision_rt)
  ) %>%
  mutate(
    rt_ok = !is.na(decision_rt) & decision_rt >= rt_min & decision_rt <= rt_max
  ) %>%
  # Carry-over control, the Experiment 1 analogue of the previous utterance's
  # length: how long the preceding decision took.
  arrange(subj, trial_n) %>%
  group_by(subj) %>%
  mutate(prev_len = lag(if_else(rt_ok, decision_rt, NA_real_))) %>%
  ungroup()
```



```{r}
# 4. Read aligned TextGrids + join PCIbex metadata per experiment
library(readtextgrid)
library(fs)

read_tgs <- function(aligned_dir, exp_name) {
  tg_files <- dir_ls(aligned_dir, recurse = TRUE, glob = "*.TextGrid")
  if (length(tg_files) == 0) {
    return(tibble())
  }

  map_dfr(tg_files, \(fp) {
    tg <- read_textgrid(fp)
    stem <- path_file(fp) %>% str_remove("\\.TextGrid$")
    sona_id <- path_file(path_dir(fp))

    # Parse stem: resp_m{mb}_block{bl}_p{pass}_{verb}_{tense}
    parts <- str_match(
      stem,
      "^resp_m(\\d+)_block(\\d+)_p(\\d+)_([a-z]+)_(past|future|pres)$"
    )

    tg %>%
      filter(text != "") %>%
      transmute(
        exp = exp_name,
        sona_id,
        recording = stem,
        metablock = parts[, 2],
        block = parts[, 3],
        pass = parts[, 4],
        verb = parts[, 5],
        tense = parts[, 6],
        tier = tier_name,
        word_phone = text,
        t_start = xmin,
        t_end = xmax,
        duration = xmax - xmin
      )
  })
}

read_pcibex_meta <- function(filepath, prime_col = "tense", also_prime = NULL) {
  read.pcibex(filepath) %>%
    clean_names() %>%
    subj_id() %>%
    filter(parameter == "Filename", str_detect(value, "^resp")) %>%
    mutate(
      sona_id = as.character(sona_id_url),
      recording = str_remove(value, "\\.webm$")
    ) %>%
    select(
      sona_id,
      recording,
      block,
      list = assigned_list,
      subj,
      tense,
      regularity,
      entity,
      target_sentence = target_canonical_sentence
    ) %>%
    mutate(regularity = str_to_upper(regularity)) %>%
    add_trial_info(tense_col = prime_col) %>%
    # A second, non-primary priming variable: whether the OTHER factor repeated
    # from the preceding trial. Burn positions are defined on block position, so
    # both priming columns mark the same trials as "burn".
    {
      if (is.null(also_prime)) {
        .
      } else {
        rename(., prime_status_main = prime_status) %>%
          add_trial_info(tense_col = also_prime) %>%
          rename(prime_reg = prime_status, prime_status = prime_status_main)
      }
    } %>%
    select(-block, -tense) # drop: conflicts with stem-parsed columns in alignments
}

# phon: priming is by regularity (regular vs irregular morphophonology). Tense is
#       constant (all past), so tense priming is undefined here.
# syn:  priming is by tense (past vs future); regularity repetition is tracked
#       alongside it as prime_reg.
phon_meta <- read_pcibex_meta("data/phon_june6.csv", prime_col = "regularity")
syn_meta <- read_pcibex_meta(
  "data/syn_june6.csv",
  prime_col = "tense",
  also_prime = "regularity"
)

alignments <- bind_rows(
  read_tgs("analysis/mfa_aligned/phon", "phon") %>%
    left_join(phon_meta, by = c("sona_id", "recording")),
  read_tgs("analysis/mfa_aligned/syntax", "syntax") %>%
    left_join(syn_meta, by = c("sona_id", "recording"))
) %>%
  filter(str_detect(recording, "^resp"))

glimpse(alignments)
```


```{r}
# 4b. Response accuracy screen
#
# Trial-level flags compare the spoken word sequence (MFA words tier, i.e. the
# ASR transcript) against the TargetCanonicalSentence logged by PCIbex for that
# trial. Subjects failing the screen are dropped from `alignments` here, so every
# downstream chunk works on retained subjects only.

past_forms <- c(
  blow = "blew",
  build = "built",
  carry = "carried",
  climb = "climbed",
  dig = "dug",
  drag = "dragged",
  drink = "drank",
  eat = "ate",
  paint = "painted",
  peel = "peeled",
  play = "played",
  push = "pushed",
  read = "read",
  ride = "rode",
  shake = "shook",
  smell = "smelled",
  spin = "spun",
  stir = "stirred",
  sweep = "swept",
  wash = "washed"
)

resp_spoken <- alignments %>%
  filter(tier == "words") %>%
  group_by(exp, sona_id, recording, verb, tense, entity, target_sentence) %>%
  arrange(t_start, .by_group = TRUE) %>%
  summarize(spoken_raw = paste(word_phone, collapse = " "), .groups = "drop")

response_flags <- resp_spoken %>%
  mutate(
    said = str_squish(str_to_lower(str_remove_all(spoken_raw, "[^a-zA-Z ]"))),
    target_words = str_squish(str_to_lower(str_remove_all(
      target_sentence,
      "[^a-zA-Z ]"
    ))),
    past_target = coalesce(unname(past_forms[verb]), paste0(verb, "ed")),

    # correct number of words
    n_words_ok = str_count(said, "\\S+") == str_count(target_words, "\\S+"),
    # correct head: the subject entity for this trial is named
    head_ok = !is.na(entity) &
      str_detect(said, paste0("\\b", str_to_lower(entity), "\\b")),
    # correct verb, in the inflection the trial called for
    verb_ok = case_when(
      # "blew" and "blue" are perfect homophones: the ASR cannot distinguish them
      # and writes the colour on most blow trials. Accepting it recovers correct
      # productions. "blowed" stays an error, being a real over-regularisation
      # rather than a transcription artefact.
      tense == "past" ~ str_detect(said, paste0("\\b", past_target, "\\b")) |
        (verb == "smell" & str_detect(said, "\\bsmelt\\b")) |
        (verb == "blow" & str_detect(said, "\\bblue\\b")),
      TRUE ~ str_detect(said, paste0("\\b", verb, "\\b"))
    ),
    # correct tense marking on the verb itself
    tense_ok = case_when(
      tense == "past" ~ verb_ok & !str_detect(said, "\\bwill\\b"),
      tense == "future" ~ verb_ok & str_detect(said, "\\bwill\\b"),
      TRUE ~ NA
    ),
    # adverbial register: tense carried by "in the past/future" instead of the
    # verb, e.g. "the wizard climbs a ladder in the future". No tense planning on
    # the verb, so these trials cannot speak to the question and are excluded.
    adverbial = str_detect(said, "\\bin the (past|future)\\b"),
    empty_resp = said %in% c("", "sil"),
    exact = said == target_words,

    # Scoring stops at the verb. Every measure (utterance onset, NP onset,
    # subject duration, gap, verb duration) lies at or before the verb, so what
    # follows it is irrelevant. Word count stays as a diagnostic but no longer
    # excludes trials: it rejected correct productions differing only in the
    # object's determiner, and the four items whose target object is not "a/an"
    # (blow/bubbles, drink/coffee, sweep/the floor, play/the guitar) absorbed
    # nearly all of that loss.
    utt_ok = head_ok &
      verb_ok &
      tense_ok &
      !adverbial &
      !empty_resp
  ) %>%
  select(
    exp,
    sona_id,
    recording,
    exact,
    n_words_ok,
    head_ok,
    verb_ok,
    tense_ok,
    adverbial,
    empty_resp,
    utt_ok
  )

subject_stats <- response_flags %>%
  group_by(exp, sona_id) %>%
  summarize(
    n_trials = n(),
    accuracy = mean(utt_ok, na.rm = TRUE),
    adverbial_rate = mean(adverbial, na.rm = TRUE),
    empty_rate = mean(empty_resp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Every participant has 108 trials in the CSV. Those below that lost one
  # upload chunk (one metablock-3 block, both passes = 12 trials); their data
  # are incomplete, not duplicated.
  mutate(complete_session = n_trials == n_trials_full)

# Sequential exclusion cascade. Each step sees only the participants that
# survived the earlier ones, so the counts add up to the final N.
filter_steps <- list(
  # Adverbial register first: it is a task-compliance criterion, and a response
  # that marks tense adverbially carries no tense planning on the verb whatever
  # its accuracy. Running it before the accuracy screen keeps the two counts
  # interpretable rather than letting accuracy absorb these participants.
  list(
    on = filter_adverbial,
    name = "Adverbial register",
    desc = sprintf(
      paste(
        "marked tense with \"in the past/future\" rather than on",
        "the verb on more than %.0f%% of trials"
      ),
      100 * max_adverbial_rate
    ),
    keep = function(d) d$adverbial_rate <= max_adverbial_rate
  ),
  list(
    on = filter_low_accuracy,
    name = "Low accuracy",
    desc = sprintf(
      "passed fewer than %.0f%% of trials on the utterance criteria",
      100 * min_subject_accuracy
    ),
    keep = function(d) d$accuracy >= min_subject_accuracy
  ),
  list(
    on = filter_incomplete,
    name = "Incomplete session",
    desc = sprintf(
      "had fewer than %d recordings, from a failed upload chunk",
      n_trials_full
    ),
    keep = function(d) d$complete_session
  )
)

alive <- subject_stats
filter_log <- tibble()
excluded <- tibble()

for (st in filter_steps) {
  if (!filter_participants || !st$on) {
    next
  }
  keep <- st$keep(alive)
  filter_log <- bind_rows(
    filter_log,
    alive %>%
      mutate(.keep_it = keep) %>%
      group_by(exp) %>%
      summarize(
        step = st$name,
        desc = st$desc,
        n_before = n(),
        n_removed = sum(!.keep_it),
        n_after = sum(.keep_it),
        .groups = "drop"
      )
  )
  excluded <- bind_rows(
    excluded,
    alive %>% filter(!keep) %>% mutate(step = st$name)
  )
  alive <- alive %>% filter(keep)
}

kept_subjects <- alive
subject_screen <- subject_stats %>%
  mutate(
    subject_ok = paste(exp, sona_id) %in%
      paste(kept_subjects$exp, kept_subjects$sona_id)
  )

if (filter_participants) {
  alignments <- alignments %>%
    semi_join(kept_subjects, by = c("exp", "sona_id"))
}

# Criterion-by-criterion pass rates, before any exclusion
response_flags %>%
  group_by(exp) %>%
  summarize(
    n = n(),
    across(
      c(
        exact,
        n_words_ok,
        head_ok,
        verb_ok,
        tense_ok,
        utt_ok,
        adverbial,
        empty_resp
      ),
      ~ mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) %>%
  print()

if (nrow(excluded)) {
  excluded %>% arrange(exp, accuracy) %>% print(n = Inf)
}
```




```{r}
# 5. Syntax + Phon: correct utterances + timing measures
# Toggle correctness criterion:
#   strict = TRUE  → spoken words must exactly match target sentence
#   strict = FALSE → utt_ok from the 4b screen (word count + head + verb + tense,
#                    excluding adverbial-register and empty responses)
strict_matching <- FALSE

syn_words <- alignments %>%
  filter(exp == "syntax", tier == "words") %>%
  group_by(sona_id, recording) %>%
  arrange(t_start, .by_group = TRUE) %>%
  mutate(word_n = row_number()) %>%
  ungroup()

# Reconstruct spoken form + tense-based correctness
# past   → word 3 is NOT "will" (speaker used a past form)
# future → word 3 IS "will"
syn_spoken <- syn_words %>%
  group_by(sona_id, recording) %>%
  summarize(spoken = paste(word_phone, collapse = " "), .groups = "drop")

syn_w3 <- syn_words %>%
  filter(word_n == 3) %>%
  select(sona_id, recording, w3 = word_phone)

syn_check <- syn_words %>%
  distinct(
    sona_id,
    recording,
    tense,
    verb,
    entity,
    regularity,
    subj,
    list,
    trial_n,
    block_trial_n,
    prime_status,
    prime_reg,
    target_sentence
  ) %>%
  left_join(syn_spoken, by = c("sona_id", "recording")) %>%
  left_join(syn_w3, by = c("sona_id", "recording")) %>%
  left_join(
    response_flags %>%
      select(
        sona_id,
        recording,
        utt_ok,
        adverbial,
        n_words_ok,
        head_ok,
        verb_ok,
        tense_ok
      ),
    by = c("sona_id", "recording")
  ) %>%
  mutate(
    target_words = str_to_lower(str_remove_all(
      target_sentence,
      "[^a-zA-Z ]"
    )) %>%
      str_squish(),
    correct = if (!filter_trials) {
      TRUE
    } else if (strict_matching) {
      spoken == target_words
    } else {
      coalesce(utt_ok, FALSE)
    }
  )

# Timing: pivot words 1–3 wide (the | entity | will/verb)
syn_timing <- syn_words %>%
  filter(word_n <= 3) %>%
  select(sona_id, recording, word_n, t_start, t_end) %>%
  pivot_wider(
    names_from = word_n,
    values_from = c(t_start, t_end),
    names_sep = "_w"
  )

# Join and compute measures
syn_analysis <- syn_check %>%
  left_join(syn_timing, by = c("sona_id", "recording")) %>%
  mutate(
    utt_onset = t_start_w1, # onset of "the"
    np_onset = t_start_w2,
    np_head_dur = t_end_w2 - t_start_w2, # duration of entity noun (NP head)
    np_offset = t_end_w2, # offset of full subject NP
    crit_onset = t_start_w3, # onset of "will" (future) or past verb
    gap = t_start_w3 - t_end_w2 # silent interval between NP and verb/will
  ) %>%
  select(
    sona_id,
    recording,
    subj,
    list,
    tense,
    verb,
    entity,
    regularity,
    trial_n,
    block_trial_n,
    prime_status,
    prime_reg,
    spoken,
    correct,
    adverbial,
    n_words_ok,
    head_ok,
    verb_ok,
    tense_ok,
    utt_onset,
    np_onset,
    np_head_dur,
    np_offset,
    crit_onset,
    gap
  )

syn_analysis %>% filter(correct) %>% glimpse()
```


```{r}
# 8. Phon: same timing measures
phon_words <- alignments %>%
  filter(exp == "phon", tier == "words") %>%
  group_by(sona_id, recording) %>%
  arrange(t_start, .by_group = TRUE) %>%
  mutate(word_n = row_number()) %>%
  ungroup()

phon_spoken <- phon_words %>%
  group_by(sona_id, recording) %>%
  summarize(spoken = paste(word_phone, collapse = " "), .groups = "drop")

phon_w3 <- phon_words %>%
  filter(word_n == 3) %>%
  select(sona_id, recording, w3 = word_phone)

phon_timing <- phon_words %>%
  filter(word_n <= 3) %>%
  select(sona_id, recording, word_n, t_start, t_end) %>%
  pivot_wider(
    names_from = word_n,
    values_from = c(t_start, t_end),
    names_sep = "_w"
  )

phon_analysis <- phon_words %>%
  distinct(
    sona_id,
    recording,
    tense,
    verb,
    entity,
    regularity,
    subj,
    list,
    trial_n,
    block_trial_n,
    prime_status,
    target_sentence
  ) %>%
  left_join(phon_spoken, by = c("sona_id", "recording")) %>%
  left_join(phon_w3, by = c("sona_id", "recording")) %>%
  left_join(phon_timing, by = c("sona_id", "recording")) %>%
  left_join(
    response_flags %>%
      select(
        sona_id,
        recording,
        utt_ok,
        adverbial,
        n_words_ok,
        head_ok,
        verb_ok,
        tense_ok
      ),
    by = c("sona_id", "recording")
  ) %>%
  mutate(
    target_words = str_to_lower(str_remove_all(
      target_sentence,
      "[^a-zA-Z ]"
    )) %>%
      str_squish(),
    correct = if (!filter_trials) {
      TRUE
    } else if (strict_matching) {
      spoken == target_words
    } else {
      coalesce(utt_ok, FALSE)
    },
    utt_onset = t_start_w1,
    np_onset = t_start_w2,
    np_head_dur = t_end_w2 - t_start_w2,
    np_offset = t_end_w2,
    crit_onset = t_start_w3,
    gap = t_start_w3 - t_end_w2
  ) %>%
  select(
    sona_id,
    recording,
    subj,
    list,
    tense,
    verb,
    entity,
    regularity,
    trial_n,
    block_trial_n,
    prime_status,
    spoken,
    correct,
    adverbial,
    n_words_ok,
    head_ok,
    verb_ok,
    tense_ok,
    utt_onset,
    np_onset,
    np_head_dur,
    np_offset,
    crit_onset,
    gap
  )

phon_analysis %>% filter(correct) %>% glimpse()
```

---


```{r, include=FALSE}
knitr::opts_chunk$set(include = TRUE, echo = FALSE)
library(knitr)

# ─── SHARED REPORT PREP ──────────────────────────────────────────────────────
# 1. All timing measures go to milliseconds here, once. Everything downstream
#    (tables, plots, models) is in ms.
# 2. Surface-form frequency: the Zipf frequency of the word form the trial
#    actually requires — the past form for past trials, the bare form for
#    future trials. Generated by analysis/make_verb_frequencies.py.
verb_freq <- read_csv("data/verb_frequencies.csv", show_col_types = FALSE)

# Word-level durations, with the phone count MFA aligned inside each word.
# Duration measures are then expressed per phoneme, so that a segmentally longer
# word is not counted as slower speech: "swept" and "carried" differ in length
# before any planning effect is considered.
phones_in_words <- {
  w <- alignments %>%
    filter(tier == "words") %>%
    select(
      exp,
      sona_id,
      recording,
      verb,
      tense,
      word = word_phone,
      w_start = t_start,
      w_end = t_end
    ) %>%
    group_by(exp, sona_id, recording) %>%
    arrange(w_start, .by_group = TRUE) %>%
    mutate(word_n = row_number()) %>%
    ungroup()
  ph <- alignments %>%
    filter(tier == "phones", !word_phone %in% c("sil", "sp", "spn", "")) %>%
    select(sona_id, recording, p_start = t_start, p_end = t_end)
  w %>%
    left_join(
      ph,
      by = c("sona_id", "recording"),
      relationship = "many-to-many"
    ) %>%
    filter(p_start >= w_start - 1e-6, p_end <= w_end + 1e-6) %>%
    count(
      exp,
      sona_id,
      recording,
      word_n,
      word,
      w_start,
      w_end,
      verb,
      tense,
      name = "n_phones"
    )
}

# The verb is found by matching the word against the form the trial required
# (past form for past trials, bare form for future), not by position: "will"
# takes slot 3 in the future condition and pushes the verb to slot 4.
word_metrics <- phones_in_words %>%
  left_join(verb_freq %>% select(verb, past_form), by = "verb") %>%
  mutate(target_form = if_else(tense == "past", past_form, verb)) %>%
  group_by(exp, sona_id, recording) %>%
  summarize(
    subj_nphones = n_phones[match(2, word_n)],
    verb_dur = {
      i <- match(TRUE, word == first(target_form))
      if (is.na(i)) NA_real_ else (w_end - w_start)[i]
    },
    verb_nphones = {
      i <- match(TRUE, word == first(target_form))
      if (is.na(i)) NA_integer_ else n_phones[i]
    },
    .groups = "drop"
  )

# Articulation rate for the trial, computed only from words that are NOT
# measured (i.e. excluding the subject noun and the verb). Using the whole
# utterance would be endogenous: a slowly produced subject would lower the very
# rate used to adjust it.
speech_rate <- phones_in_words %>%
  left_join(verb_freq %>% select(verb, past_form), by = "verb") %>%
  mutate(target_form = if_else(tense == "past", past_form, verb)) %>%
  group_by(exp, sona_id, recording) %>%
  filter(word_n != 2, word != first(target_form)) %>%
  summarize(
    rate = sum(n_phones) / sum(w_end - w_start),
    .groups = "drop"
  )

# Spoken duration of each utterance. AutoRecordMS is a constant 4500 ms for
# every trial, so the raw recording length carries no information; the length
# that varies is how long the participant actually spoke.
utt_duration <- alignments %>%
  filter(tier == "words") %>%
  group_by(exp, sona_id, recording) %>%
  summarize(utt_dur = max(t_end) - min(t_start), .groups = "drop")

attach_word_metrics <- function(df, exp_name) {
  df %>%
    left_join(
      word_metrics %>% filter(exp == exp_name) %>% select(-exp),
      by = c("sona_id", "recording")
    ) %>%
    left_join(
      speech_rate %>% filter(exp == exp_name) %>% select(-exp),
      by = c("sona_id", "recording")
    ) %>%
    left_join(
      utt_duration %>% filter(exp == exp_name) %>% select(-exp),
      by = c("sona_id", "recording")
    ) %>%
    # Carry-over control: how long the participant spoke on the PRECEDING trial.
    # trial_n is contiguous within participant, so lag() gives the true
    # predecessor; the first trial of each 12-trial cycle is a burn trial and is
    # excluded anyway, so no lag crosses a block boundary into the analysis set.
    arrange(sona_id, trial_n) %>%
    group_by(sona_id) %>%
    mutate(prev_len = lag(utt_dur)) %>%
    ungroup() %>%
    mutate(
      subj_dur_pp = np_head_dur / subj_nphones,
      verb_dur_pp = verb_dur / verb_nphones,
      z_rate = as.numeric(scale(rate)),
      z_prev = as.numeric(scale(prev_len)),
      entity = factor(entity)
    )
}
syn_analysis <- attach_word_metrics(syn_analysis, "syntax")
phon_analysis <- attach_word_metrics(phon_analysis, "phon")

time_cols <- c(
  "utt_onset",
  "np_onset",
  "np_head_dur",
  "np_offset",
  "crit_onset",
  "gap",
  "verb_dur",
  "subj_dur_pp",
  "verb_dur_pp"
)
syn_analysis <- syn_analysis %>% mutate(across(all_of(time_cols), ~ . * 1000))
phon_analysis <- phon_analysis %>% mutate(across(all_of(time_cols), ~ . * 1000))

# An onset of exactly 0 means the aligner placed speech at the very start of the
# recording: the participant was already talking, so the latency is not
# measurable rather than being zero.
drop_zero_onsets <- function(df) {
  df %>%
    mutate(across(
      any_of(c("utt_onset", "np_onset", "crit_onset")),
      ~ if_else(.x <= 0, NA_real_, .x)
    ))
}
syn_analysis <- drop_zero_onsets(syn_analysis)
phon_analysis <- drop_zero_onsets(phon_analysis)


attach_freq <- function(df) {
  df %>%
    left_join(
      verb_freq %>% select(verb, zipf_lemma, zipf_past),
      by = "verb"
    ) %>%
    mutate(
      zipf_form = if_else(tense == "past", zipf_past, zipf_lemma),
      # Frequency enters as a RANK over the distinct word forms: least frequent
      # = 1, most frequent = K. Ranking discards the spacing between Zipf
      # values, so a handful of very frequent verbs no longer dominates an
      # otherwise flat predictor.
      rank_freq = dense_rank(zipf_form),
      # Centred, then divided by its SD. Centring makes the intercept the
      # mid-frequency item; the SD division is a units choice that keeps the
      # existing normal(0, 0.3) prior on b meaningful. On raw centred ranks
      # (range +/- 9.5) that prior would imply implausible effects across the
      # range and would stop regularising.
      z_freq = as.numeric(scale(rank_freq))
    )
}
syn_analysis <- attach_freq(syn_analysis)
phon_analysis <- attach_freq(phon_analysis)

measure_levels <- c(
  "Utterance onset",
  "NP onset",
  "Subject duration\n(ms/phone)",
  "Verb duration\n(ms/phone)",
  "Gap\n(NP-verb/will)"
)

long_measures <- function(df) {
  df %>%
    filter(correct, prime_status != "burn") %>%
    select(
      sona_id,
      tense,
      regularity,
      prime_status,
      `Utterance onset` = utt_onset,
      `NP onset` = np_onset,
      `Subject duration\n(ms/phone)` = subj_dur_pp,
      `Verb duration\n(ms/phone)` = verb_dur_pp,
      `Gap\n(NP-verb/will)` = gap
    ) %>%
    pivot_longer(
      all_of(measure_levels),
      names_to = "measure",
      values_to = "value"
    ) %>%
    mutate(measure = factor(measure, levels = measure_levels))
}

syn_plot_data <- long_measures(syn_analysis)
phon_plot_data <- long_measures(phon_analysis)

cumulative_words <- function(exp_name, analysis_df) {
  alignments %>%
    filter(exp == exp_name, tier == "words") %>%
    group_by(sona_id, recording) %>%
    arrange(t_start, .by_group = TRUE) %>%
    mutate(word_n = row_number()) %>%
    ungroup() %>%
    semi_join(
      analysis_df %>% filter(correct),
      by = c("sona_id", "recording")
    ) %>%
    filter(prime_status != "burn", word_n <= 6) %>%
    mutate(across(c(t_start, t_end), ~ . * 1000))
}
syn_cumulative <- cumulative_words("syntax", syn_analysis)
phon_cumulative <- cumulative_words("phon", phon_analysis)

# 3. Counts for inline use.
concept_n_subj <- n_distinct(concept$sona_id_url)
concept_total <- nrow(exp)
concept_burn <- exp %>% filter(prime_status == "burn") %>% nrow()
concept_included <- exp %>% filter(prime_status != "burn", rt_ok) %>% nrow()
concept_rt_trimmed <- exp %>% filter(prime_status != "burn", !rt_ok) %>% nrow()

# How strongly tense repetition and regularity repetition co-vary in the
# conceptual sequence (they are anti-correlated by list construction).
concept_rep <- exp %>%
  filter(prime_status != "burn", prime_reg != "burn")
concept_rrep_given_tprime <- mean(
  concept_rep$prime_reg[concept_rep$prime_status == "primed"] == "primed"
)
concept_rrep_given_tswitch <- mean(
  concept_rep$prime_reg[concept_rep$prime_status == "unprimed"] == "primed"
)

exp_counts <- function(df) {
  list(
    n_subj = n_distinct(df$sona_id),
    total = nrow(df),
    wrong = sum(!df$correct, na.rm = TRUE),
    burn = sum(df$prime_status == "burn", na.rm = TRUE),
    included = sum(df$correct & df$prime_status != "burn", na.rm = TRUE)
  )
}
syn_c <- exp_counts(syn_analysis)
phon_c <- exp_counts(phon_analysis)
syn_c$pct_wrong <- round(100 * syn_c$wrong / syn_c$total)
phon_c$pct_wrong <- round(100 * phon_c$wrong / phon_c$total)

# ─── REPORT PLOT HELPERS ─────────────────────────────────────────────────────
# Facets always in a single row; y always milliseconds.
condition_means_plot <- function(plot_data, cond, cond_label, cond_levels) {
  plot_data %>%
    within_se(value, sona_id, {{ cond }}, prime_status, strata = measure) %>%
    ggplot(aes(
      x = prime_status,
      y = m,
      shape = {{ cond }},
      fill = {{ cond }},
      linetype = {{ cond }},
      group = {{ cond }}
    )) +
    facet_wrap(~measure, scales = "free_y", nrow = 1) +
    geom_line(position = pd, linewidth = 0.7, colour = "black") +
    geom_errorbar(
      aes(ymin = lwr, ymax = upr),
      position = pd,
      width = 0,
      linewidth = 0.7,
      colour = "black",
      linetype = "solid"
    ) +
    geom_point(position = pd, size = 2.5, colour = "black") +
    cond_shapes(cond_label, cond_levels) +
    cond_fills(cond_label, cond_levels) +
    cond_linetypes(cond_label, cond_levels) +
    labs(x = "Prime status", y = "Time (ms)") +
    theme_paper(base_size = 9)
}

timeline_plot <- function(cum_data, cond, cond_label) {
  cum_data %>%
    group_by({{ cond }}, prime_status, word_n) %>%
    summarize(
      m_onset = mean(t_start, na.rm = TRUE),
      m_offset = mean(t_end, na.rm = TRUE),
      se_onset = sd(t_start, na.rm = TRUE) / sqrt(n()),
      se_offset = sd(t_end, na.rm = TRUE) / sqrt(n()),
      .groups = "drop"
    ) %>%
    pivot_longer(
      c(m_onset, m_offset),
      names_to = "bound",
      values_to = "m",
      names_prefix = "m_"
    ) %>%
    mutate(
      se = if_else(bound == "onset", se_onset, se_offset),
      cond = interaction({{ cond }}, prime_status, sep = " / ")
    ) %>%
    ggplot(aes(
      x = word_n + if_else(bound == "offset", 0.3, 0),
      y = m,
      shape = cond,
      fill = cond,
      linetype = cond,
      group = interaction(cond, bound)
    )) +
    geom_line(linewidth = 0.8, alpha = 0.5, colour = "black") +
    geom_errorbar(
      aes(ymin = m - 1.96 * se, ymax = m + 1.96 * se),
      width = 0,
      linewidth = 0.6,
      colour = "black",
      linetype = "solid"
    ) +
    geom_point(size = 2.5, colour = "black") +
    scale_shape_manual(name = cond_label, values = c(16, 21, 17, 24)) +
    scale_fill_manual(
      name = cond_label,
      values = c("black", "white", "black", "white")
    ) +
    scale_linetype_manual(
      name = cond_label,
      values = c("solid", "22", "solid", "22")
    ) +
    scale_x_continuous(
      breaks = 1:4,
      labels = c("the", "NP", "will / Vpast", "V / obj")
    ) +
    labs(x = "Sentence region", y = "Time from trial onset (ms)") +
    theme_paper(base_size = 9) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

means_table <- function(plot_data, cond, caption) {
  d <- plot_data %>%
    within_se(value, sona_id, {{ cond }}, prime_status, strata = measure) %>%
    mutate(
      measure = str_replace_all(as.character(measure), "\n", " "),
      Condition = as.character({{ cond }}),
      `Prime status` = prime_status,
      N = n,
      Mean = sprintf("%.0f", m),
      SE = sprintf("%.1f", se)
    )
  side_by_side(d, c("Condition", "Prime status", "N", "Mean", "SE"), caption)
}

# ─── MODEL WRAPPER ───────────────────────────────────────────────────────────
# One call per experiment fits every measure. `fixed` is the full population
# formula; `subj_slopes` / `item_slopes` name the terms that get random slopes
# (frequency is between-verb, so it never takes a by-verb slope).
fit_timing <- function(
  data,
  dv,
  fixed,
  subj_slopes,
  item_slopes,
  tag,
  family = shifted_lognormal(),
  priors = priors_lognormal,
  subject = "sona_id",
  extra_re = NULL,
  adapt_delta = 0.95
) {
  file <- file.path("analysis/fits", paste0(tag, "_", dv))
  if (!fit_models && !file.exists(paste0(file, ".rds"))) {
    return(NULL)
  }
  # Experiment 1 is keyed by `subj`, the production experiments by `sona_id`.
  stopifnot(all(c(subject, "verb", dv) %in% names(data)))
  main <- as.formula(sprintf(
    "%s ~ %s + (1 + %s | %s) + (1 + %s | verb)%s",
    dv,
    fixed,
    subj_slopes,
    subject,
    item_slopes,
    if (is.null(extra_re)) "" else paste(" +", extra_re)
  ))
  # A hurdle family splits the response into "did it happen at all" and "how
  # large was it". The gap is 67-79% exact zeros, so the same predictors are
  # allowed to act on both parts: priming may change how OFTEN speakers pause
  # rather than how long they pause for. In brms `hu` is the probability of a
  # ZERO, so a positive hu coefficient means MORE no-pause trials.
  is_hurdle <- grepl("hurdle", family$family %||% "", fixed = TRUE)
  form <- if (is_hurdle) {
    bf(main, as.formula(sprintf("hu ~ %s + (1 | %s)", fixed, subject)))
  } else {
    main
  }
  brm(
    form,
    data = data,
    family = family,
    prior = priors,
    chains = 4,
    cores = 4,
    iter = 4000,
    warmup = 1000,
    control = list(adapt_delta = adapt_delta, max_treedepth = 12),
    seed = 1234,
    file = file,
    file_refit = "on_change"
  )
}

# Facet titles for the coefficient plots, matching the condition-mean figures.
measure_pretty <- c(
  utt_onset = "Utt. onset",
  np_onset = "NP onset",
  subj_dur_pp = "Subject dur.",
  verb_dur_pp = "Verb dur.",
  gap = "Gap length"
)

# ─── MODEL LADDER ────────────────────────────────────────────────────────────
# Each spec is fitted to every measure of the experiment. `slopes` is the
# by-participant random slope structure, always a subset of `fixed`: frequency,
# entity and rate are between-item or trial-level covariates and take no
# participant slope.
# Spec 6 (entity as a maximal random effect) is not fitted. With only three
# entity levels the funnel geometry produced 198-434 divergent transitions at
# adapt_delta = 0.95 and still 47-198 at 0.99, at roughly 8x the runtime of any
# other spec. Divergences mean the posterior was not properly explored, so those
# estimates would be biased rather than merely noisy. Spec 5 estimates the same
# entity contrasts as fixed effects and samples cleanly. The machinery in
# fit_timing() (extra_re, adapt_delta) is retained so the attempt is
# reproducible.
specs_tense <- list(
  list(
    id = "s1",
    label = "1: Tense x Prime",
    fixed = "c_tense * c_prime",
    slopes = "c_tense * c_prime"
  ),
  list(
    id = "s2",
    label = "2: + Frequency",
    fixed = "c_tense * c_prime + z_freq",
    slopes = "c_tense * c_prime"
  ),
  list(
    id = "s3",
    label = "3: + Reg. prime",
    fixed = "c_tense * c_prime + z_freq + c_rprime",
    slopes = "c_tense * c_prime + c_rprime"
  ),
  list(
    id = "s5",
    label = "5: + Entity, rate",
    fixed = "c_tense * c_prime + z_freq + c_rprime + entity + z_rate",
    slopes = "c_tense * c_prime + c_rprime"
  )
)

# Experiment 1 is a keypress task with no audio, so there is no articulation
# rate to control for; its spec 5 adds entity alone.
specs_tense_norate <- map(specs_tense, function(sp) {
  sp$fixed <- sub(" \\+ z_rate$", "", sp$fixed)
  sp
})

specs_reg <- list(
  list(
    id = "s1",
    label = "1: Regularity x Prime + Freq",
    fixed = "c_reg * c_prime + z_freq",
    slopes = "c_reg * c_prime"
  ),
  list(
    id = "s5",
    label = "2: + Entity, rate",
    fixed = "c_reg * c_prime + z_freq + entity + z_rate",
    slopes = "c_reg * c_prime"
  )
)

# Spec 5 controls for the subject noun and the trial's articulation rate. The
# three entity nouns differ sharply in per-phoneme duration (Chef 162, Pirate
# 118, Wizard 107 ms/phone), so leaving entity out loads that variance onto the
# residual for every duration measure.
fit_ladder <- function(data, measures, specs, tag, subject = "sona_id") {
  set_names(specs, map_chr(specs, "label")) %>%
    map(function(sp) {
      set_names(measures, coalesce(measure_pretty[measures], measures)) %>%
        map(
          ~ fit_timing(
            data,
            .x,
            sp$fixed,
            sp$slopes,
            "c_prime",
            paste0(tag, "_", sp$id),
            extra_re = sp$extra_re,
            adapt_delta = sp$adapt_delta %||% 0.95,
            family = if (.x == "gap") {
              hurdle_lognormal()
            } else {
              shifted_lognormal()
            },
            priors = if (.x == "gap") priors_hurdle else priors_lognormal,
            subject = subject
          )
        ) %>%
        compact()
    })
}

# One table + one figure per spec, so every model in the ladder is inspectable.
# The reported model. The ladder is kept for the comparison table, but every
# figure and coefficient table in the body comes from this spec.
headline_spec <- "5: + Entity, rate"
headline_spec_reg <- "2: + Entity, rate"

pick_headline <- function(ladder) {
  nm <- intersect(c(headline_spec, headline_spec_reg), names(ladder))
  if (!length(nm)) {
    return(list())
  }
  ladder[[nm[1]]]
}

# ── Item-level forest plot ───────────────────────────────────────────────────
# By-verb deviations from the population intercept: how much each item pulls the
# measure up or down once the fixed effects are accounted for. Points are
# posterior medians, bars 95% CrIs; a verb whose interval excludes zero is
# behaving differently from the rest of the set.
item_forest <- function(models, exp_label) {
  if (!length(models)) {
    return(invisible(NULL))
  }
  d <- imap_dfr(models, function(fit, nm) {
    r <- ranef(fit, groups = "verb", probs = c(0.025, 0.975))$verb
    tibble(
      measure = nm,
      verb = dimnames(r)[[1]],
      est = r[, "Estimate", "Intercept"],
      lo = r[, "Q2.5", "Intercept"],
      hi = r[, "Q97.5", "Intercept"]
    )
  }) %>%
    mutate(measure = factor(measure, levels = names(models)))

  ord <- d %>%
    filter(measure == levels(d$measure)[1]) %>%
    arrange(est) %>%
    pull(verb)

  ggplot(d, aes(x = est, y = factor(verb, levels = ord))) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "gray50") +
    geom_errorbar(
      aes(xmin = lo, xmax = hi),
      orientation = "y",
      width = 0,
      linewidth = 0.5,
      colour = "black"
    ) +
    geom_point(size = 1.6, colour = "black") +
    facet_wrap(~measure, nrow = 1, scales = "free_x") +
    labs(
      x = "By-verb deviation from the intercept",
      y = NULL,
      title = exp_label
    ) +
    theme_paper(base_size = 9, legend_position = "none") +
    theme(axis.text.y = element_text(size = 6.5))
}

# ── Bayesian sanity checks ───────────────────────────────────────────────────
# Divergences mean the sampler could not explore part of the posterior, so the
# estimates are biased rather than merely noisy. Rhat > 1.01 or bulk ESS < 400
# means the chains have not mixed well enough to trust the summaries.
convergence_table <- function(models, caption) {
  if (!length(models)) {
    return(invisible(NULL))
  }
  tbl <- imap_dfr(models, function(fit, nm) {
    np <- nuts_params(fit)
    dr <- summarise_draws(as_draws_df(fit), "rhat", "ess_bulk", "ess_tail")
    dr <- dr[!is.na(dr$rhat), ]
    tibble(
      measure = nm,
      divergences = sum(subset(np, Parameter == "divergent__")$Value),
      max_rhat = round(max(dr$rhat), 3),
      min_ess_bulk = round(min(dr$ess_bulk)),
      min_ess_tail = round(min(dr$ess_tail))
    )
  }) %>%
    mutate(ok = divergences == 0 & max_rhat <= 1.01 & min_ess_bulk >= 400)
  print(kable(tbl, caption = caption))
  invisible(tbl)
}

# Posterior predictive check: simulated datasets from the fitted model against
# the observed distribution. Systematic mismatch means the likelihood is wrong,
# regardless of how well the chains converged.
pp_grid <- function(models, exp_label) {
  if (!length(models)) {
    return(invisible(NULL))
  }
  ps <- imap(models, function(fit, nm) {
    pp_check(fit, ndraws = 50) +
      labs(title = nm, x = NULL, y = NULL) +
      theme_paper(base_size = 7, legend_position = "none") +
      theme(
        plot.title = element_text(size = 7, face = "bold"),
        axis.text = element_text(size = 5.5)
      )
  })
  wrap_plots(ps, nrow = 1) +
    plot_annotation(title = paste(exp_label, "- posterior predictive checks"))
}

report_ladder <- function(ladder, term_labels, exp_label, fig_prefix) {
  for (nm in names(ladder)) {
    models <- ladder[[nm]]
    if (!length(models)) {
      next
    }
    cat(sprintf("\n\n**%s -- model %s**\n\n", exp_label, nm))
    report_models(
      models,
      term_labels,
      sprintf(
        "%s, model %s. Positive = slower/longer. * marks terms with P(dir) at or above 0.90.",
        exp_label,
        nm
      ),
      figure_name = sprintf(
        "%s-%s",
        fig_prefix,
        tolower(gsub("[^a-z0-9]+", "-", tolower(nm)))
      )
    )
  }
}

# Predictive comparison across the ladder, per measure.
compare_ladder <- function(ladder, caption) {
  ms <- unique(unlist(lapply(ladder, names)))
  rows <- map_dfr(ms, function(m) {
    fits <- compact(lapply(ladder, function(l) l[[m]]))
    if (length(fits) < 2) {
      return(NULL)
    }
    # loo_compare() dispatches on its first argument: a bare list of brmsfit
    # objects lands in the default method and errors. Pass the loo objects.
    out <- try(
      {
        fits <- lapply(fits, add_criterion, "loo")
        loos <- lapply(fits, function(f) f$criteria$loo)
        cmp <- loo_compare(loos)
        tibble(
          measure = m,
          model = rownames(cmp),
          elpd_diff = round(cmp[, "elpd_diff"], 1),
          se_diff = round(cmp[, "se_diff"], 1)
        )
      },
      silent = TRUE
    )
    # A failed comparison must not discard the fits that took hours to produce.
    if (inherits(out, "try-error")) {
      warning(
        "LOO comparison failed for ",
        m,
        ": ",
        conditionMessage(attr(out, "condition"))
      )
      NULL
    } else {
      out
    }
  })
  if (nrow(rows)) print(kable(rows, caption = caption))
}

fit_all_measures <- function(
  data,
  measures,
  fixed,
  subj_slopes,
  item_slopes,
  tag
) {
  set_names(measures, coalesce(measure_pretty[measures], measures)) %>%
    map(
      ~ fit_timing(
        data,
        .x,
        fixed,
        subj_slopes,
        item_slopes,
        tag,
        family = if (.x == "gap") hurdle_lognormal() else shifted_lognormal(),
        priors = if (.x == "gap") priors_hurdle else priors_lognormal
      )
    ) %>%
    compact()
}


# Readable table: one block per measure, terms spelled out, and the rows with
# most of the posterior on one side of zero in bold.
# Probabilities are never printed as 0.00 or 1.00, which would claim more
# certainty than a finite posterior sample can support.
fmt_p <- function(p) {
  ifelse(p >= 0.995, "> 0.99", ifelse(p <= 0.005, "< 0.01", sprintf("%.2f", p)))
}

# Long tables are split into two side-by-side panels so a five-measure model
# fits on one page. The split falls on a measure boundary, never inside one.
side_by_side <- function(d, cols, caption, font_size = 7) {
  blocks <- split(
    seq_len(nrow(d)),
    factor(d$measure, levels = unique(d$measure))
  )
  sizes <- lengths(blocks)
  cut <- which.min(abs(cumsum(sizes) - sum(sizes) / 2))
  left <- unlist(blocks[seq_len(cut)])
  right <- unlist(blocks[-seq_len(cut)])
  if (!length(right)) {
    k <- kable(
      d[, cols, drop = FALSE],
      booktabs = TRUE,
      caption = caption,
      align = paste(c("l", rep("r", length(cols) - 1)), collapse = "")
    ) %>%
      kableExtra::kable_styling(
        latex_options = "hold_position",
        font_size = font_size
      ) %>%
      kableExtra::pack_rows(
        index = table(factor(d$measure, levels = unique(d$measure)))
      )
    if (any(d$strong)) {
      k <- kableExtra::row_spec(k, which(d$strong), bold = TRUE)
    }
    return(k)
  }

  # The measure appears once per block, in its own column, so the two halves can
  # sit next to each other in a single table. Row-level bolding is not available
  # once the halves are side by side, so strong terms are marked with a star.
  panel <- function(idx) {
    dd <- d[idx, , drop = FALSE]
    first <- !duplicated(as.character(dd$measure))
    out <- data.frame(
      Measure = ifelse(first, as.character(dd$measure), ""),
      dd[, cols, drop = FALSE],
      check.names = FALSE
    )
    last <- names(out)[ncol(out)]
    out[[last]] <- paste0(out[[last]], ifelse(dd$strong, "*", ""))
    out
  }
  L <- panel(left)
  R <- panel(right)
  n <- max(nrow(L), nrow(R))
  pad <- function(x, n) {
    while (nrow(x) < n) {
      x[nrow(x) + 1, ] <- rep("", ncol(x))
    }
    x
  }
  both <- cbind(pad(L, n), pad(R, n))
  hdr <- c("Measure", "Term", "Est.", "95% CrI", "P(dir)")
  if (length(cols) == 5) {
    hdr <- c("Measure", "Condition", "Prime", "N", "Mean", "SE")
  }

  kable(
    both,
    booktabs = TRUE,
    caption = caption,
    row.names = FALSE,
    col.names = rep(hdr, 2),
    # kable inserts \addlinespace every five rows by default, which lands inside
    # a block once the two halves are side by side. Suppressing it needs the
    # format stated explicitly.
    format = "latex",
    linesep = "",
    align = paste(rep(c("l", "l", rep("r", ncol(L) - 2)), 2), collapse = "")
  ) %>%
    kableExtra::kable_styling(
      latex_options = "hold_position",
      font_size = font_size
    )
}

model_table <- function(tbl, term_labels, caption) {
  # Labels for the control terms, which are the same in every model.
  extra <- c(
    entityPirate = "Pirate vs Chef",
    entityWizard = "Wizard vs Chef",
    z_rate = "Articulation rate",
    z_freq = "Frequency",
    c_rprime = "Regularity priming"
  )
  d <- tbl %>%
    filter(!grepl("Intercept", .variable)) %>%
    mutate(measure = factor(measure, levels = unique(measure))) %>%
    arrange(measure) %>%
    mutate(
      Term = label_terms(.variable, term_labels),
      strong = pmax(p_greater_0, 1 - p_greater_0) >= 0.90,
      Estimate = sprintf("%.3f", estimate),
      `95% CrI` = sprintf("[%.3f, %.3f]", lower_95, upper_95),
      # P(direction): the posterior mass on the side of zero the estimate is on.
      `P(dir)` = fmt_p(pmax(p_greater_0, 1 - p_greater_0))
    ) %>%
    select(measure, Term, Estimate, `95% CrI`, `P(dir)`, strong)

  if (n_distinct(d$measure) > 2) {
    return(side_by_side(
      d,
      c("Term", "Estimate", "95% CrI", "P(dir)"),
      caption,
      font_size = 6
    ))
  }
  k <- d %>%
    select(-measure, -strong) %>%
    kable(booktabs = TRUE, align = "lrrr", caption = caption,
          format = "latex", linesep = "") %>%
    kableExtra::kable_styling(
      latex_options = "hold_position",
      font_size = 8
    ) %>%
    kableExtra::pack_rows(
      index = table(factor(d$measure, levels = unique(d$measure)))
    )
  if (any(d$strong)) {
    k <- kableExtra::row_spec(k, which(d$strong), bold = TRUE)
  }
  k
}


report_ladder <- function(ladder, term_labels, exp_label, fig_prefix) {
  for (nm in names(ladder)) {
    models <- ladder[[nm]]
    if (!length(models)) {
      next
    }
    cat(sprintf("\n\n**%s -- model %s**\n\n", exp_label, nm))
    report_models(
      models,
      term_labels,
      sprintf(
        "%s, model %s. Positive = slower/longer. * marks terms with P(dir) at or above 0.90.",
        exp_label,
        nm
      ),
      figure_name = sprintf(
        "%s-%s",
        fig_prefix,
        tolower(gsub("[^a-z0-9]+", "-", tolower(nm)))
      )
    )
  }
}

# Predictive comparison across the ladder, per measure.
compare_ladder <- function(ladder, caption) {
  ms <- unique(unlist(lapply(ladder, names)))
  rows <- map_dfr(ms, function(m) {
    fits <- compact(lapply(ladder, function(l) l[[m]]))
    if (length(fits) < 2) {
      return(NULL)
    }
    # loo_compare() dispatches on its first argument: a bare list of brmsfit
    # objects lands in the default method and errors. Pass the loo objects.
    out <- try(
      {
        fits <- lapply(fits, add_criterion, "loo")
        loos <- lapply(fits, function(f) f$criteria$loo)
        cmp <- loo_compare(loos)
        tibble(
          measure = m,
          model = rownames(cmp),
          elpd_diff = round(cmp[, "elpd_diff"], 1),
          se_diff = round(cmp[, "se_diff"], 1)
        )
      },
      silent = TRUE
    )
    # A failed comparison must not discard the fits that took hours to produce.
    if (inherits(out, "try-error")) {
      warning(
        "LOO comparison failed for ",
        m,
        ": ",
        conditionMessage(attr(out, "condition"))
      )
      NULL
    } else {
      out
    }
  })
  if (nrow(rows)) print(kable(rows, caption = caption))
}

fit_all_measures <- function(
  data,
  measures,
  fixed,
  subj_slopes,
  item_slopes,
  tag
) {
  set_names(measures, coalesce(measure_pretty[measures], measures)) %>%
    map(
      ~ fit_timing(
        data,
        .x,
        fixed,
        subj_slopes,
        item_slopes,
        tag,
        family = if (.x == "gap") hurdle_lognormal() else shifted_lognormal(),
        priors = if (.x == "gap") priors_hurdle else priors_lognormal
      )
    ) %>%
    compact()
}


# Readable table: one block per measure, terms spelled out, and the rows with
# most of the posterior on one side of zero in bold.

report_ladder <- function(ladder, term_labels, exp_label, fig_prefix) {
  for (nm in names(ladder)) {
    models <- ladder[[nm]]
    if (!length(models)) {
      next
    }
    cat(sprintf("\n\n**%s -- model %s**\n\n", exp_label, nm))
    report_models(
      models,
      term_labels,
      sprintf(
        "%s, model %s. Positive = slower/longer. * marks terms with P(dir) at or above 0.90.",
        exp_label,
        nm
      ),
      figure_name = sprintf(
        "%s-%s",
        fig_prefix,
        tolower(gsub("[^a-z0-9]+", "-", tolower(nm)))
      )
    )
  }
}

# Predictive comparison across the ladder, per measure.
compare_ladder <- function(ladder, caption) {
  ms <- unique(unlist(lapply(ladder, names)))
  rows <- map_dfr(ms, function(m) {
    fits <- compact(lapply(ladder, function(l) l[[m]]))
    if (length(fits) < 2) {
      return(NULL)
    }
    # loo_compare() dispatches on its first argument: a bare list of brmsfit
    # objects lands in the default method and errors. Pass the loo objects.
    out <- try(
      {
        fits <- lapply(fits, add_criterion, "loo")
        loos <- lapply(fits, function(f) f$criteria$loo)
        cmp <- loo_compare(loos)
        tibble(
          measure = m,
          model = rownames(cmp),
          elpd_diff = round(cmp[, "elpd_diff"], 1),
          se_diff = round(cmp[, "se_diff"], 1)
        )
      },
      silent = TRUE
    )
    # A failed comparison must not discard the fits that took hours to produce.
    if (inherits(out, "try-error")) {
      warning(
        "LOO comparison failed for ",
        m,
        ": ",
        conditionMessage(attr(out, "condition"))
      )
      NULL
    } else {
      out
    }
  })
  if (nrow(rows)) print(kable(rows, caption = caption))
}

fit_all_measures <- function(
  data,
  measures,
  fixed,
  subj_slopes,
  item_slopes,
  tag
) {
  set_names(measures, coalesce(measure_pretty[measures], measures)) %>%
    map(
      ~ fit_timing(
        data,
        .x,
        fixed,
        subj_slopes,
        item_slopes,
        tag,
        family = if (.x == "gap") hurdle_lognormal() else shifted_lognormal(),
        priors = if (.x == "gap") priors_hurdle else priors_lognormal
      )
    ) %>%
    compact()
}

# One place where raw parameter names become readable labels.
label_terms <- function(v, term_labels) {
  extra <- c(
    entityPirate = "Pirate vs Chef",
    entityWizard = "Wizard vs Chef",
    z_rate = "Articulation rate",
    z_freq = "Frequency",
    c_rprime = "Regularity priming"
  )
  bare <- sub("^b_(hu_)?", "", v)
  out <- unname(term_labels[paste0("b_", bare)])
  out[is.na(out)] <- unname(extra[bare[is.na(out)]])
  out[is.na(out)] <- bare[is.na(out)]
  out
}

term_order <- function(lab, term_labels) {
  factor(
    lab,
    levels = rev(unique(c(
      unname(term_labels),
      "Regularity priming",
      "Frequency",
      "Articulation rate",
      "Pirate vs Chef",
      "Wizard vs Chef",
      sort(unique(lab))
    )))
  )
}

coef_table <- function(models) {
  if (!length(models)) {
    return(NULL)
  }
  imap_dfr(models, function(fit, nm) {
    fit %>%
      gather_draws(`b_.*`, regex = TRUE) %>%
      group_by(.variable) %>%
      summarize(
        estimate = median(.value),
        lower_95 = quantile(.value, 0.025),
        upper_95 = quantile(.value, 0.975),
        p_greater_0 = mean(.value > 0),
        .groups = "drop"
      ) %>%
      # The pause model has two parts: the duration of a pause when one occurs,
      # and the probability of producing none. They are separate panels.
      mutate(measure = if_else(grepl("^b_hu_", .variable), "No pause lik.", nm))
  })
}

coef_plot <- function(tbl, term_labels) {
  keep <- unname(term_labels)
  d <- tbl %>%
    filter(!grepl("Intercept", .variable)) %>%
    filter(label_terms(.variable, term_labels) %in% keep) %>%
    mutate(
      lab = label_terms(.variable, term_labels),
      lab = term_order(lab, term_labels),
      measure = factor(measure, levels = unique(measure)),
      pdir = pmax(p_greater_0, 1 - p_greater_0)
    ) %>%
    droplevels() %>%
    group_by(measure) %>%
    mutate(.text_x = max(upper_95) + 0.30 * (max(upper_95) - min(lower_95))) %>%
    ungroup()

  ggplot(d, aes(x = estimate, y = lab)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "gray50") +
    geom_errorbar(
      aes(xmin = lower_95, xmax = upper_95),
      orientation = "y",
      width = 0.15,
      linewidth = 0.7,
      colour = "black"
    ) +
    geom_point(size = 2.4, colour = "black") +
    geom_text(aes(x = .text_x, label = fmt_p(pdir)), hjust = 0.5, size = 2.7) +
    geom_text(
      data = d %>% distinct(measure, .text_x),
      aes(x = .text_x, y = Inf, label = "P(dir)"),
      inherit.aes = FALSE,
      vjust = -0.7,
      size = 2.7,
      fontface = "bold"
    ) +
    facet_wrap(~measure, nrow = 1, scales = "free_x") +
    scale_x_continuous(expand = expansion(mult = c(0.10, 0.28)), n.breaks = 4) +
    coord_cartesian(clip = "off") +
    labs(x = "Estimate (log scale)", y = NULL) +
    theme_paper(legend_position = "none") +
    theme(
      axis.text.y = element_text(face = "bold", size = 8),
      plot.margin = margin(t = 18, r = 8, b = 4, l = 4, unit = "pt")
    )
}


report_models <- function(models, term_labels, caption, figure_name = NULL) {
  if (!length(models)) {
    cat("\n*Models not fitted. Set `fit_models <- TRUE` and re-render.*\n\n")
    return(invisible(NULL))
  }
  tbl <- coef_table(models)
  print(model_table(tbl, term_labels, caption))

  # The pause model is on a different scale from the timing measures, so it gets
  # its own panel row rather than being squeezed alongside them.
  gap_names <- c("Gap length", "No pause lik.")
  main <- tbl %>% filter(!measure %in% gap_names)
  gapp <- tbl %>% filter(measure %in% gap_names)

  p1 <- if (nrow(main)) coef_plot(main, term_labels) else NULL
  p2 <- if (nrow(gapp)) coef_plot(gapp, term_labels) else NULL
  p <- if (!is.null(p1) && !is.null(p2)) {
    p1 / p2 + patchwork::plot_layout(heights = c(1, 1))
  } else if (!is.null(p1)) {
    p1
  } else {
    p2
  }

  if (!is.null(figure_name) && !is.null(p)) {
    save_figure(
      p,
      figure_name,
      width = 7.2,
      height = if (!is.null(p1) && !is.null(p2)) 5.4 else 3
    )
  }
  print(p)
}


library(brms)
library(tidybayes)
dir.create("analysis/fits", recursive = TRUE, showWarnings = FALSE)

priors_lognormal <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 0.3), class = "b"),
  prior(normal(0, 0.3), class = "sd"),
  prior(lkj(2), class = "cor")
)

priors_hurdle <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 0.3), class = "b"),
  prior(normal(0, 0.3), class = "sd"),
  prior(lkj(2), class = "cor"),
  prior(normal(0, 1), class = "Intercept", dpar = "hu"),
  prior(normal(0, 1), class = "b", dpar = "hu"),
  prior(normal(0, 1), class = "sd", dpar = "hu")
)

priors_gaussian <- c(
  prior(normal(0, 500), class = "Intercept"),
  prior(normal(0, 100), class = "b"),
  prior(normal(0, 100), class = "sd"),
  prior(lkj(2), class = "cor")
)

code_half <- function(x, plus) {
  out <- ifelse(x == plus, 0.5, -0.5)
  if (length(unique(na.omit(out))) < 2) {
    stop(sprintf(
      "code_half(): '%s' matches nothing, so the contrast is constant. Levels present: %s",
      plus,
      paste(unique(x), collapse = ", ")
    ))
  }
  out
}

# Population-level terms, following the switches above. `cond` is the design
# factor for the experiment (c_tense or c_reg).
fixed_terms <- function(cond) {
  core <- if (fit_three_way) {
    sprintf("%s * c_prime * c_rprime", cond)
  } else if (fit_prime_interaction) {
    sprintf("%s * c_prime + c_prime * c_rprime", cond)
  } else {
    sprintf("%s * c_prime + c_rprime", cond)
  }
  paste(core, "+ z_freq")
}

# Every term the switches can produce, in display order. coef_plot() drops the
# ones a given fit does not contain.
prime_term_labels <- function(
  cond_label,
  cond_term,
  prime_label = "Tense priming",
  prime_short = "Priming"
) {
  # Only the terms that can appear in the reported model. Interactions with
  # regularity priming are never fitted; it is a control.
  setNames(
    c(
      cond_label,
      prime_label,
      "Regularity priming",
      "Frequency",
      paste0(cond_label, " x ", prime_short)
    ),
    c(
      paste0("b_", cond_term),
      "b_c_prime",
      "b_c_rprime",
      "b_z_freq",
      paste0("b_", cond_term, ":c_prime")
    )
  )
}

timing_measures <- c(
  "utt_onset",
  "np_onset",
  "subj_dur_pp",
  "verb_dur_pp",
  "gap"
)

phon_fit_data <- phon_analysis %>%
  filter(correct, prime_status != "burn") %>%
  mutate(
    c_reg = code_half(regularity, "IRREGULAR"),
    c_prime = code_half(prime_status, "primed"),
    sona_id = factor(sona_id),
    verb = factor(verb)
  ) %>%
  drop_na(utt_onset, z_freq, z_rate)

syn_fit_data <- syn_analysis %>%
  filter(correct, prime_status != "burn") %>%
  mutate(
    c_tense = code_half(tense, "future"),
    c_prime = code_half(prime_status, "primed"),
    c_rprime = code_half(prime_reg, "primed"),
    sona_id = factor(sona_id),
    verb = factor(verb)
  ) %>%
  drop_na(utt_onset, z_freq, z_rate)

concept_fit_data <- exp %>%
  filter(prime_status != "burn", rt_ok) %>%
  left_join(verb_freq %>% select(verb, zipf_past), by = "verb") %>%
  mutate(
    c_tense = code_half(tense, "future"),
    c_prime = code_half(prime_status, "primed"),
    c_rprime = code_half(prime_reg, "primed"),
    rank_freq = dense_rank(zipf_past),
    z_freq = as.numeric(scale(rank_freq)),
    z_prev = as.numeric(scale(prev_len)),
    subj = factor(subj),
    verb = factor(verb),
    entity = factor(entity)
  ) %>%
  drop_na(decision_rt, z_freq)
```

```{r}
# Helpers so every number in the prose is pulled from the fits and the data
# rather than typed in.
MOD <- "s5"

bb <- function(ex, dv, term, dpar = NULL) {
  f <- sprintf("analysis/fits/%s_%s_%s.rds", ex, MOD, dv)
  d <- posterior::as_draws_df(readRDS(f))
  v <- if (is.null(dpar)) paste0("b_", term) else paste0("b_", dpar, "_", term)
  x <- d[[v]]
  list(
    b = median(x),
    lo = quantile(x, .025),
    hi = quantile(x, .975),
    p = max(mean(x > 0), mean(x < 0)),
    pgt = mean(x > 0),
    pct = 100 * (exp(median(x)) - 1)
  )
}

# "b = 0.04, 95% CrI [0.01, 0.07], P(b > 0) = 0.99"
rep_b <- function(ex, dv, term, dpar = NULL, digits = 3) {
  z <- bb(ex, dv, term, dpar)
  sprintf(
    "$\\beta$ = %.*f, 95%% CrI [%.*f, %.*f], $P_{dir}$ = %s",
    digits,
    z$b,
    digits,
    z$lo,
    digits,
    z$hi,
    fmt_p(z$p)
  )
}

# "1090 ms (SE 15.6)"
ms <- function(df, v, ..., unit = " ms") {
  d <- df %>% filter(...)
  x <- d[[v]][!is.na(d[[v]])]
  sprintf("%.0f%s (SE %.1f)", mean(x), unit, sd(x) / sqrt(length(x)))
}
e1 <- exp %>% filter(prime_status != "burn", rt_ok)
e2 <- phon_analysis %>% filter(correct, prime_status != "burn")
e3 <- syn_analysis %>% filter(correct, prime_status != "burn")
prate <- function(df, ...) {
  d <- df %>% filter(...)
  sprintf(
    "%.1f%% (SE %.1f)",
    100 * mean(d$gap > 0),
    100 * sqrt(mean(d$gap > 0) * (1 - mean(d$gap > 0)) / nrow(d))
  )
}
```

# Report

Three experiments test whether tense is planned before articulation begins, and
whether the cost of that planning is morphosyntactic (selecting a tense feature)
or morphophonological (retrieving the form that realises it). Experiment 1 uses a
speeded keypress decision; Experiments 2 and 3 use spoken production, with timing
recovered from forced-aligned recordings.

All timing measures are in milliseconds. Error bars are 95% intervals built on
Morey (2008) within-subject standard errors. Conditions are distinguished by
marker fill and line type rather than colour, so the figures survive greyscale
printing.

## Data and preprocessing

Spoken responses were uploaded per block as `.webm`, converted to 16 kHz mono
wav, transcribed with AssemblyAI, and force-aligned against those transcripts
with the Montreal Forced Aligner (`english_us_arpa` acoustic model and
dictionary, out-of-vocabulary items passed through G2P). Word and phone
boundaries come from the resulting TextGrids. The commands are listed in the
appendix.

Recordings that never reached the server, and sessions with no usable audio, are
outside the participant screen below: they simply have no timing data. One
morphophonology participant uploaded 101 empty five-byte files and contributes no
recordings at all.

```{r, include=TRUE, results='asis'}
# 4c. Participant filtering write-up. Reflects whichever switches are set above.
exp_label <- function(e) c(phon = "morphophonology", syntax = "morphosyntax")[e]

n_start <- subject_stats %>% count(exp, name = "n")
n_end <- kept_subjects %>% count(exp, name = "n")

phrase <- function(df, unit = "participants") {
  paste(sprintf("%d %s %s", df$n, exp_label(df$exp), unit), collapse = " and ")
}

cat("## Participant filtering\n\n")

if (!filter_participants || nrow(filter_log) == 0) {
  cat(sprintf(
    paste(
      "No participant-level exclusions were applied. All %s with",
      "aligned recordings enter the analyses below.\n\n"
    ),
    phrase(n_start)
  ))
} else {
  cat(sprintf(
    paste(
      "Forced alignment yielded %s. Exclusions were applied in the",
      "order below; each step acts only on the participants that",
      "survived the preceding steps.\n\n"
    ),
    phrase(n_start)
  ))

  for (i in seq_len(nrow(distinct(filter_log, step)))) {
    s <- distinct(filter_log, step, desc)[i, ]
    rows <- filter_log %>% filter(step == s$step)
    counts <- paste(
      sprintf(
        "%d of %d in %s (%d remaining)",
        rows$n_removed,
        rows$n_before,
        exp_label(rows$exp),
        rows$n_after
      ),
      collapse = ", and "
    )
    cat(sprintf(
      "**Step %d, %s.** Participants who %s were excluded: %s.\n\n",
      i,
      tolower(s$step),
      s$desc,
      counts
    ))
  }

  cat(sprintf(
    "After all participant-level exclusions, %s remain.\n\n",
    phrase(n_end)
  ))
}

if (filter_trials) {
  trial_counts <- response_flags %>%
    semi_join(kept_subjects, by = c("exp", "sona_id")) %>%
    group_by(exp) %>%
    summarize(n = n(), kept = sum(utt_ok, na.rm = TRUE), .groups = "drop")

  cat(sprintf(
    paste(
      "Within the retained participants, individual trials were excluded",
      "when the response did not name the correct subject entity, did not",
      "use the target verb in the tense the trial called for, did not match",
      "the target sentence in length, was empty, or marked tense with an",
      "adverbial instead of on the verb. This retained %s.\n\n"
    ),
    paste(
      sprintf(
        "%d of %d trials (%.0f%%) in %s",
        trial_counts$kept,
        trial_counts$n,
        100 * trial_counts$kept / trial_counts$n,
        exp_label(trial_counts$exp)
      ),
      collapse = " and "
    )
  ))
} else {
  cat(
    "No trial-level exclusions were applied; all trials from retained participants enter the analyses.\n\n"
  )
}
```


## Experiment 1: Conceptual task

### Prediction

Repetition should have no effect here. If the repetition effects in the
production experiments come from reusing linguistic material, then a task that
requires no verb form to be produced gives nothing to reuse. Participants decide
which time a picture refers to and press a key. Deciding about the future may
take longer than deciding about the past, but repeating a decision about the same
time should not speed up the next one, and the effect of repetition should not
differ between the two tenses. A repetition effect here would mean the effects in
the other two experiments need not be linguistic at all.


`r concept_n_subj` participants learned character–verb–time pairings (Chef,
Pirate, Wizard x 20 verbs) and made speeded keypress decisions (C = Past,
M = Future). The manipulation is **tense priming**: whether the current trial's
tense matches the previous trial's (primed) or switches (unprimed). No speech is
recorded, so the utterance screen does not apply here.

### Filtering

Of `r concept_total` trials, `r concept_burn` boundary trials (positions 1, 6, 7
and 12 of each 12-trial block) were excluded. A further `r concept_rt_trimmed`
trials fell outside the `r rt_min`–`r rt_max` ms response window and were dropped
as anticipations or disengagement: the raw responses run from 0 ms to 173 s, and
the upper tail distorts both the means and the lognormal fit. That leaves
**`r concept_included` trials**.

### Means

```{r}
exp %>%
  filter(prime_status != "burn", rt_ok) %>%
  group_by(Tense = tense, `Prime status` = prime_status) %>%
  summarize(
    N = n(),
    `Mean RT (ms)` = round(mean(decision_rt, na.rm = TRUE)),
    SE = round(sd(decision_rt, na.rm = TRUE) / sqrt(n()), 1),
    .groups = "drop"
  ) %>%
  kable(caption = "Experiment 1: decision RT by tense x prime status")
```

Regularity repetition is also tracked. The two are not independent: regularity
repeats on `r round(100 * concept_rrep_given_tprime)`% of trials where tense
repeats but `r round(100 * concept_rrep_given_tswitch)`% of trials where it
switches, so the two priming effects are estimated jointly in one model rather
than separately.

```{r}
exp %>%
  filter(prime_status != "burn", prime_reg != "burn", rt_ok) %>%
  group_by(`Regularity prime` = prime_reg, `Tense prime` = prime_status) %>%
  summarize(
    N = n(),
    `Mean RT (ms)` = round(mean(decision_rt, na.rm = TRUE)),
    SE = round(sd(decision_rt, na.rm = TRUE) / sqrt(n()), 1),
    .groups = "drop"
  ) %>%
  kable(caption = "Experiment 1: decision RT by regularity prime x tense prime")
```

### Plot

```{r, fig.height=2.1, fig.width=4.2}
p_concept <- exp %>%
  filter(prime_status != "burn", rt_ok) %>%
  mutate(measure = "Decision RT") %>%
  within_se(decision_rt, subj, tense, prime_status, strata = measure) %>%
  ggplot(aes(
    x = prime_status,
    y = m,
    shape = tense,
    fill = tense,
    linetype = tense,
    group = tense
  )) +
  geom_line(position = pd, linewidth = 0.7, colour = "black") +
  geom_errorbar(
    aes(ymin = lwr, ymax = upr),
    position = pd,
    width = 0,
    linewidth = 0.7,
    colour = "black",
    linetype = "solid"
  ) +
  geom_point(position = pd, size = 2.5, colour = "black") +
  cond_shapes("Tense", c("past", "future")) +
  cond_fills("Tense", c("past", "future")) +
  cond_linetypes("Tense", c("past", "future")) +
  labs(x = "Prime status", y = "Decision RT (ms)") +
  theme_paper(base_size = 9)

save_figure(p_concept, "fig-exp1-concept-means", width = 4.6, height = 2.1)
p_concept
```

### Modelling

The model estimates each measure separately. Positive coefficients mean a later
onset or a longer duration. The measures are the time from the start of the trial
to the first word, the time to the subject noun, the duration of the subject noun
per phone, the duration of the verb per phone, and the interval between the
subject noun and the verb.

Speakers reach the verb without pausing on most trials, so the interval is
modelled in two parts. **Gap length** is how long the pause lasts on the trials where
there is one. **No pause lik.** is how likely the speaker is to run
straight into the verb; a positive coefficient there means fewer pauses. The
first part is on the log scale in milliseconds, the second on the log-odds scale.


```{r, results='asis', fig.height=2.5, fig.width=6}
concept_ladder <- fit_ladder(
  concept_fit_data,
  "decision_rt",
  specs_tense_norate,
  "concept",
  subject = "subj"
)

# ── Reported model: spec 5 (design x prime + freq + reg-prime + entity + rate)
concept_ladder_main <- pick_headline(concept_ladder)

report_models(
  concept_ladder_main,
  prime_term_labels("Tense", "c_tense"),
  "Experiment 1, reported model (spec 5). Positive = slower/longer.",
  figure_name = "fig-exp1-coefs-main"
)
```

### Items

By-verb deviations from the intercept in the reported model. A verb whose
interval excludes zero behaves differently from the rest of the item set.

```{r, fig.height=3, fig.width=5}
p_items <- item_forest(concept_ladder_main, "Experiment 1")
if (!is.null(p_items)) {
  save_figure(p_items, "fig-exp1-item-forest", width = 6, height = 3)
  p_items
}
```



---


### Results

Decisions were slower for future than for past. Future trials averaged
`r ms(e1, "decision_rt", tense == "future")` and past trials
`r ms(e1, "decision_rt", tense == "past")`. The model supports this difference
(`r rep_b("concept", "decision_rt", "c_tense")`).

Tense repetition did not affect decision times. Primed trials averaged
`r ms(e1, "decision_rt", prime_status == "primed")` and unprimed trials
`r ms(e1, "decision_rt", prime_status == "unprimed")`, and the model gives
`r rep_b("concept", "decision_rt", "c_prime")`. The interaction between tense and
tense repetition is absent
(`r rep_b("concept", "decision_rt", "c_tense:c_prime")`), with the posterior
centred on zero.

This is what we expected. The conceptual task requires participants to retrieve
which time a picture refers to and press a key, with no verb form to produce.
Deciding about the future takes longer than deciding about the past, but
repeating a decision about the same time does not make the next one faster.


---

## Experiment 2: Morphophonology

### Prediction

Repetition should act near the verb. The form of the past tense is not needed
until the verb is articulated, so preparing it should leave the time to begin
speaking unchanged and appear on the words immediately around the verb. The
effect should be larger for irregular verbs, whose past forms have to be
retrieved individually, than for regular verbs, which share one ending. The
clearest place to see this is the interaction between regularity and repetition,
on the duration of the subject noun and on whether the speaker pauses before
reaching the verb.


`r phon_c$n_subj` participants produced past-tense descriptions of the same
scenes (*The pirate spun a top*). The manipulation is **morphophonological
priming** by verb regularity: whether the current trial's verb shares its past
formation with the previous trial's. All trials are past tense, so tense is held
constant and regularity carries the design.

**Measures.** Utterance onset (when *the* begins), NP onset, NP head duration
(the entity name), and the gap between NP offset and the verb.

### Filtering

Of `r phon_c$total` trials, `r phon_c$wrong` (`r phon_c$pct_wrong`%) failed the
utterance screen and `r phon_c$burn` were boundary trials, leaving
**`r phon_c$included` trials** from `r phon_c$n_subj` participants.

### Means

```{r}
means_table(
  phon_plot_data,
  regularity,
  "Experiment 2: timing measures by regularity x prime status"
)
```

### Plot

```{r, fig.height=2, fig.width=8}
p_phon <- condition_means_plot(
  phon_plot_data,
  regularity,
  "Regularity",
  c("IRREGULAR", "REGULAR")
)
save_figure(p_phon, "fig-exp2-phon-means", width = 7.2, height = 2.2)
p_phon
```

### Modelling

The model estimates each measure separately. Positive coefficients mean a later
onset or a longer duration. The measures are the time from the start of the trial
to the first word, the time to the subject noun, the duration of the subject noun
per phone, the duration of the verb per phone, and the interval between the
subject noun and the verb.

Speakers reach the verb without pausing on most trials, so the interval is
modelled in two parts. **Gap length** is how long the pause lasts on the trials where
there is one. **No pause lik.** is how likely the speaker is to run
straight into the verb; a positive coefficient there means fewer pauses. The
first part is on the log scale in milliseconds, the second on the log-odds scale.


```{r, results='asis', fig.height=5, fig.width=8}
phon_ladder <- fit_ladder(phon_fit_data, timing_measures, specs_reg, "phon")

# ── Reported model: spec 5 (design x prime + freq + reg-prime + entity + rate)
phon_ladder_main <- pick_headline(phon_ladder)

report_models(
  phon_ladder_main,
  prime_term_labels(
    "Regularity",
    "c_reg",
    prime_label = "Regularity priming",
    prime_short = "Priming"
  ),
  "Experiment 2, reported model (spec 5). Positive = slower/longer.",
  figure_name = "fig-exp2-coefs-main"
)
```

### Items

By-verb deviations from the intercept in the reported model. A verb whose
interval excludes zero behaves differently from the rest of the item set.

```{r, fig.height=2.5, fig.width=8}
p_items <- item_forest(phon_ladder_main, "Experiment 2")
if (!is.null(p_items)) {
  save_figure(p_items, "fig-exp2-item-forest", width = 7.2, height = 4)
  p_items
}
```



---


### Results

Repetition of the past-tense form shortened the verb. After an irregular prime,
irregular verbs were produced in `r ms(e2, "verb_dur_pp", regularity == "IRREGULAR", prime_status == "primed", unit = " ms/phone")`
against `r ms(e2, "verb_dur_pp", regularity == "IRREGULAR", prime_status == "unprimed", unit = " ms/phone")`
when unprimed. Regular verbs changed less, from
`r ms(e2, "verb_dur_pp", regularity == "REGULAR", prime_status == "unprimed", unit = " ms/phone")`
to `r ms(e2, "verb_dur_pp", regularity == "REGULAR", prime_status == "primed", unit = " ms/phone")`.
The model supports the repetition effect
(`r rep_b("phon", "verb_dur_pp", "c_prime")`) but does not separate the two verb
types (`r rep_b("phon", "verb_dur_pp", "c_reg:c_prime")`).

Repetition also changed the subject noun, and here the two verb types behave
differently. Before a regular verb, the subject was
`r ms(e2, "subj_dur_pp", regularity == "REGULAR", prime_status == "unprimed", unit = " ms/phone")`
when unprimed and `r ms(e2, "subj_dur_pp", regularity == "REGULAR", prime_status == "primed", unit = " ms/phone")`
when primed. Before an irregular verb the two were close,
`r ms(e2, "subj_dur_pp", regularity == "IRREGULAR", prime_status == "unprimed", unit = " ms/phone")`
and `r ms(e2, "subj_dur_pp", regularity == "IRREGULAR", prime_status == "primed", unit = " ms/phone")`.
The model gives `r rep_b("phon", "subj_dur_pp", "c_reg:c_prime")` for the
interaction, which we read as support given the direction and the size of the
posterior mass above zero.

The clearest effect is on whether speakers paused at all before the verb. Pauses
were most frequent before an unprimed irregular verb, on
`r prate(e2, regularity == "IRREGULAR", prime_status == "unprimed")` of trials,
and dropped to `r prate(e2, regularity == "IRREGULAR", prime_status == "primed")`
when the preceding trial had also used an irregular. Regular verbs showed no such
change, `r prate(e2, regularity == "REGULAR", prime_status == "unprimed")` when
unprimed and `r prate(e2, regularity == "REGULAR", prime_status == "primed")` when
primed. The model separates all three terms: irregular verbs are followed by more
pauses (`r rep_b("phon", "gap", "c_reg", dpar = "hu")`), repetition reduces
pausing (`r rep_b("phon", "gap", "c_prime", dpar = "hu")`), and the reduction is
larger for irregulars (`r rep_b("phon", "gap", "c_reg:c_prime", dpar = "hu")`).
The coefficients are on the probability of producing no pause, so a positive
value means fewer pauses.

Speech onset was unaffected. Primed trials began at
`r ms(e2, "utt_onset", prime_status == "primed")` and unprimed trials at
`r ms(e2, "utt_onset", prime_status == "unprimed")`
(`r rep_b("phon", "utt_onset", "c_prime")`).

Repetition of the past-tense form therefore acts late. It leaves the time to
begin speaking unchanged and shows up on the subject noun, on the verb itself,
and on whether the speaker pauses before reaching the verb. This is what we
predicted for a form-based effect: the form is not needed until the verb is
articulated, so preparing it does not delay the start of the sentence.


---

## Experiment 3: Morphosyntax

### Prediction

Repetition should act early. Choosing between past and future is a choice about
the structure of the sentence, made before articulation begins, so its effects
should appear at the start of the utterance rather than around the verb. The
interaction between tense and repetition at the onset of the subject noun is the
measure this predicts, and the later measures should show little.


`r syn_c$n_subj` participants produced past or future descriptions of the same
scenes (*The pirate spun a top* vs *The pirate will spin a top*). The
manipulation is **tense priming**, identical in logic to Experiment 1, but the
response is spoken, so the dependent measures are timing rather than keypress
latency.

**Measures.** As in Experiment 2, with the gap running from NP offset to the
tense-bearing word (*will* in the future, the past-marked verb in the past).

### Filtering

Of `r syn_c$total` trials, `r syn_c$wrong` (`r syn_c$pct_wrong`%) failed the
utterance screen and `r syn_c$burn` were boundary trials, leaving
**`r syn_c$included` trials** from `r syn_c$n_subj` participants.

### Means

```{r}
means_table(
  syn_plot_data,
  tense,
  "Experiment 3: timing measures by tense x prime status"
)
```

As in Experiment 1, regularity repetition is tracked alongside tense repetition
and the two enter the models together.

```{r}
syn_analysis %>%
  filter(correct, prime_status != "burn", prime_reg != "burn") %>%
  group_by(`Regularity prime` = prime_reg, `Tense prime` = prime_status) %>%
  summarize(
    N = n(),
    `Utterance onset (ms)` = round(mean(utt_onset, na.rm = TRUE)),
    `NP head dur (ms)` = round(mean(np_head_dur, na.rm = TRUE)),
    `Gap (ms)` = round(mean(gap, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  kable(caption = "Experiment 3: timing by regularity prime x tense prime")
```

### Plot

```{r, fig.height=2, fig.width=8}
p_syn <- condition_means_plot(
  syn_plot_data,
  tense,
  "Tense",
  c("past", "future")
)
save_figure(p_syn, "fig-exp3-syn-means", width = 7.2, height = 2.2)
p_syn
```


### Modelling

The model estimates each measure separately. Positive coefficients mean a later
onset or a longer duration. The measures are the time from the start of the trial
to the first word, the time to the subject noun, the duration of the subject noun
per phone, the duration of the verb per phone, and the interval between the
subject noun and the verb.

Speakers reach the verb without pausing on most trials, so the interval is
modelled in two parts. **Gap length** is how long the pause lasts on the trials where
there is one. **No pause lik.** is how likely the speaker is to run
straight into the verb; a positive coefficient there means fewer pauses. The
first part is on the log scale in milliseconds, the second on the log-odds scale.


```{r, results='asis', fig.height=5, fig.width=8}
syn_ladder <- fit_ladder(syn_fit_data, timing_measures, specs_tense, "syn")

# ── Reported model: spec 5 (design x prime + freq + reg-prime + entity + rate)
syn_ladder_main <- pick_headline(syn_ladder)

report_models(
  syn_ladder_main,
  prime_term_labels("Tense", "c_tense"),
  "Experiment 3, reported model (spec 5). Positive = slower/longer.",
  figure_name = "fig-exp3-coefs-main"
)
```

### Items

By-verb deviations from the intercept in the reported model. A verb whose
interval excludes zero behaves differently from the rest of the item set.

```{r, fig.height=2.5, fig.width=8}
p_items <- item_forest(syn_ladder_main, "Experiment 3")
if (!is.null(p_items)) {
  save_figure(p_items, "fig-exp3-item-forest", width = 7.2, height = 4)
  p_items
}
```



---


### Results

Tense affected the sentence throughout. Future sentences reached the subject noun
at `r ms(e3, "np_onset", tense == "future")` against
`r ms(e3, "np_onset", tense == "past")` for past
(`r rep_b("syn", "np_onset", "c_tense")`). The subject noun was shorter in the
future, `r ms(e3, "subj_dur_pp", tense == "future", unit = " ms/phone")` against
`r ms(e3, "subj_dur_pp", tense == "past", unit = " ms/phone")`
(`r rep_b("syn", "subj_dur_pp", "c_tense")`), and the verb was longer,
`r ms(e3, "verb_dur_pp", tense == "future", unit = " ms/phone")` against
`r ms(e3, "verb_dur_pp", tense == "past", unit = " ms/phone")`
(`r rep_b("syn", "verb_dur_pp", "c_tense")`).

Tense repetition on its own did nothing. Primed trials reached the subject noun
at `r ms(e3, "np_onset", prime_status == "primed")` and unprimed trials at
`r ms(e3, "np_onset", prime_status == "unprimed")`
(`r rep_b("syn", "np_onset", "c_prime")`), and the same holds for the other
measures.

Repetition interacted with tense at the subject noun. In the future, primed
trials reached the noun at
`r ms(e3, "np_onset", tense == "future", prime_status == "primed")` and unprimed
trials at `r ms(e3, "np_onset", tense == "future", prime_status == "unprimed")`.
In the past the ordering reverses,
`r ms(e3, "np_onset", tense == "past", prime_status == "primed")` when primed and
`r ms(e3, "np_onset", tense == "past", prime_status == "unprimed")` when unprimed.
The model gives `r rep_b("syn", "np_onset", "c_tense:c_prime")`. The posterior
leans in the predicted direction but the interval is wide, and with
`r n_distinct(e3$sona_id)` participants we treat this as suggestive rather than
established. A replication would need roughly 45 participants to resolve it.

The interaction appears early, before the verb has been reached, which is where
we expected an effect of tense selection to appear. Later measures do not show
it: neither the duration of the subject noun
(`r rep_b("syn", "subj_dur_pp", "c_tense:c_prime")`) nor the duration of the verb
(`r rep_b("syn", "verb_dur_pp", "c_tense:c_prime")`) separates the conditions.


---



# Appendix: model comparison and checks

## Model comparison

Models were compared by leave-one-out cross-validation. Differences are relative
to the best model for that measure; a difference smaller than about twice its
standard error is not meaningful.

```{r, results='asis'}
compare_ladder(concept_ladder, "Experiment 1: model comparison")
compare_ladder(phon_ladder, "Experiment 2: model comparison")
compare_ladder(syn_ladder, "Experiment 3: model comparison")
```

## Convergence

No divergent transitions, $\hat{R}$ at or below 1.01, and effective sample sizes
in the thousands for every model reported.

```{r, results='asis'}
convergence_table(concept_ladder_main, "Experiment 1: convergence")
convergence_table(phon_ladder_main, "Experiment 2: convergence")
convergence_table(syn_ladder_main, "Experiment 3: convergence")
```

## Posterior predictive checks

Each panel compares the observed distribution of a measure (dark line) against
data simulated from the fitted model (light lines).

```{r, fig.height=2, fig.width=8}
pp_grid(phon_ladder_main, "Experiment 2")
```

```{r, fig.height=2, fig.width=7.2}
pp_grid(syn_ladder_main, "Experiment 3")
```

```{r, fig.height=2, fig.width=8}
pp_grid(concept_ladder_main, "Experiment 1")
```


# Appendix: statistical models

All models were fitted with brms (Stan), four chains of 4000 iterations with the
first 1000 discarded, `adapt_delta` = 0.95.

## Contrast coding

All categorical predictors are sum coded at $\pm 0.5$, so each coefficient is the
difference between the two levels and the intercept is the grand mean.

| predictor | +0.5 | -0.5 | used in |
|---|---|---|---|
| Tense | future | past | Experiments 1 and 3 |
| Regularity | irregular | regular | Experiment 2 |
| Tense priming | primed | unprimed | Experiments 1 and 3 |
| Regularity priming | primed | unprimed | all three |

In Experiment 2 regularity priming is the manipulation and appears as *Priming*
in the tables. In Experiments 1 and 3 the manipulation is tense priming, and
regularity priming enters only as a control.

The subject noun (Chef, Pirate, Wizard) is treated as a factor with Chef as the
reference level. Frequency is the rank of the produced word form in the SUBTLEX
Zipf scale, from least to most frequent, centred and divided by its standard
deviation. Ranking rather than the Zipf value itself keeps a small number of very
frequent verbs from dominating an otherwise flat predictor. Articulation rate is
the number of phones per second on that trial, computed over the words that are
not being measured, and is also centred and scaled.

## Response distributions

Onsets and durations are positive and right skewed and were fitted with a shifted
lognormal distribution. The shift estimates a floor below which no response
occurs.

The interval between the subject noun and the verb is zero on
`r sprintf("%.0f%%", 100*mean(e2$gap == 0))` of trials in Experiment 2 and
`r sprintf("%.0f%%", 100*mean(e3$gap == 0))` in Experiment 3, because speakers
often move into the verb without pausing. A single continuous distribution
describes this badly, so it was fitted with a hurdle lognormal: one part gives the
probability of producing no pause, the other the duration of the pause when there
is one. Both parts take the same predictors, which allows repetition to change how
often speakers pause separately from how long they pause. Coefficients labelled as
being on the probability of no pause come from the first part, where a positive
value means fewer pauses.

## Model structure

Every measure was fitted with the same structure. For Experiments 1 and 3,

$$
\text{measure} \sim \text{Tense} \times \text{Tense priming} + \text{Regularity priming} + \text{Frequency} + \text{Character} + \text{Rate}
$$
$$
+ \; (1 + \text{Tense} \times \text{Tense priming} + \text{Regularity priming} \mid \text{participant}) + (1 + \text{Tense priming} \mid \text{verb})
$$

and for Experiment 2,

$$
\text{measure} \sim \text{Regularity} \times \text{Priming} + \text{Frequency} + \text{Character} + \text{Rate}
$$
$$
+ \; (1 + \text{Regularity} \times \text{Priming} \mid \text{participant}) + (1 + \text{Priming} \mid \text{verb})
$$

Experiment 1 has no recordings and therefore no rate term.


Each measure was fitted with the same population-level terms: the design factor
(tense or regularity), repetition, their interaction, frequency, the subject noun,
articulation rate, and, in Experiments 1 and 3, form repetition as a control.
Random effects are by participant, with slopes for the design factor, repetition
and their interaction, and by verb, with a slope for repetition. Frequency, the
subject noun and articulation rate vary between items or within a trial and take
no by-participant slope.

Regularity priming is included because it is not independent of tense priming in
these materials. The form repeats on
`r sprintf("%.0f%%", 100*mean((e3 %>% filter(prime_status=="primed", prime_reg!="burn"))$prime_reg == "primed"))`
of trials where tense repeats and
`r sprintf("%.0f%%", 100*mean((e3 %>% filter(prime_status=="unprimed", prime_reg!="burn"))$prime_reg == "primed"))`
of trials where it switches, so leaving it out would let part of a form-repetition
effect appear as a tense priming effect. It is never entered into an
interaction.

## Priors

| parameter | prior |
|---|---|
| intercept | Normal(0, 1) |
| population-level coefficients | Normal(0, 0.3) |
| random-effect standard deviations | Normal(0, 0.3) |
| random-effect correlations | LKJ(2) |
| hurdle intercept and coefficients | Normal(0, 1) |

Coefficients are on the log scale, so Normal(0, 0.3) places most of its mass on
effects smaller than a factor of two and rules out nothing the data could show.

## Reporting

Each coefficient is reported as its posterior median, a 95% credible interval,
and $P_{dir}$, the proportion of the posterior lying on the same side of zero as
the median. $P_{dir}$ runs from 0.5, meaning the posterior is split evenly and
the sign is undetermined, to 1, meaning the whole posterior falls on one side.
It is not a significance test and has no threshold attached; we describe an
effect as supported when most of the posterior is on one side and the interval is
narrow enough to be informative. Values are never printed as 0 or 1, since a
finite posterior sample cannot establish either.

## Convergence and model checking

Every model reported here converged: no divergent transitions, $\hat{R} \le 1.01$,
and bulk effective sample sizes above 400. Posterior predictive checks compare
simulated data against the observed distribution for each measure. Posterior
contraction, the reduction in variance from prior to posterior, is above 0.94 for
the frequency, subject noun and articulation rate terms and between 0.56 and 0.99
for the interactions, so the posteriors are determined by the data rather than by
the priors.

Models were compared by leave-one-out cross-validation. Adding the subject noun
and articulation rate improves prediction on the duration measures by a wide
margin, which is why they are in the reported model: the three subject nouns
differ by more than 50 ms per phone, far more than any effect of the manipulation.


# Appendix: recordings and measurement

Spoken responses were recorded in the browser, one file per trial, and uploaded
in blocks. Recordings were converted to 16 kHz mono audio and transcribed
automatically. Each transcript was then aligned to its recording with the
Montreal Forced Aligner using its English acoustic model and pronunciation
dictionary, with pronunciations generated for words outside the dictionary. The
alignment gives the start and end time of every word and every phone, and all
timing measures are read from it.

Speech onset is the start of the first word. The onset of the subject noun and
its duration are taken from the second word. The interval before the verb runs
from the end of the subject noun to the start of the verb, and is zero when the
two are adjacent. The verb is located by matching the transcript against the
form the trial called for rather than by position, since the future adds a word
before it.

Durations of the subject noun and the verb are divided by the number of phones
the aligner placed inside them. Words differ in length, and without this a longer
word would count as slower speech. Articulation rate is the number of phones per
second over the remaining words of the sentence, which leaves the two measured
words out of their own control variable.

Transcripts were scored against the sentence the trial called for. A response was
kept if it named the right character and used the right verb in the right tense.
Material after the verb was not scored, since every measure lies at or before the
verb. Two adjustments were made after inspecting the transcripts. The past tense
of *blow* is a homophone of *blue* and was transcribed as the colour on most
trials, so both spellings were accepted. *Blowed* was kept as an error, being a
real regularisation rather than a transcription problem. Responses that marked
time with an adverbial instead of on the verb, such as *the wizard climbs a
ladder in the past*, were excluded, since they leave nothing for the verb to
carry.

Participants were excluded if fewer than half their responses met these criteria.
